### 2022-07-27 manuscript.Rmd
implement POLMM for ordinal variables

---
output: word_document
bibliography: doc/references.bib
csl: doc/nature.csl
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinical AD; however, clinical AD does not always correspond with AD neuropathology and vice-versa.[@nelson2012] We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:

1.  identify novel genetic risk loci for individual NPE
2.  identify common sources of genetic risk (pleiotropy) between NPE
3.  provide evidence that the effects of clinical AD genetic risk loci may be mediated by specific NPE.

# Methods

We used genotype and NPE data from four aged autopsy cohorts: National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Memory and Aging Project (ROSMAP), Adult Changes in Thought (ACT), and the AD Neuroimaging Initiative (ADNI).

## Quality Control

Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS,[@shade2022] with the exception that all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021]

1.  Retrieve participants labeled as non-Hispanic white or similar (European, etc.) if available.
2.  Perform imputation on TOPMed Imputation Server (ROSMAP and ADNI had already been imputed using same reference panel by collaborators, so this step was skipped for those cohorts).
3.  Covert VCF files to PLINK file set, keeping variants will minor allele frequencies (MAF) of 0.1% or greater.
4.  Merge sub-cohorts of same study (e.g. NACC cohorts 1-12 and ACT 1-3).
5.  Remove duplicate participants in each study.
6.  Remove participants without autopsy data available.
7.  Iteratively remove variants and participants with excess missingness until no variant is missing in more than 5% of participants and no participant is missing more than 5% of variants.
8.  Remove participants with unusually high/low heterozygosity (+/- 3 SD cutoff).
9.  Merge with 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012] and remove participants without primarily European ancestry as determined by principal component analysis.
10. Remove participants with rare neurological diseases that substantially raise the likelihood that present NPE could be due to unusual pathological processes (NACC only, as the other cohorts did not have this information available).
11. Remove duplicate participants between studies, preferentially keeping the observation in the larger cohort (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
12. Merge cohorts.
13. Harmonize NPE variables to analyze.
14. Create list of related ($\ge$ 2nd-degree relatives) individuals using KING[@manichaikul2010] to remove for each NPE variable, randomly keeping one participant in each related cluster.
15. Create genetic principal components using the PC-AiR method in the GENESIS package in R 4.1.[@conomos2015]

We note that there are substantial differences in the way that NPE are collected in different cohorts,(cite) and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes and minimizing computational load for analysis. Thus, some ordinal variables were collapsed to binary variables, or synthetic NPE may have been created using categorical variables from one cohort and continuous variables from another. All code used for harmonization will be made available upon publication, and a brief overview of the harmonization strategy for each NPE will be included in the Supplemental Materials.

## Statistical Analyses

All the analyses discussed below were done on the data set with the combined cohorts.

### Single-variant GWAS

Single-variant logistic regression analyses were performed in PLINK 2.0.[@chang2015; @purcell] Covariates included age at death, sex, cohort, and the first 10 genetic principle components. Continuous covariates were standardized to a mean of 0 and a variance of 1. An additive mode of inheritance was assumed in all analyses. Variants achieving a p-value of $< 5\times 10^{-8}$ were considered "genome-wide" significant. To identify independent genetic risk loci, variants were clumped using PLINK 1.9[@chang] using a linkage-disequilibrium threshold of $r^2 \le 0.05$.

Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package.[@bi2021] Dense kinship matrices were used to account for relatedness.

### Conditional and mediation analyses

For genetic loci associated with multiple NPE, we considered mediation hypotheses to test whether the association of the lead variant with one NPE was mediated by another NPE. Regression analyses were run in R using the `glm` or `polr` functions and mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.

### Colocalization analyses

All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues. Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs. A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) \> 80% was used as a threshold for evidence of colocalization.

### 2022-07-27 code/aux/np_sample_sizes.R

@@ -1,0 +1,302 @@
+library(data.table)
+library(magrittr)
+library(stringi)
+library(readxl)
+
+na_strings <- c("-9", "-4", "9999", "999", "888")
+# Universal Data Set + NP
+uds <- fread("/data_global/nacc/investigator_nacc57.csv", 
+             na.strings = na_strings)
+# NP data not longitudinal, so just keep obs from last visit
+uds <- uds[NACCVNUM == NACCAVST]
+
+# Minimal Data Set + NP
+mds <- fread("/data_global/nacc/fardo09062019.csv", 
+             na.strings = na_strings)
+
+#------------------------------------------------------
+# subset participants with NP data and NP variables
+#------------------------------------------------------
+
+# all autopsied have NPFORMVER
+uds[!(is.na(NPFORMVER)), np := 1]
+table(uds$np, useNA = "a")
+
+mds[!(is.na(NPFORMVER)), np := 1]
+table(mds$np, useNA = "a")
+
+np_vars <- c(colnames(uds)[grep("NP", colnames(uds))])
+np_vars_nacc <- fread("data/nacc_np_vars_names_nacc.txt", header = FALSE)$V1
+np_vars_remove <- fread("data/nacc_uds_var_names_np.txt", header = FALSE)$V1
+np_vars <- np_vars[!(np_vars) %in% np_vars_remove]
+np_vars <- c(np_vars_nacc, np_vars)
+uds_np <- uds[np == 1, c(..np_vars, "RACE", "HISPANIC"), with =FALSE]
+uds_np[, uds := 1]
+# month of death and year of death not in MDS, but still want to have for UDS
+mds[, `:=`(NACCMOD = NA, NACCYOD = NA)]
+np_mds_vars <- np_vars[!(np_vars %in% 
+                           c("NPARTAG", "NPATGSEV", "NPATGAMY",
+                             "NPATGAM1", "NPATGAM2", "NPATGAM3", 
+                             "NPATGAM4", "NPATGAM5", "NPATGFRN", 
+                             "NPATGFR1", "NPATGFR2", "NPATGFR3", 
[…]

### 2022-07-27 code/aux/rosmap_wgs_sample_size.R

@@ -1,0 +1,30 @@
+library(data.table)
+library(stringi)
+library(readxl)
+
+linker <- fread("~/ROSMAP_biospecimen_metadata.csv")
+wgs <- fread("~/ROSMAP_assay_wholeGenomeSeq_metadata.csv")
+clinical <- fread("~/ROSMAP_clinical.csv")
+np <- setDT(read_xlsx(
+  "/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx"))
+
+linker_wgs <- merge(linker[assay == "wholeGenomeSeq", 
+                           .(specimenID, individualID)], 
+                    wgs[, .(specimenID)], 
+                    by = "specimenID")
+linker_wgs <- merge(linker_wgs[, .(individualID)], 
+                    clinical[, .(individualID, projid)], "individualID")
+linker_wgs[, projid := as.character(projid)]
+
+filler_zeroes <- Vectorize(function(id, maxchar = 8) {
+  paste(rep_len(0, maxchar - nchar(id)), collapse = "")
+}, vectorize.args = "id")
+
+linker_wgs[nchar(projid) < 8, 
+           paste("Number if IDs with < 8 chars:", .N)]
+
+linker_wgs[, projid := paste0(filler_zeroes(projid), projid)]
+
+np <- np[!is.na(age_death) & !is.na(ceradsc)]
+
+np_wgs <- merge(np, linker_wgs[, .(projid)], "projid")

### 2022-07-27 code/coloc/find_qtls.py

@@ -20,9 +20,11 @@ if len(snps) == 0:
 
 
 snp_set = set(snps)
-eqtl_dir = "/data_global/GTEx/GTEx_Analysis_v8_QTLs/GTEx_Analysis_v8_eQTL_Eur/eqtls/"
-sqtl_dir = ("/data_global/GTEx/GTEx_Analysis_v8_QTLs/GTEx_Analysis_v8_sQTL_Eur/" +
-            "GTEx_Analysis_v8_sQTL_EUR/"
+eqtl_dir = ("/data_global/GTEx/GTEx_Analysis_v8_QTLs/"
+              "GTEx_Analysis_v8_eQTL_Eur/eqtls/"
+)
+sqtl_dir = ("/data_global/GTEx/GTEx_Analysis_v8_QTLs/"
+            "GTEx_Analysis_v8_sQTL_Eur/GTEx_Analysis_v8_sQTL_EUR/"
 )
 eqtl_dir_files = os.listdir(eqtl_dir)
 sqtl_dir_files = os.listdir(sqtl_dir)

### 2022-07-27 code/conditional/mega_apoe_chr19.sh

@@ -1,0 +1,47 @@
+
+phenos=$(cat data/mega/mega_np.pheno | awk 'NR==1{for (i=3; i<(NF); i++) printf $i "\n"; print $NF}')
+
+# for pheno in $phenos
+# do
+# python code/gwas/plink2_logistic.py \
+#   -f data/mega/mega_np \
+#   -o output/gwas/mega/apoe_chr19/mega \
+#   -p data/mega/mega_np.pheno \
+#   -c data/mega/mega_np_apoe.covar \
+#   -r data/mega/related_rm/"$pheno".remove \
+#   --chrom 19 \
+#   --pheno_name "$pheno"
+# done
+
+# for pheno in $phenos
+# do
+#   cat output/gwas/mega/apoe_chr19/mega."$pheno".glm.logistic | \
+#     awk '$14 < 0.01 {print $3}' \
+#     > tmp/"$pheno"_apoe_p0.01.txt
+#   
+#   plink2 \
+#     --pfile data/mega/mega_np \
+#     --extract tmp/"$pheno"_apoe_p0.01.txt \
+#     --export A \
+#     --export-allele data/mega/varID_ALT.txt \
+#     --out tmp/"$pheno"_apoe_p0.01
+# done
+
+run_ord_reg () {
+  ord_pheno=$1
+  bin_pheno=$2
+  ./code/gwas/ordinal/clm_regression.R \
+  -c data/mega/mega_np_apoe.covar \
+  -g tmp/"$bin_pheno"_apoe_p0.01.raw \
+  -o tmp/"$ord_pheno"_apoe_ord_results.txt \
+  -p data/mega/mega_np_ord.pheno \
+  --phenotype "$ord_pheno" \
+  -r data/mega/related_rm/"$ord_pheno".remove
+}
+
+# run_ord_reg caa_ord caa
+# run_ord_reg cerad cerad3
+# run_ord_reg braak braak56
+run_ord_reg diffuse_abeta diffuse_abeta3
+run_ord_reg late late23
+run_ord_reg lewy_body lewy_body123

### 2022-07-27 code/gwas/gene_names.R

@@ -1,0 +1,38 @@
+
+
+#======================================================
+# annote a variant to the nearest gene
+#======================================================
+
+library(data.table)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("-v", "--variant_file", 
+                    help = "path to file with variant positions")
+parser$add_argument("-g", "--gene_file",  
+                    help = "path to file with gene positions")
+
+args <- parser$parse_args(
+  c('-v', 'output/gwas/mega/braak56.assoc.logistic',
+    '-g', 'raw_data/NCBI38.gene.loc'))
+
+genes <- fread(args$gene_file)
+gene_pos <- melt(genes, id.vars = c("V2", "V6"), measure.vars = c("V3", "V4"))
+snps <- fread(args$variant_file)
+
+snps_top <- snps[P %in% snps[, min(P), CHR]$V1]
+
+find_closest_gene <- function(chr, bp, gene_pos) {
+  chr <- as.character(chr)
+  closest_gene <-
+    gene_pos[V2 == chr
+    ][
+      abs(value - bp) == min(abs(value - bp)), V6]
+  return(closest_gene)
+}
+
+find_closest_gene <- Vectorize(find_closest_gene, 
+                               vectorize.args = c("chr", "bp"))
+
+snps_top[, Gene := find_closest_gene(CHR, BP)]

### 2022-07-27 code/gwas/ordinal/01_snp_list.R

@@ -28,4 +28,5 @@ for (i in seq_len(n_lists)) {
 }
 
 list_index <- data.table(1:n_lists)
-fwrite(list_index, file = "data/tmp/snp_list_index.tmp", row.names = F, col.names = F, quote = F, sep = " ")
+fwrite(list_index, file = "tmp/snp_list_index.tmp", 
+       row.names = F, col.names = F, quote = F, sep = " ")

### 2022-07-27 code/gwas/ordinal/polmm.sh

@@ -1,0 +1,22 @@
+
+phenos=$(cat data/mega/mega_np_ord.pheno | cut -d " " -f 6- | awk 'NR == 1{print}')
+
+for pheno in "vcid"
+  do
+  null_model=output/gwas/mega/polmm/"$pheno"_null_model.Rds
+  ./code/gwas/ordinal/polmm_null_model.R \
+    --covar data/mega/mega_np.covar \
+    --pheno data/mega/mega_np_ord.pheno \
+    --plink tmp/mega_np_prune \
+    --phenotype $pheno \
+    --out $null_model \
+    --ncores 100
+
+  ./code/gwas/ordinal/polmm_regression.R \
+    -m $null_model \
+    -l data/mega/SnpMatrix/SnpMatrix_list.Rds \
+    -o output/gwas/mega/polmm/"$pheno"_polmm_results.Rds
+  done
+
+
+

### 2022-07-27 code/gwas/ordinal/polmm_regression.R

@@ -1,0 +1,65 @@
+#! /usr/bin/Rscript --vanilla
+#=================================
+# POLMM SNP regression
+#=================================
+
+library(data.table)
+library(snpStats)
+library(POLMM)
+library(stringi)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("-m", "--model", 
+                    help = "path to POLMM null model .Rds file")
+parser$add_argument("-l", "--snp_matrix_list",
+                    default = "data/mega/SnpMatrix/SnpMatrix_list.Rds",
+                    help = paste("path to .Rds file of vector", 
+                                 "of SnpMatrix .Rds filenames"))
+parser$add_argument("-o", "--out", help = "path to text file to save results")
+args <- parser$parse_args()
+
+if (is.null(args$model)) {
+  stop("--model is a required argument")
+}
+
+if (!file.exists(args$model)) {stop(paste0(args$model, "doesn't exist"))}
+if (!file.exists(args$snp_matrix_list)) {
+  stop(paste0(args$snp_matrix_list, "doesn't exist"))
+}
+
+null_model <- readRDS(args$model)
+snp_mat_files <- readRDS(args$snp_matrix_list)
+results_list <- vector("list", length = length(snp_mat_files))
+
+for (i in 1:length(snp_mat_files)) {
+  mat <- readRDS(snp_mat_files[i])
+  g <- mat$genotypes@.Data
+  class(g) <- "numeric"
+  g <- g - 1
+  g[g == -1] <- NA
+  chrs <- mat$map$chromosome
+  polmm <- setDT(POLMM(
+    objNull = null_model,
+    Geno.mtx = g,
+    chrVec = chrs
+    )
+  )
+  results_list[[i]] <- polmm
+  rm(mat, g, chrs, polmm)
+}
+
+results <- rbindlist(results_list)
+
+results[, `:=`(pval.spa = as.numeric(pval.spa),
+               chr = as.integer(chr),
+               missing.rate = as.numeric(missing.rate),
+               Stat = as.numeric(Stat),
+               VarW = as.numeric(VarW),
+               VarP = as.numeric(VarP),
+               pval.norm = as.numeric(pval.norm),
+               MAF = as.numeric(MAF),
+               beta = as.numeric(beta),
+               switch.allele = as.logical(switch.allele))]
+
+saveRDS(results, file = args$out)

### 2022-07-27 code/gwas/ordinal_old/01_snp_list.R

@@ -1,31 +1,0 @@
-##---------------------------------------------------------------------
-## make lists of snps of length 30K for analysis blocks 
-##---------------------------------------------------------------------
-
-library(pacman)
-p_load(data.table, magrittr)
-# args <- commandArgs(trailingOnly = TRUE)
-
-# load snps from .bim file
-bim_file <- "data/plink/adc_np.bim"# args[1]
-bim <- fread(bim_file, header = F) %>% 
-  .[, .(V1, V2)] %>% 
-  setnames(., c("V1", "V2"), c("chr", "snp")) %>% 
-  .[, chr := factor(chr)]
-
-raw_length <- 30000 # desired approximate number of variants for each .raw file
-n_lists <- nrow(bim) %/% raw_length
-bim[, list_n := rep(1:n_lists, length.out=.N)]
-
-for (i in seq_len(n_lists)) {
-  fwrite(bim[list_n == i, .(snp)], 
-         file = paste0("data/tmp/ordinal_snp_list_", i, ".tmp"), 
-         row.names = F, 
-         col.names = F, 
-         quote = F, 
-         sep = " "
-  )
-}
-
-list_index <- data.table(1:n_lists)
-fwrite(list_index, file = "data/tmp/snp_list_index.tmp", row.names = F, col.names = F, quote = F, sep = " ")

### 2022-07-27 code/gwas/ordinal_old/03_regression.sh

@@ -1,9 +1,0 @@
-#!/bin/bash
-
-phenotype=$1
-
-<data/tmp/snp_list_index.tmp xargs -n 1 \
-  Rscript --vanilla --slave code/analysis/ordinal/03a_regression.R \
-    data/plink/adc_np.covar \
-    data/adc_np_ord.txt data/related_rm/"$phenotype".remove \
-    $phenotype

### 2022-07-27 code/gwas/ordinal_old/08_make.raw.top.snps.sh

@@ -1,15 +1,0 @@
-#!/bin/bash
-
-##--------------------------------------------
-## make .raw file for top snps for brant test
-##--------------------------------------------
-
-# clump top snps
-cat output/nacc/ordinal_results.txt | awk '{print$2,$7}' | awk '{if (NR == 1) print "SNP P"; else print $0}' > data/tmp/ordinal_results.tmp
-
-plink \
-  --bfile data/nacc/nacc \
-  --clump data/tmp/ordinal_results.tmp \
-  --clump-r2 0.05 \
-  --clump-p1 1e-5 \
-  --out output/nacc/ordinal

### 2022-07-27 code/heritability/format_sumstats_plink2.R

@@ -1,0 +1,94 @@
+#!/usr/bin/Rscript --vanilla
+
+#=====================================
+# Format plink .assoc.logistic file
+# into LDAK summary statistics format
+# Predictor A1 A2 Z
+#=====================================
+
+library(data.table)
+library(magrittr)
+library(stringi)
+library(rtracklayer)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("--sumstats")
+parser$add_argument("--out")
+parser$add_argument("--build", default="hg38")
+parser$add_argument("--tags", default="raw_data/ldak.thin.hapmap.gbr.tagging",
+                    help="Tagging file for LDAK heritability model")
+args <- parser$parse_args()
+
+results_path <- args$sumstats
+out_path <- args$out
+tags_path <- args$tags
+
+if (args$build %in% c("hg19", "hg38")) {
+  build <- args$build
+  print(paste("Variant positions are assumed to map to build:", build))
+  if (build == "hg38") {
+    print("Variants will be remapped to build hg19.")
+  }
+} else {
+  stop("Build must be either: hg19, hg38")
+}
+
+results <- fread(results_path)
+results[A1 != ALT, T_STAT := -T_STAT]
+results <- results[, .(`#CHROM`, POS, ID, REF, ALT, OBS_CT, T_STAT)]
+setnames(results, colnames(results), c('CHR', 'BP', 'SNP', 'A2', 'A1', 'n', 'Z'))
+
+
+
+# LDAK only accepts SNPs for genetic correlation
+snp_alleles <- c("A", "C", "T", "G")
+sumstats <- results[A1 %in% snp_alleles & A2 %in% snp_alleles]
+sumstats[, ChrBP := paste0(CHR, ":", BP)]
+sumstats <- sumstats[!(duplicated(ChrBP))]
+sumstats[, seq_name := paste0("chr", CHR)]
+
+# convert to hg19 build if current build is hg38
+if (build == "hg38") {
+  setnames(sumstats, "BP", "BP_hg38", skip_absent = TRUE)
+  
+  # commented out because of issues installing GEOquery
+  # copy over chain file
+  if (!file.exists("raw_data/hg38ToHg19.over.chain")) {
+    file.copy("/data_global/liftOver/hg38ToHg19.over.chain.gz",
[…]

### 2022-07-27 code/imputation/vcf_to_plink2.py

@@ -1,0 +1,38 @@
+#!/usr/bin/python3
+
+def vcf_to_plink2(vcf):
+    sys.stdout = sys.__stdout__
+    
+    cmd = ['plink2', '--vcf', vcf, '--geno', args.geno, \
+    '--maf', args.maf, '--make-pgen', "--out", args.out, '--vcf-half-call', 'm']
+    
+    subprocess.run(cmd)
+
+if __name__ == '__main__':
+
+    import argparse
+    import subprocess
+    import sys
+    import os
+
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--vcf', 
+                        required = True, 
+                        help = "vcf file path")
+    parser.add_argument('--geno', 
+                        default = '0.2', 
+                        help = "max fraction of people can be missing alleles")
+    parser.add_argument('--maf', 
+                        default = '0.001', 
+                        help = "minimum minor allele frequency") 
+    parser.add_argument('--out', 
+                        default = 'data/tmp/tmp', 
+                        help = "output fileset prefix")
+
+    args = parser.parse_args()
+    
+    cwd = os.getcwd()
+
+    vcf_file = args.vcf
+    
+    vcf_to_plink2(vcf_file)    

### 2022-07-27 code/magma/02_subset_pheno.R

@@ -1,0 +1,43 @@
+#! /use/bin/Rscript --vanilla
+#=============================================
+# Reformat and subset .pheno files for MAGMA
+#=============================================
+
+library(data.table)
+library(argparse)
+
+parser <- ArgumentParser()
+
+parser$add_argument('-p', '--pheno', help  = 'path to phenotype file')
+parser$add_argument('-r', '--remove', 
+                    help  = 'path to file with sample IDs to exclude')
+parser$add_argument('--phenotype', 
+                    help  = 'phenotype name')
+parser$add_argument('-o', '--out', help  = 'path to output file')
+
+
+args <- parser$parse_args()
+
+# #----------------------------------------------------
+# # test args
+# #----------------------------------------------------
+# args$pheno <- 'data/mega/mega_np_ord.pheno'
+# args$remove <- 'data/mega/related_rm/braak56.remove'
+# args$phenotype <- 'braak56'
+# args$out <- 'data/mega/magma/braak56.pheno'
+# #----------------------------------------------------
+
+phenotype <- args$phenotype
+pheno <- fread(args$pheno, colClasses = list(character = c('FID', 'IID')),
+               na.strings = '-1')
+pheno <- pheno[!is.na(get(phenotype))]
+rel_rm <- fread(args$remove, colClasses = c('character', 'character'))
+setnames(rel_rm, colnames(rel_rm), colnames(pheno)[1:2])
+
+
+pheno <- pheno[!rel_rm, on = names(rel_rm)]
+pheno <- pheno[, .(FID, IID, get(..phenotype))]
+pheno[, V3 := V3 + 1]
+setnames(pheno, 'V3', paste(phenotype))
+
+fwrite(pheno, file = args$out, sep = ' ', quote = FALSE)

### 2022-07-27 code/magma/magma.sh

@@ -1,0 +1,54 @@
+phenos=$(cat data/mega/mega_np.pheno | \
+         awk 'NR==1{for (i=3; i<NF; i++) printf $i "\n"; print $NF}')
+         
+# for pheno in $phenos
+# do
+#   Rscript --vanilla code/magma/02_subset_pheno.R \
+#     -p data/mega/mega_np.pheno \
+#     -o tmp/"$pheno".tmp \
+#     -r data/mega/related_rm/"$pheno".remove \
+#     --phenotype "$pheno" 
+# done
+
+nprocs=29
+# for pheno in $phenos
+# do
+#   seq 1 $nprocs | xargs -n 1 -P $nprocs -I % \
+#     magma \
+#     --batch % $nprocs \
+#     --bfile data/mega/mega_np \
+#     --pheno file=tmp/"$pheno".tmp \
+#     --covar file=data/mega/mega_np.covar \
+#     --gene-annot data/mega/magma/snp_gene_annot_hg38.genes.annot \
+#     --out tmp/"$pheno"_magma
+# 
+#   magma --merge tmp/"$pheno"_magma --out output/magma/"$pheno"
+# done
+
+# ordinal phenotypes
+phenos=$(cat data/mega/mega_np_ord.pheno | \
+         awk 'NR==1{for (i=3; i<NF; i++) printf $i "\n"; print $NF}')
+         
+for pheno in $phenos
+do
+  Rscript --vanilla code/magma/02_subset_pheno.R \
+    -p data/mega/mega_np_ord.pheno \
+    -o tmp/"$pheno".tmp \
+    -r data/mega/related_rm/"$pheno".remove \
+    --phenotype "$pheno"
+done
+
+for pheno in $phenos
+do
+  seq 1 $nprocs | xargs -n 1 -P $nprocs -I % \
+    magma \
+    --batch % $nprocs \
+    --bfile data/mega/mega_np \
+    --pheno file=tmp/"$pheno".tmp \
+    --covar file=data/mega/mega_np.covar \
+    --gene-annot data/mega/magma/snp_gene_annot_hg38.genes.annot \
+    --genes-only \
+    --out tmp/"$pheno"_magma
+
+  magma --merge tmp/"$pheno"_magma --out output/magma/"$pheno"
+done

### 2022-07-27 code/mvlmm/01_mega_mvlmm_adnc.sh

@@ -1,0 +1,39 @@
+cat output/gwas/mega/mega_adnc_pca.PC1.glm.linear | \
+  awk '$14 < 5e-4 {print $3}' \
+  > tmp/gemma_snps.tmp
+
+cat data/mega/mega_np_ord.pheno | \
+  awk '$3!=-1 && $4!=-1 && $5!=-1 && $10!=-1 && NR!=1 {
+       print$1,$2,0,0,0,$3,$4,$5,$10
+  }' \
+  > tmp/mega_mvlmm_adnc.txt
+
+plink2 \
+  --pfile data/mega/mega_np \
+  --keep tmp/mega_mvlmm_adnc.txt  \
+  --make-bed \
+  --out tmp/mega_mvlmm_adnc
+
+
+ldak5.2.linux \
+  --bfile tmp/mega_mvlmm_adnc \
+  --calc-kins-direct tmp/mega_mvlmm_adnc \
+  --kinship-gz YES  \
+  --ignore-weights YES  \
+  --power -0.25 \
+  --hwe-stand NO  \
+  --max-threads 32
+
+Rscript --vanilla code/mvlmm/01a_format_grm_covar.txt
+
+~/software/GEMMA-0.98.5/bin/gemma \
+  -bfile tmp/mega_mvlmm_adnc  \
+  -k tmp/mega_mvlmm_adnc_grm.txt  \
+  -km 2 \
+  -n 1 2 3 4  \
+  -lmm  \
+  -c tmp/mega_mvlmm_adnc.covar  \
+  -outdir ./tmp/ \
+  -o mega_mvlmm_adnc \
+  -snps tmp/gemma_snps.tmp
+

### 2022-07-27 code/np_coloc/loci.R

@@ -1,10 +1,11 @@
 library(data.table)
 library(magrittr)
 source('code/functions/plink2_to_plink_format.R')
+`%!in%` <- function(x, v) {!(x %in% v)}
 
-pheno <- fread("data/mega/mega_np.pheno")
+pheno <- fread("data/mega/mega_np.pheno", na.strings = '-1')
 phenotypes <- colnames(pheno)[3:length(colnames(pheno))]
-
+phenotypes <- phenotypes[phenotypes %!in% c("lewy")]
 file_path <- function(phenotype, prefix="output/gwas/mega") {
   paste0(prefix, '/mega.', phenotype, '.glm.logistic')
 }
@@ -21,10 +22,10 @@ for (phenotype in phenotypes) {
 }
 
 # 2. Create subsets of GWAS sumstats with P-values below threshold
-threshold <- 1e-4
+threshold <- 1e-5
 for (phenotype in phenotypes) {
 assign(paste0(phenotype, "_top"),
-       get(phenotype)[P < threshold])
+       get(phenotype)[CHR != 19][P < threshold])
 }
 
 # 3. Merge subset sumstats for phenotype pairs where OR are in same direction

### 2022-07-27 code/prep/adc/02_np_adgc.R

@@ -4,7 +4,7 @@ library(magrittr)
 
 na_strings <- c("-9", "-4", "9999", "999", "888")
 # Universal Data Set + NP
-uds <- fread("/data_global/nacc/investigator_nacc56.csv", 
+uds <- fread("/data_global/nacc/investigator_nacc57.csv", 
              na.strings = na_strings)
 # NP data not longitudinal, so just keep obs from last visit
 uds <- uds[NACCVNUM == NACCAVST]
@@ -13,6 +13,13 @@ uds <- uds[NACCVNUM == NACCAVST]
 mds <- fread("/data_global/nacc/fardo09062019.csv", 
              na.strings = na_strings)
 
+# list of neuropath variables
+np_vars_nacc <- fread("data/nacc_np_vars_names_nacc.txt", header = FALSE)$V1
+np_vars_remove <- fread("data/nacc_uds_var_names_np.txt", header = FALSE)$V1
+
+# PLINK .fam file
+adgc <- fread("data/adc/adc.psam", check.names = TRUE)
+setnames(adgc, "X.FID", "FID")
 #------------------------------------------------------
 # subset participants with NP data and NP variables
 #------------------------------------------------------
@@ -25,8 +32,7 @@ mds[!(is.na(NPFORMVER)), np := 1]
 table(mds$np, useNA = "a")
 
 np_vars <- colnames(uds)[grep("NP", colnames(uds))]
-np_vars_nacc <- fread("data/nacc_np_vars_names_nacc.txt", header = FALSE)$V1
-np_vars_remove <- fread("data/nacc_uds_var_names_np.txt", header = FALSE)$V1
+
 np_vars <- np_vars[!(np_vars) %in% np_vars_remove]
 np_vars <- c(np_vars_nacc, np_vars)
 uds_np <- uds[np == 1, ..np_vars]
@@ -53,11 +59,12 @@ np <- np[!((NACCID %in% np_dup_id$NACCID) & uds == 0)]
 # add variable indicating genotype data available in ADGC_HRC
 #-------------------------------------------------------------
 
-adgc <- fread("tmp/adc_no_naccid_dups.fam", header = FALSE)
-np_adgc <-  merge(np, adgc[, .(V1, V2)], by.x = "NACCID", by.y = "V2")
+
+np_adgc <- merge(np, adgc, by.x = "NACCID", by.y = "IID")
[…]

### 2022-07-27 code/prep/adc/05_remove_ethnic_outliers.R

@@ -21,8 +21,12 @@ merged <- fread("tmp/adc_1000g_merged_pca.eigenvec", header = F) %>%
 otg.merged <- merge(merged, otg, by = c("sample"), all = T)
 
 # reverse PCs if needed to make European pop top right of plot & label samples
-if (otg.merged[super_pop == "EUR", mean(PC1.norm)] < 0) {otg.merged[, `:=`(PC1 = -PC1, PC1.norm = -PC1.norm)]}
-if (otg.merged[super_pop == "EUR", mean(PC2.norm)] < 0) {otg.merged[, `:=`(PC2 = -PC2, PC2.norm = -PC2.norm)]}
+if (otg.merged[super_pop == "EUR", mean(PC1.norm)] < 0) {
+  otg.merged[, `:=`(PC1 = -PC1, PC1.norm = -PC1.norm)]
+  }
+if (otg.merged[super_pop == "EUR", mean(PC2.norm)] < 0) {
+  otg.merged[, `:=`(PC2 = -PC2, PC2.norm = -PC2.norm)]
+  }
 #otg.merged[is.na(super_pop), super_pop := "adc"]
 
 #ggplot of first 2 normalized PCs
@@ -38,8 +42,10 @@ pca.plot <- ggplot(otg.merged, aes(PC1.norm, PC2.norm, color = as.factor(super_p
 radius <- 0.38  #######
 #######################
 
-CircleFun <- function(center = c(otg.merged[super_pop == "EUR", mean(PC1.norm)],
-                                 otg.merged[super_pop == "EUR", mean(PC2.norm)]), 
+CircleFun <- function(center = c(otg.merged[super_pop == "EUR", 
+                                            mean(PC1.norm)],
+                                 otg.merged[super_pop == "EUR", 
+                                            mean(PC2.norm)]), 
                       r = 1, npoints = 100){
   tt <- seq(0,2*pi,length.out = npoints)
   xx <- center[1] + r * cos(tt)
@@ -50,7 +56,9 @@ CircleFun <- function(center = c(otg.merged[super_pop == "EUR", mean(PC1.norm)],
 ##-------------------
 ## Make plot
 ##-------------------
-pca.plot2 <- ggplot(otg.merged, aes(PC1.norm, PC2.norm, color = as.factor(super_pop))) +
+pca.plot2 <- ggplot(otg.merged, 
+                    aes(PC1.norm, PC2.norm, 
[…]

### 2022-07-27 code/prep/merge_plink2_filesets.py

@@ -1,0 +1,199 @@
+
+#============================
+# merge two plink filesets
+#============================
+
+def tmp_file(pfile, suffix=''):
+    return tmp_path + "/" + os.path.basename(pfile) + suffix
+
+def strip_a1a2(pfile, suffix):
+    # strips the A1:A2 suffixes from variant rsIDs
+    out_path = tmp_file(pfile, suffix)
+    with open(out_path, "w") as file_out:
+        cat_proc = subprocess.Popen(["cat", pfile + ".pvar"], 
+                                    stdout=subprocess.PIPE)
+        awk_proc = subprocess.Popen(["awk", 
+                                     "{print gensub(/:[ACTG]*:[ACTG]*/, \"\", 1)}"],
+                                    stdin=cat_proc.stdout,
+                                    stdout=file_out)
+        awk_proc.wait()
+
+def remove_dup_rsids(pfile, suffix, suffix1):
+    bim_file = tmp_file(pfile, suffix)
+    out_path = tmp_file(pfile, "_dup_rsids.tmp")
+    with open(out_path, "w") as file_out:
+        cat_proc = subprocess.Popen(["cat", bim_file], 
+                                    stdout=subprocess.PIPE)
+        cut_proc = subprocess.Popen(['awk', '{print $2}'], 
+                                    stdin=cat_proc.stdout,
+                                    stdout=subprocess.PIPE)
+        sort_proc = subprocess.Popen(['sort'], 
+                                    stdin=cut_proc.stdout,
+                                    stdout=subprocess.PIPE)
+        uniq_proc = subprocess.Popen(['uniq', '-d'], 
+                                    stdin=sort_proc.stdout,
+                                    stdout=file_out)
+        uniq_proc.wait()
+    
+    cmd = ['plink', '--pfile', pfile, '--bim', bim_file, '--make-bed',
+           '--out', tmp_file(pfile, suffix1)]
+
+    if os.stat(out_path).st_size != 0:
+        cmd += ['--exclude', out_path]
+
+    subprocess.run(cmd)
+
+def print_variants(pfile, suffix, suffix1):
+    with open(tmp_file(pfile, suffix)) as in_file:
+        with open(tmp_file(pfile, suffix1), 'w') as out_file:
+            subprocess.run(['awk', '{print$2}'],
+                           stdin=in_file,
+                           stdout=out_file)
[…]

### 2022-08-08 manuscript.Rmd
implement SAIGE for binary outcomes

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
+```
+
+```{r}
+library(data.table)
+library(flextable)
 ```
@@
 
+## Participants
+
+```{r}
+np <- readRDS("data/mega/mega_np.Rds")
+nadrc <- np[, length(table(NACCADC))]
+```
+
 We used genotype and NPE data from four aged autopsy cohorts: National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Memory and Aging Project (ROSMAP), Adult Changes in Thought (ACT), and the AD Neuroimaging Initiative (ADNI).
+All study participants were de-identified and deceased, and archival samples were exclusively used.
+Therefore, our study does not fall under the definition of "Human Subjects Research," according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- involving the collection/study of data or specimens if publicly availablr, or/or recorded such that subjects cannot be identified.
 
-## Quality Control
+This analysis used NACC data from `r nadrc` National Institute of Aging-funded Alzheimer's Disease Research Centers (ADRC).
+Individual ADRC use different recruitment methods and perform autopsies on-site, but participant data at each ADRC are collected using a standard form (<https://files.alz.washington.edu/documentation/np11-form.pdf>) and submitted to NACC where they are aggregated.
+The NACC Neuropathology (NP) data set based on the first version of this form was originally implemented in 2001[@besser2018], and this analysis uses data from then through the December 2021 freeze.
+Participants were excluded if they did not have autopsy data available or if they were noted in the NP data set to have at least one of 19 conditions that could potentially invalidate results.
+These conditions include brain tumors, severe head trauma, and fronto-temporal dementia (see **Supplementary Materials** for full list of variables used).
 
-Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS,[@shade2022] with the exception that all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021]
+ROSMAP has been previously described and consists of harmonized data from two longitudinal cohorts: The Religious Orders Study (ROS) and the Memory and Aging Project (MAP).[@bennett2018] ROS began in 1994 and has recruited over 1,100 Catholic priests, nuns, and brothers across the United States.
+MAP started in 1997 and enrolls community members in northeastern Illinois.
+The ROSMAP NP data used in this study was received from Rush University researchers in January 2020.
 
-1.  Retrieve participants labeled as non-Hispanic white or similar (European, etc.) if available.
-2.  Perform imputation on TOPMed Imputation Server (ROSMAP and ADNI had already been imputed using same reference panel by collaborators, so this step was skipped for those cohorts).
-3.  Covert VCF files to PLINK file set, keeping variants will minor allele frequencies (MAF) of 0.1% or greater.
-4.  Merge sub-cohorts of same study (e.g. NACC cohorts 1-12 and ACT 1-3).
-5.  Remove duplicate participants in each study.
-6.  Remove participants without autopsy data available.
-7.  Iteratively remove variants and participants with excess missingness until no variant is missing in more than 5% of participants and no participant is missing more than 5% of variants.
-8.  Remove participants with unusually high/low heterozygosity (+/- 3 SD cutoff).
-9.  Merge with 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012] and remove participants without primarily European ancestry as determined by principal component analysis.
-10.
-Remove participants with rare neurological diseases that substantially raise the likelihood that present NPE could be due to unusual pathological processes (NACC only, as the other cohorts did not have this information available).
-11. Remove duplicate participants between studies, preferentially keeping the observation in the larger cohort (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
-12. Merge cohorts.
-13. Harmonize NPE variables to analyze.
-14. Create list of related ($\ge$ 2nd-degree relatives) individuals using KING[@manichaikul2010] to remove for each NPE variable, randomly keeping one participant in each related cluster.
-15. Create genetic principal components using the PC-AiR method in the GENESIS package in R 4.1.[@conomos2015]
+The ACT study began in 1994 and recruited residents in the greater Seattle area aged 65 years and older without dementia at time of enrollment.21 The goal of the study has expanded to include three cohorts and continuous enrollment using the same enrollment criteria and has a current total of 4,960 participants across all three cohorts.
+The ACT NP data used in this study were obtained from Kaiser Permanente in May 2020.
 
-We note that there are substantial differences in the way that NPE are collected in different cohorts,(cite) and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes and minimizing computational load for analysis.
-Thus, some ordinal variables were collapsed to binary variables, or synthetic NPE may have been created using categorical variables from one cohort and continuous variables from another.
-All code used for harmonization will be made available upon publication, and a brief overview of the harmonization strategy for each NPE will be included in the Supplemental Materials.
+The ADNI (<https://adni.loni.usc.edu>) was launched in 2003 as a public-private research partnership, led by Principal Investigator Michael W.
+Weiner, MD.
+The primary goal of ADNI has been to test whether serial magnetic resonance imaging (MRI), positron emission tomography (PET), other biological markers, and clinical and neuropsychological assessment can be combined to measure the progression of mild cognitive impairment (MCI) and early AD.
+A subset of ADNI participants undergo autopsy and receive neuropathological phenotyping.
+The ADNI NP data used in this study was downloaded from the ADNI website in October 2021.
+
+## Genotype data and quality control
+
+Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS.[@shade2022] Genotype data for all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021] NACC and ACT raw genotype data were obtained from the September 2020 freeze of the Alzheimer's Disease Genetics Consortium (ADGC) and subsequently imputed, while pre-imputed ROSMAP and ADNI genotype data were received from collaborators in the Hohman lab at Vanderbilt University in March 2021 and December 2021, respectively.
+
+Quality control on genotype data was performed separately for each cohort.
+Firstly, only participants labeled as non-Hispanic white or similar (European, etc.) were used if available.
+After imputation, duplicate samples and samples without autopsy data available were removed.
+Genetic variants and participants were iteratively removed under no variants were missing in more than 5% of participants and no participants were missing more than 5% of variants (however, average genotype coverage was 99.7%, and no variants or participants were actually removed during this process).
+Then, participants with unusually high or low (+/- 3 standard deviations from mean) were removed.
+Finally, participants were merged with the 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012], and participants with substantial non-European ancestry, as determined by distance from EUR superpopulation centroid in principal component analysis using the first two principal components, were removed.
+Some participants had NP and genotype data available in more than one cohort.
+In these cases, records in the larger cohort were preferentially kept and discarded in other cohorts (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
+
+## Defining and harmonizing NPE
+
+We note that there are substantial differences in the way that NPE are collected in different cohorts, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
+Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another.
+All code used for harmonization is available at <https://www.github.com/lincoln-shade/np_phewas>, and an overview of the harmonization strategy for each NPE is included in the **Supplemental Materials**.
 
@@
 
-All the analyses discussed below were done on the data set with the combined cohorts.
-
 ### Single-variant GWAS
 
-Single-variant logistic regression analyses were performed in PLINK 2.0.[@chang2015; @purcell] Covariates included age at death, sex, cohort, and the first 10 genetic principle components.
-Continuous covariates were standardized to a mean of 0 and a variance of 1.
+Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package,[@bi2021] while binary variables were analyzed with logistic mixed-effects models using the SAIGE R package.[@zhou2018] Covariates included age at death, sex, cohort, and the first 10 genetic principle components created using the PC-AiR method in the GENESIS R package.[@conomos2015].
+Dense genetic relationship matrices (GRM) created using a pruned set of independent ($r^2 < 0.2$) were used to account for relatedness between participants.
 An additive mode of inheritance was assumed in all analyses.
-Variants achieving a p-value of $< 5\times 10^{-8}$ were considered "genome-wide" significant.
-To identify independent genetic risk loci, variants were clumped using PLINK 1.9[@chang] using a linkage-disequilibrium threshold of $r^2 \le 0.05$.
+Analyses proceeded in two stages: in stage one, null models with fixed covariates and GRM were fitted using either POLMM or SAIGE.
+In stage 2, score tests were performed on each variant with saddle-point approximation used to calculate p-values.
+Variants achieving a p-value of $p < 5\times 10^{-8}$ were considered "genome-wide" significant.
+To identify independent risk loci, variants were clumped using PLINK 1.9[@chang] with a linkage-disequilibrium threshold of $r^2 \le 0.05$.
 
-Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package.[@bi2021] Dense kinship matrices were used to account for relatedness.
+For NPE available in multiple cohorts, single-variant analyses were first performed in NACC participants.
+Lead variants with p-values exceeding a suggestive threshold of $p < 1 \times 10^{-6}$ were then tested in ROSMAP, ACT, and/or ADNI as available for attempted validation.
+Participants were then pooled for genome-wide mega-analysis.
+
+To examine whether variants in the APOE region independent of $\epsilon$ haplotypes influence NPE risk, we re-analyzed Chromosome 19 using APOE $\epsilon$ diplotype as an additional covariate for NPE with significant association signals within the APOE locus.
+APOE diplotypes were determined using the rs7412 and rs429358 variants.
+
+### Multivariate analyses
+
+Because many NPE have high phenotype correlation that single-outcome association analysis cannot take into account, we sought to perform genetic association analyses that could better account for the common co-occurrence of of NPE.
+We first assessed the phenotype correlation of NPE used in the pooled GWAS using polychoric correlation and grouped NPE visually using dendrograms.&&&
+
+### Colocalization analyses
+
+All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues and in ROSMAP bulk RNA-seq data in the dorsolateral prefrontal cortex.
+Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs.
+For ordinal variables, cut points widely used in previous studies were chosen to determine case-control proportions.
+A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) $> 80\%$ was used as a threshold for evidence of colocalization.
+
+To investigate whether shared GWAS signals drive association between multiple NPE, colocalization analysis was also performed on loci with variants exceeding the suggestive threshold of $p < 1 \times 10^{-6}$ for more than one NPE in the pooled mega-analytic GWAS.
 
@@
 For genetic loci associated with multiple NPE, we considered mediation hypotheses to test whether the association of the lead variant with one NPE was mediated by another NPE.
-Regression analyses were run in R using the `glm` or `polr` functions and mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.
+Regression analyses were run in R using the `glm` function for binary outcome traits or `polr` function in the MASS[@venables2002] package for ordinal traits.
+Mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.
+To assess the possibility of reverse causation, mediation analyses were performed with both NPE in each pair as outcomes.
 
-### Colocalization analyses
+# Results
 
-All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues.
-Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs.
-A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) \> 80% was used as a threshold for evidence of colocalization.
+```{r}
+covar <- fread("data/mega/mega_np.covar")
+nmega <- formatC(covar[, .N], big.mark = ",")
+nnacc <- formatC(covar[FID == "0", .N], big.mark = ",")
+nrosmap <- formatC(covar[FID == "1", .N], big.mark = ",")
+nact <- formatC(covar[FID == "2", .N], big.mark = ",")
+nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")
+```
+
+```{r}
+t1_env <- new.env()
+source("code/manuscript/01_table1_participant_demographics.R",
+       local = t1_env)
+t1dt <- t1_env$t1dt
+```
+
+In total, `r nmega` participants across all cohorts underwent autopsy, had genotype data available, and passed quality control measures.
+This included `r nnacc` NACC, `r nrosmap` ROSMAP, `r nact` ACT, and `r nadni` ADNI participants.
+Participant demographics are shown in **Table 1**.
+NACC participants were younger (mean age-of-death 81 years) compared to ROSMAP (90 years) and ACT (88 years) participants, and had more balanced sex ratios, with 50% of NACC participants female vs.
+68% and 58% in ROSMAP and ACT, respectively.
+NACC participants were also more likely to have an APOE $\epsilon 4$ allele (55%) vs.
+ROSMAP (25%) or ACT (28%).
+Across the 12 NPE harmonized across all four cohorts, NACC and ADNI tended to show more advanced neuropathology.
+Using CERAD neuritic plaque score as an example, 9% of NACC participants were rated as "none" for CERAD score, whereas 24% and 23% of ROSMAP and ACT participants were rated as "none," respectively.
+At the other extreme, 65% of NACC participants were rated as having "severe" neuritic plaque pathology, whereas only 33% and 27% of ROSMAP and ACT participants, respectively, were rated as such.
+This trend can be seen across **XX** of the NPE harmonized across cohorts and is shown in **Supplementary Table S2**.
+
+## Single-variant analysis
+
+# Tables
+
+```{r}
+t1_env$t1
+```

### 2022-08-08 code/SAIGE/step1_fitNULLGLMM.R

@@ -1,0 +1,252 @@
+#!/usr/bin/env Rscript
+
+options(stringsAsFactors=F)
+
+## load R libraries
+library(SAIGE)
+require(optparse) #install.packages("optparse")
+
+print(sessionInfo())
+
+## set list of cmd line arguments
+option_list <- list(
+  make_option("--plinkFile", type="character",default="",
+    help="Path to plink file for creating the genetic relationship matrix (GRM). minMAFforGRM can be used to specify the minimum MAF and maxMissingRate can be used to specify the maximum missing rates  of markers in the plink file to be used for constructing GRM. Genetic markers are also randomly selected from the plink file to estimate the variance ratios"),
+  make_option("--bedFile", type="character",default="",
+    help="Path to bed file. If plinkFile is specified, 'plinkFile'.bed will be used"),
+  make_option("--bimFile", type="character",default="",
+    help="Path to bim file. If plinkFile is specified, 'plinkFile'.bim will be used"),
+  make_option("--famFile", type="character",default="",
+    help="Path to fam file. If plinkFile is specified, 'plinkFile'.fam will be used"),
+  make_option("--phenoFile", type="character", default="",
+    help="Required. Path to the phenotype file. The file can be either tab or space delimited. The phenotype file has a header and contains at least two columns. One column is for phentoype and the other column is for sample IDs. Additional columns can be included in the phenotype file for covariates in the null model. Please specify the names of the covariates using the argument covarColList and specify categorical covariates using the argument qCovarColList. All categorical covariates must also be included in covarColList."),
+  make_option("--phenoCol", type="character", default="",
+    help="Required. Column name for phenotype to be tested in the phenotype file, e.g CAD"),
+  make_option("--traitType", type="character", default="binary",help="Required. binary or quantitative [default=binary]"),
+  make_option("--invNormalize", type="logical",default=FALSE,
+    help="Optional. Only for quantitative. Whether to perform the inverse normalization for the phenotype [default='FALSE']"),
+  make_option("--covarColList", type="character", default="",
+    help="List of covariates (comma separated)"),
+  make_option("--qCovarColList", type="character", default="",
[…]

### 2022-08-08 code/manuscript/01_table1_participant_demographics.R

@@ -1,0 +1,50 @@
+
+library(data.table)
+library(magrittr)
+library(flextable)
+library(table1)
+
+pord <- fread("data/mega/mega_np_ord.pheno", na.strings = '-1')
+pbin <- fread("data/mega/mega_np.pheno", na.strings = '-1')
+pheno <- merge(pord, pbin, c("FID", "IID"))
+covar <- fread("data/mega/mega_np.covar")
+covar_apoe <- fread("data/mega/mega_np_apoe.covar", stringsAsFactors = TRUE)
+covar <- merge(covar, 
+               covar_apoe[, .(FID, IID, apoe)], 
+               c("FID", "IID"), 
+               all.x = TRUE)
+covar <- covar[, .(FID, IID, msex, age_death, apoe)]
+covar[, Sex := factor(msex, labels = c("Female", "Male"))]
+
+dt <- merge(pheno, covar, by = c("FID", "IID"))
+setnames(dt, "age_death", "Age of Death")
+dt[, Study := factor(FID, labels = c("NACC", "ROSMAP", "ACT", "ADNI"))]
+dt[, CERAD := ordered(cerad, 
+                        labels = c("None", "Mild", "Moderate", "Severe"))]
+
+# Display apoe variable as number of APOE e4 alleles for table
+dt[apoe %in% c("apoe22", "apoe23", "apoe33"), 
+     `APOE e4 alleles` := 0]
+dt[apoe %in% c("apoe24", "apoe34"), 
+     `APOE e4 alleles` := 1]
+dt[apoe %in% c("apoe44"), 
+     `APOE e4 alleles` := 2]
+dt[, `APOE e4 alleles` := ordered(`APOE e4 alleles`)]
+
+t1dt <- 
+  as.data.frame(
+    table1(~ Sex + `Age of Death` + `APOE e4 alleles` + CERAD | Study,
+           data = dt
+           )
+    ) 
+
+t1 <- flextable(t1dt) %>% 
+  autofit() %>% 
+  padding(padding = 2, part = "all") %>%
+  add_header_lines("Table 1: Participant Demographics") %>% 
+  add_footer_lines("Selected NPE are shown in Table 1. All NPE summaries are shown in Supplementary Materials Table S2.") %>% 
+  add_footer_lines("Key: SD, standard deviation; Min, minimum; Max, maximum. Age of Death variable is integer for NACC, ADNI, and ACT but continuous in ROSMAP.")
+
+saveRDS(t1, 
[…]

### 2022-08-08 code/multivar/pheno_cors.R

@@ -6,9 +6,11 @@
 
 library(data.table)
 library(magrittr)
+library(polycor)
 
 bin_pheno <- fread("data/mega/mega_np.pheno", na.strings = "-1")
 ord_pheno <- fread("data/mega/mega_np_ord.pheno", na.strings = "-1")
+ord_pheno[, `:=`(vcid  = NULL, part_def = NULL)]
 pheno <- merge(ord_pheno, bin_pheno[, .(FID, IID, hs, grossinf, microinf)],
                by = c('FID', 'IID'))
 # pheno[, hs := NULL]
@@ -25,7 +27,7 @@ for (i in colnames(pheno[, 3:ncol(pheno)])) {
 
 pheno_4k <- character()
 for (i in colnames(pheno)) {
-  if (pheno[!(is.na(get(i))), .N] >= 1000) {
+  if (pheno[!(is.na(get(i))), .N] >= 2000) {
     pheno_4k <- c(pheno_4k, i)
   }
 }
@@ -34,12 +36,16 @@ pheno_4k <- pheno_4k[complete.cases(pheno_4k)]
 
 n_cols <- ncol(pheno_4k) - 2L
 cor_mat <- matrix(nrow = n_cols, ncol = n_cols)
-colnames(cor_mat) <- rownames(cor_mat) <- colnames(pheno_4k[, -c("FID", "IID")])
+colnames(cor_mat) <- 
+  rownames(cor_mat) <- 
+  colnames(pheno_4k[, -c("FID", "IID")])
 for (i in 1L:n_cols) {
   pheno1 <- colnames(pheno_4k)[i + 2L]
   for (j in i:n_cols) {
     pheno2 <- colnames(pheno_4k)[j + 2L]
-    cor_mat[i, j] <- cor_mat[j, i] <- pheno_4k[, cor(get(..pheno1), get(..pheno2), use = "c", method = 's')]
+    cor_mat[i, j] <- cor_mat[j, i] <- 
+      pheno_4k[, polychor(get(..pheno1), 
+                          get(..pheno2))]
   }
 }
 
@@ -47,6 +53,7 @@ cor_mat[is.na(cor_mat)] <- 0
 cor_mat <- round(cor_mat, 2)
 
 pca1 <- prcomp(pheno_4k[, -c("FID", "IID")], scale. = T, center = T)
+pca2 <- princomp(cor_mat)
 pcs <- data.table(FID = pheno_4k$FID, 
                   IID = pheno_4k$IID, 
                   pca1$x)

### 2022-08-08 code/prep/mega/02_make_pheno.R

@@ -191,9 +191,10 @@ mega_np[!is.na(vcid) & (arteriol > 2 | atheroscler > 2), vcid := 3]
 mega_np[!is.na(vcid) & microinf == 1, vcid := 4]
 mega_np[!is.na(vcid) & grossinf == 1, vcid := 5]
 
-plink_pheno_vars <- c(colnames(pheno), 'diffuse_abeta3', 'late23', 
-                      'lewy_body123', 'part_def34')
-plink_pheno_vars <- plink_pheno_vars[plink_pheno_vars != "lewy"]
+plink_pheno_vars <- c("FID", "IID", "hs", "microinf", "grossinf")
+# plink_pheno_vars <- c(colnames(pheno), 'diffuse_abeta3', 'late23', 
+                      # 'lewy_body123', 'part_def34')
+# plink_pheno_vars <- plink_pheno_vars[plink_pheno_vars != "lewy"]
 
 fwrite(mega_np[, ..plink_pheno_vars], 
        file = "data/mega/mega_np.pheno", 

### 2022-08-16 manuscript.Rmd
add tables for manuscript, write more on manuscript, write SAIGE scripts for NACC, write POLMM for ROSMAP, ACT, and ADNI

@@
 
-Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinical AD; however, clinical AD does not always correspond with AD neuropathology and vice-versa.[@nelson2012] We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:
+Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinically diagnosed AD; however, clinically diagnosed AD is increasingly recognized as a heterogeneous disease that can reflect multiple independent or co-occurring pathological processes.
+Aged individuals in research autopsy cohorts often have more than one proteinopathy present in their brains at time of death.[@karanth2020].
+Vascular factors, especially small-vessel disease, also contributes to cognitive decline.
+
+does not always correspond with AD neuropathology and vice-versa.[@nelson2012]
+
+We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:
 
@@
 
-## Defining and harmonizing NPE
+## Defining and harmonizing NPE for analysis
 
-We note that there are substantial differences in the way that NPE are collected in different cohorts, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
+We note that there are substantial differences in the way that some NPE are collected in different cohorts, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
 Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another.
-All code used for harmonization is available at <https://www.github.com/lincoln-shade/np_phewas>, and an overview of the harmonization strategy for each NPE is included in the **Supplemental Materials**.
+All code used for harmonization is available at <https://www.github.com/lincoln-shade/np_phewas>.
+
+In total, eleven NPE were harmonized across the four cohorts used.
+This included eight ordinal outcomes and three binary outcomes.
+Ordinal outcomes harmonized were brain arteriolosclerosis, atherosclerosis in the circle of Willis, Braak Stage for neurofibrillary tangles (NFT), cerebral amyloid angiopathy, Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, diffuse amyloid-beta plaques, limbic-predominant age-related TDP-43 encephalopathy neuropathologic change (LATE-NC), and Lewy body pathology.
+Braak Stage ranges from zero (absent NFT) to six (diffuse NFT throughout cortex and large loss of neurons).[@braak1991] LATE-NC had four levels, with "None" indicating lack of TDP-43 proteinopathy, "Mild" indicating TDP-43 deposits in the amygdala only, "Moderate" indicating deposits in the amygdala and limbic system, and "Severe" indicating deposits in amygdala, limbic system, and neocortex.
+Except where noted otherwise, other ordinal NPE were coded in four levels, "None," "Mild," "Moderate," or "Severe." Diffuse amyloid plaque pathology followed the Thal staging of amyloid deposition in NACC, ACT, and ADNI.
+In ROSMAP, diffuse plaques are examined and quantified in five regions (midfrontal cortex, entorhinal cortex, inferior parietal cortex, and hippocampus), then scaled by each regions standard deviation and then averaged.
+ROSMAP participants were labeled as "None" if their averaged score was equal to zero, "Mild" if their score was higher than zero but $\le$ 0.5, "Moderate" if their score was between 0.5 and one, and "Severe" if their score was above one.
+These labels roughly corresponded to score quartiles.
+
+The three binary NPE outcomes included hippocampal sclerosis (HS), gross infarctions, and microinfarcts.
+Participants were labeled as cases for HS if they have unilateral or bilateral HS, cases for gross infarctions in they have at least one chronic or acute gross infarction, and cases for microinfarcts if they had one or more microinfarcts in any region observed.
 
@@
 
-To examine whether variants in the APOE region independent of $\epsilon$ haplotypes influence NPE risk, we re-analyzed Chromosome 19 using APOE $\epsilon$ diplotype as an additional covariate for NPE with significant association signals within the APOE locus.
-APOE diplotypes were determined using the rs7412 and rs429358 variants.
+To examine whether variants in the *APOE* region, defined as the region with 200 kilo-basepairs (KBP) from the start and end of the APOE coding region, independent of *APOE* $\epsilon$ haplotypes influence NPE risk, we re-analyzed Chromosome 19 using *APOE* $\epsilon$ diplotype as an additional covariate for NPE with significant association signals within the *APOE* locus in the main mega-analysis.
+*APOE* diplotypes were determined using the rs7412 and rs429358 variants according to the SNPedia online reference.[@apoe-s] Diplotype $\epsilon3/\epsilon3$ was used as the reference, while $\epsilon2/\epsilon2$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, and $\epsilon4/\epsilon4$ were adjusted for using fixed effect indicator variables.
+For NPE with multiple independent ($r^2 < 0.05$) association signals in the *APOE* region in the main analysis not adjusting for *APOE* $\epsilon$ diplotype, only the most significant signal was reported in our results.
+Additional signals were only reported if they were significant in the *APOE*-adjusted models.
 
@@
 Because many NPE have high phenotype correlation that single-outcome association analysis cannot take into account, we sought to perform genetic association analyses that could better account for the common co-occurrence of of NPE.
-We first assessed the phenotype correlation of NPE used in the pooled GWAS using polychoric correlation and grouped NPE visually using dendrograms.&&&
+We first assessed the phenotype correlation of NPE used in the pooled GWAS using polychoric correlation and grouped NPE visually using a dendrogram.
+Principal components analysis (PCA) was then performed using the psych R package on each group of NPE.
+The first PC (PC1) of each group was then used as the outcome variable for GWAS using SAIGE in a two-stage analysis.
+In the first stage, PC1 was regression on fixed-effects covariates using the GRM to account for relatedness.
+Because PC1 was not normally distributed, the residuals of stage one were then rank inverse normalized and regressed on covariates and GRM again in stage two.
+Score tests were then performed on genetic variants and saddle-point approximation used to estimate p-values.
+
+### Gene-based analyses
+
+Gene start and end sites were determined using NCBI version 38 gene regions.
+Variants were mapped to a gene if they were located within 10 KBP of the gene's start or end sites.
+Using an unrelated set of the pooled participant data set, gene-based analyses were peformed using MAGMA v1.10.[@leeuw2015] The same fixed-effects covariates used in single-variant analyses were used, and the top variant principal components that accounted for 99.9% of variance in a gene's region were used to test for significance using an F-test.
+Genes with $p < 2.5 \times 10^{-6}$ were considered significantly associated with NPE.
 
@@
 
-All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues and in ROSMAP bulk RNA-seq data in the dorsolateral prefrontal cortex.
-Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs.
+All variants in each logistic GWAS with p-values $< 1 \times 10^{-5}$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues and in ROSMAP bulk RNA-seq data in the dorsolateral prefrontal cortex.
+Colocalization analysis using the coloc R package was then performed on GWAS loci with significant QTLs.
 For ordinal variables, cut points widely used in previous studies were chosen to determine case-control proportions.
@@
 
-To investigate whether shared GWAS signals drive association between multiple NPE, colocalization analysis was also performed on loci with variants exceeding the suggestive threshold of $p < 1 \times 10^{-6}$ for more than one NPE in the pooled mega-analytic GWAS.
+To investigate whether shared GWAS signals drive association between multiple NPE, colocalization analysis was also performed on loci with variants exceeding a p-value threshold of $p < 1 \times 10^{-4}$ and concordant effect direction for more than one NPE in the pooled mega-analytic GWAS.
 
@@
 ROSMAP (25%) or ACT (28%).
-Across the 12 NPE harmonized across all four cohorts, NACC and ADNI tended to show more advanced neuropathology.
+Across the 13 NPE harmonized across multiple cohorts, NACC and ADNI tended to show more advanced neuropathology than ROSMAP and ACT.
 Using CERAD neuritic plaque score as an example, 9% of NACC participants were rated as "none" for CERAD score, whereas 24% and 23% of ROSMAP and ACT participants were rated as "none," respectively.
@@
 
+```{r}
+t2_env <- new.env()
+source("code/manuscript/02_table2_mega_analysis_results.R",
+       local = t2_env)
+```
+
+In the NACC analysis, no variants outside of the APOE region were significantly associated with any NPE.
+Variants within this region were significantly associated with Braak stage, CERAD score, CAA, and diffuse amyloid plaques.
+Variants reaching a suggestive significance threshold ($p < 1 \times 10^{-5}$ and replicating at a nominal significance threshold of $p < 0.05$ are shown in Table 2 below.
+Variants outside og the APOE region that reached a type-I error rate below $0.05$ using Benjamini-Hochberg adjustment included &&&.
+
+In total, fourteen independent ($r^2 < 0.05$) loci had lead variants that exceeded the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis.
+Of these, four were duplicate signals in the *APOE* region, leaving ten significant signals across 13 NPE.
+Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE*-adjusted analysis.
+Novel variant-NPE associations included a variant 12 KBP upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis and a *PIK3R5* intronic variant, rs72807981, associated with Braak stage (**Table 3**).
+Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak stage; and *TMEM106B* with HS and LATE-NC.
+
+## Multivariate analysis
+
+Visual inspection of a dendrogram (**Supplementary Figure S2**) of NPE based on polychoric correlations reveals two primary clusters of NPE: an AD NPE cluster consisting of Braak stage, CERAD score, diffuse amyloid, and CAA; and a vascular NPE cluster consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts.
+PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster.
+Only the *APOE* region was significantly associated with the AD cluster PC1, while no loci were significantly associated with the vascular PC1.
+
+## Gene-based analysis
+
+## Colocalization analyses
+
+Because CAA was the only NPE to have an associated locus in the *APOE* region with a signal independent of *APOE* $\epsilon$ diplotype, this region was exluded from NPE-NPE colocalization analysis.
+Pairs of NPE with the same variant reaching a p-value threshold of $p < 1 \times 10^{-5}$ with concordant effect directions included Braak stage and CERAD score in the *BIN1* region on Chromosome 2 and HS and LATE-NC in the *TMEM106B* region on Chromosome 7.
+Both loci showed a high probability of colocalization, with both having a PPH4 $> 99.9%$.
+
+Multiple suggestive NPE loci showed evidence of colocalization with gene expression QTL (eQTL) in GTEx.
+A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with GRN expression (PPH4 $>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL.
+The CAA-associated *APOE* locus indepedent of *APOE* $\epsilon$ diplotype also colocalized with *APOE* eQTL in GTEx (PPH4 $= 95\%$) and with a methylation QTL (mQTL) in ROSMAP (PPH4 $>99.9\%$).
+
+## Conditional and mediation analyses
+
+We performed conditional and mediation analyses on both loci that colocalized between two NPE: *BIN1*, which colocalized with Braak stage and CERAD score; and *TMEM106B*, which colocalized with HS and LATE-NC.
+In unconditional analyses (the single-variant analyses discussed above), the lead *BIN1* variant, rs6733839, had similar effect sizes for both Braak stage ($p = 3.8 \times 10^{-10}$, OR $= 1.21$) and CERAD score ($p = 6.6 \times 10^{-8}$, OR $= 1.20$).
+However, when Braak stage was included as a covariate for CERAD, the association between rs6733839 and CERAD score was markedly reduced (OR $= 1.03$, $p = 0.47$).
+In contrast, the association between rs6733839 and Braak stage was less attenuated when adjusted for CERAD score (OR $= 1.14$, \$p = \$).
+In the mediation analysis, we found that Braak stage mediated 87% of the effect of rs633839 on CERAD score, and the post-mediation effect size was not significantly different than zero.
+For the *TMEM106B* analysis, we used the lead variant for HS, rs4721058, which was significantly associated with both HS (OR $= 0.67$, $p = 1.7 \times 10^{-11}$) and LATE-NC (OR $= 0.71$, $p = 1.7 \times 10^{-8}$).
+Mediation analysis resulted in moderate mediation (\~50%) for both HS and LATE-NC.
+
+# Discussion
+
+We performed GWAS and functional analyses for eleven neuropathology endophenotypes across four high-quality neuropathology cohorts with a maximum sample size of 7,463 participants.
+Our work builds on previous genetic association studies of NPE, Alzheimer's, and related dementias and provides another attempt to better understand the complex associations between different neuropathologies and genetic risk factors.
+We discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively.
+We also investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
+However, our study has several important limitations, including sample size and heterogeneity in design between cohorts.
+
+## *COL4A1* and atherosclerosis
+
+One locus on Chromosome 12q34 with lead variant rs2000660 (MAF $= 9.9\%$ in study participants) located 12 KBP upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis.
+*COL4A1* codes for a component of collagen IV, an important component of basal lamina.
+The *COL4A1/COL4A2* locus has been previously found to be associated with numerous other vascular phenotypes, including peripheral artery disease[@klarin2019], coronary artery disease[@nikpay2015], stroke(cite), and arteriol stiffness(cite) through GWAS.
+Additionally, the locus has been associated with rare familial cerebrovascular diseases[@blevins2021] and lacunar ischemic stroke[@rannikmae2017] rs2000660 in particular has recently been found to be a risk variant for migraine.[@hautakangas2022] The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to cerebrovascular disease is likely related to disruption of the extracellular matrix.(cite) rs2000660 did not show evidence of colocalization in GTEx or ROSMAP, so the functional effect of this variant, if indeed it is has one, is not obvious from our results.
+rs2000660 was located just outside the 10 KBP window used for gene-based analysis, and the *COL4A1* and *COL4A2* genes were not nominally associated with atherosclerosis in our study ($p>0.05$).
+Furthermore, rs2000660 was not nominally associated with any other vascular NPE in our study.
+A previous GWAS in ROSMAP of cerebral atherosclerosis did not identify the *COL4A1* as a risk locus.[@vattathil2021] However, this sample size of this study was significantly smaller than the one used in the present study (1,325 vs 6,959).
+Indeed, rs2000660 only reached genome-wide significance in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0081$).
+
+## *PIK3R5* and Braak stage
+
+The other identified novel genome-wide significant locus located on Chromosome 17p13 had a lead variant of rs72807981 (MAF $= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival.
+One study which compared differential mRNA expression in aged adults with Braak stage VI vs.
+non-demented controls found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $= 0.01$).[@guennewig2021]
+
+## *APOE* and NPE
+
+Multiple variants in the APOE region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
+Additionally, APOE was also associated with LATE-NC.
+This result was unsurprising and consistent with previous GWAS of NPE.[@beecham2014 @dugan2021] While multiple loci in this region were significantly associated with each of these four NPE, only one locus with lead variant rs4803778 remained significant (OR $= 1.24$, $p = 5.8 \times 10^{-12}$) when adjusting for *APOE* $\epsilon$ diplotype.
+rs4803778 was in high LD ($r^2 = 0.97$) with rs4803779, which was also significantly associated with CAA (OR $= 1.24$, $p = 6.7 \times 10^{-12}$) and is a lead mQTL in ROSMAP for site cg09555818 ($p = 3.3 \times 10^{-91}$), indicating that this locus may be associated with CAA through epigenetic regulation of *APOE*.
+However, this locus is located approximately 30 KBP downstream of the *APOE* coding end site and situated between the *APOC4*, *APOC2*, and *CLPTM1*, and is a stronger eQTL of *APOC2* than *APOE*, although it is strongly significant and colocalizes with expression of both genes in multiple tissues.
+
+## *BIN1*, neurofibrillary tangles, and neuritic plaques
+
+An locus approximately 30 KBP downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage (OR $= 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques (OR $= 1.20$, $p = 6.6 \times 10^{-8}$).
+This locus is a known AD risk locus and is second only to *APOE* for strength of association with late-onset AD.[@bellenguez2022] However, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques (OR $= 1.05$, $p = 0.15$) nor CAA (OR $= 1.02$, $p = 0.49$).
+Previous research has identified that *BIN1* variant rs744373 is associated with tau-PET but not amyliod-PET levels[@franzmeier2019] and that *BIN1* protein expression is correlated with neurofibrillary table pathology but not with diffuse or neuritic plaques.[@holler2014] While the *BIN1* locus was associated with both neurofibrillary tangle and neuritic plaque pathology in our sample, neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles[@nelson2009], indicating that pathological process related to tau could mediate the association.
+We therefore performed mediation analysis to test the hypothesis that the association between the *BIN1* locus and neuritic plaques is mediated by overall tau burden, using Braak staging of neurofibrillary tangles as a proxy measurement.
+We found that tau burden completely mediated the association between BIN1 and neuritic plaques, as the association between *BIN1* and neuritic plaques was not longer significant after adjusting for neurofibrillary tangle pathology (OR = 1.03, p = 0.48, see **Figure X**).
+Our findings are therefore consistent with the hypothesis that the relationship between *BIN1* and Alzheimer's disease is mediated by tau rather than amyloid pathological processes.
+
+## Limitations and conclusion
+
+There are multiple limitations to the approach we took in this study.
+The four cohorts we used, while all high quality, used different recruitment and design considerations.
+NACC recruitment is handled independently at each ADRC and likely consists of a more clinically representative population, whereas MAP and ACT particpants may be more representative of the broader aged communities from which they are recruited.
+ROS recruits Catholic brothers and sisters and may represent a population more highly educated than the United States as a whole.
+These differences are reflected in the relative frequencies and distributions of NPE between cohorts, with NACC and ADNI tending to show more severe pathology than ACT and ROSMAP (**Tables 1 and S2**).
+Additionally, there is likely substantially heterogeneity in the way that NPE data are collected and graded both within and between studies.
+We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis.[@shade2022]
+
+Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is small relative to their clinical counterparts.
+A recent GWAS of AD included over 111,000 clinically diagnosed or "proxy" (based on family history) AD cases and nearly 700,000 controls.
+In contrast, our largest analyzed sample size for CERAD score totalled 7,447 participants, while LATE-NC totaled only 2,224.
+These sample reflect immense progress in autopsy cohorts from even eight years ago and will only continue to improve, but availabel sample sizes still pose limitations on power of GWAS and similar 'omics approaches to studying disease risk using NPE.
+A strength of our study is that data on multiple NPE is available for each participant, allowing for multivariate approaches to studying NPE.
+However, as our PCA-based multivariate analyses did not identify any risk loci not identified in single-outcome analyses, this approach cannot overcome sample size limitations.
+
+In conclusion, we have identified several intriguing novel loci associated with NPE and validated multiple known risk loci for AD and related dementias in NPE.
+We also provide additional context and consideration for relationships between specific risk loci and different NPE.
+Overall, our study provides additional potential avenues of investigation into the relationship between genomics, neuropathology, and dementia.
+
+\newpage
+
+# References
+
+::: {#refs}
+:::
+
+\newpage
+
 # Tables
@@
 ```
+
+```{r}
+t2_env$t2
+```

### 2022-08-16 code/SAIGE/01_make_pheno_file.R

@@ -1,16 +1,0 @@
-#===================================================
-# merge .pheno and .covar files for SAIGE analysis
-#===================================================
-
-library(data.table)
-
-pheno <- fread("data/mega/mega_np.pheno", na.strings = "-1")
-covar <- fread("data/mega/mega_np.covar")
-
-pheno <- merge(pheno, covar)
-
-fwrite(pheno, 
-       file = "data/mega/mega_np_pheno_covar.txt", 
-       sep = " ",
-       na = NA,
-       quote = FALSE)

### 2022-08-16 code/SAIGE/bind_results.R

@@ -1,0 +1,46 @@
+
+#====================================================
+# Get SAIGE results files by chromosom and rbindlist
+#====================================================
+
+library(data.table)
+library(argparse)
+library(stringi)
+
+parser <- ArgumentParser()
+parser$add_argument("-p", "--prefix", 
+                    help = paste0("prefix for results to rbind. ",
+                                  "Files should end in '_chr[0-9]*.txt'. ",
+                                  "prefix should end in '_chr'."))
+parser$add_argument("-o", "--output", 
+                    help = paste0(
+                      "path for output file. If not used, will ",
+                      "default to --prefix minus '_chr' plus '.txt'"))
+args <- parser$parse_args()
+print(args)
+
+get_results <- function(files) {
+  nfiles <- length(files)
+  file_list <- vector(mode = "list", length = nfiles)
+  for (i in 1:nfiles) {
+    file_list[[i]] <- fread(files[i])
+  }
+  
+  results <- rbindlist(file_list)
+  return(results)
+}
+
+files <- list.files(dirname(args$prefix), 
+                    paste0(basename(args$prefix), "[0-9]*.txt$"))
+print(files)
+
+results <- get_results(paste0(dirname(args$prefix), "/", files))
+setorder(results, CHR, POS)
+
+if (is.null(args$output)) {
+  out <- stri_replace_last_fixed(args$prefix, "_chr", ".txt")
+} else {
+  out <- args$output
+}
+
+fwrite(results, file = out, sep = " ", quote = FALSE, na = "NA")

### 2022-08-16 code/SAIGE/make_pheno_file.R

@@ -1,0 +1,24 @@
+#===================================================
+# merge .pheno and .covar files for SAIGE analysis
+#===================================================
+
+library(data.table)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("--pheno", "-p", help = "phenotype file")
+parser$add_argument("--covar", "-c", help = "covariate file")
+parser$add_argument("--out", "-o", help = "output file")
+parser$add_argument("--na_string", default = "-1", help = "NA value string")
+args <- parser$parse_args()
+
+pheno <- fread(args$pheno, na.strings = args$na_string)
+covar <- fread(args$covar)
+
+pheno <- merge(pheno, covar)
+
+fwrite(pheno, 
+       file = args$out, 
+       sep = " ",
+       na = NA,
+       quote = FALSE)

### 2022-08-16 code/functions/find_closest_gene.R

@@ -3,21 +3,28 @@
 # annote a variant to the nearest gene
 #======================================================
 
-find_closest_gene <- function(chr, bp, g) {
+find_closest_gene <- function(chr, bp, g, 
+                              chrom_col="chr", 
+                              pos_col="value",
+                              gene_col="gene_name") {
   # chr = chromosome of variant
   # bp = base pair of variant
   # g = data.table with following columns
-  #   V2: Chromosome of gene
-  #   value: Start or end site of gene (best to use data table with
+  #   chrom_col: Chromosome of gene
+  #   pos_col: Start or end site of gene (best to use data table with
   #     two rows per gene -- 1 with start and one with end position)
-  #   V6: Gene name
+  #   gene_col: Gene name
   require(data.table)
   g <- as.data.table(g)
-  chr <- as.character(chr)
+  snp_chr <- as.character(chr)
   closest_gene <-
-    g[V2 == chr
-    ][
-      abs(value - bp) == min(abs(value - bp)), V6]
+    g[as.character(get(chrom_col)) == snp_chr][
+      abs(get(pos_col) - bp) == min(abs(get(pos_col) - bp)), get(gene_col)]
+  
+  if (length(closest_gene) > 1) {
+    closest_gene <- paste(closest_gene, collapse = "/")
+  }
+  
   return(closest_gene)
 }
 

### 2022-08-16 code/functions/get_results.R

@@ -1,0 +1,12 @@
+
+get_results <- function(files) {
+  require(data.table)
+  nfiles <- length(files)
+  file_list <- vector(mode = "list", length = nfiles)
+  for (i in 1:nfiles) {
+    file_list[[i]] <- fread(files[i])
+  }
+  
+  results <- rbindlist(file_list)
+  return(results)
+}

### 2022-08-16 code/gwas/ordinal/polmm_act.sh

@@ -1,0 +1,41 @@
+
+mkdir output/gwas/act/polmm
+
+phenos=$(cat data/mega/act_np_ord.pheno | \
+         cut -d " " -f 3- | \
+         awk 'NR == 1{print}')
+
+for pheno in $phenos
+  do
+  # output top clumped SNPs from NACC analysis to SnpMatrix
+  snps=tmp/"$pheno"_nacc_snps.txt
+  cat output/gwas/adc/polmm/"$pheno"_polmm_results.clumped | \
+    awk 'NR>1{print $3}' \
+    > $snps
+
+  snpmat_prefix=data/mega/SnpMatrix/nacc_"$pheno"
+  ./code/gwas/ordinal/polmm_create_SnpMatrix.R \
+    --snp_list $snps \
+    -o $snpmat_prefix
+
+  # create null model
+  null_model=output/gwas/act/polmm/"$pheno"_null_model.Rds
+  ./code/gwas/ordinal/polmm_null_model.R \
+    --covar data/mega/act_np.covar \
+    --pheno data/mega/act_np_ord.pheno \
+    --plink tmp/mega_np_prune \
+    --phenotype $pheno \
+    --out $null_model \
+    --ncores 100
+
+  # run regression
+  results=output/gwas/act/polmm/"$pheno"_polmm_results.Rds
+  ./code/gwas/ordinal/polmm_regression.R \
+    -m $null_model \
+    --snp_matrix "$snpmat_prefix"SnpMatrix.Rds \
+    -o $results
+
+  Rscript --vanilla ./code/gwas/rds_to_text.R \
+  -r $results \
+  -o output/gwas/act/polmm/"$pheno"_polmm_results.txt
+  done

### 2022-08-16 code/gwas/ordinal/polmm_adni.sh

@@ -1,0 +1,41 @@
+
+mkdir output/gwas/adni/polmm
+
+phenos=$(cat data/mega/adni_np_ord.pheno | \
+         cut -d " " -f 3- | \
+         awk 'NR == 1{print}')
+
+for pheno in $phenos
+  do
+  # # output top clumped SNPs from NACC analysis to SnpMatrix
+  # snps=tmp/"$pheno"_nacc_snps.txt
+  # cat output/gwas/adc/polmm/"$pheno"_polmm_results.clumped | \
+  #   awk 'NR>1{print $3}' \
+  #   > $snps
+  # 
+  snpmat_prefix=data/mega/SnpMatrix/nacc_"$pheno"
+  # ./code/gwas/ordinal/polmm_create_SnpMatrix.R \
+  #   --snp_list $snps \
+  #   -o $snpmat_prefix
+
+  # create null model
+  null_model=output/gwas/adni/polmm/"$pheno"_null_model.Rds
+  ./code/gwas/ordinal/polmm_null_model.R \
+    --covar data/mega/adni_np.covar \
+    --pheno data/mega/adni_np_ord.pheno \
+    --plink tmp/mega_np_prune \
+    --phenotype $pheno \
+    --out $null_model \
+    --ncores 100
+
+  # run regression
+  results=output/gwas/adni/polmm/"$pheno"_polmm_results.Rds
+  ./code/gwas/ordinal/polmm_regression.R \
+    -m $null_model \
+    --snp_matrix "$snpmat_prefix"SnpMatrix.Rds \
+    -o $results
+
+  Rscript --vanilla ./code/gwas/rds_to_text.R \
+  -r $results \
+  -o output/gwas/adni/polmm/"$pheno"_polmm_results.txt
+  done

### 2022-08-16 code/gwas/ordinal/polmm_create_SnpMatrix.R

@@ -14,51 +14,65 @@ parser$add_argument("-b", "--bim", help = "PLINK .bim file",
 parser$add_argument("-m", "--snp_set_size", 
                     help = "max number of SNPs per SnpMatrix",
                     default = 50000L)
+parser$add_argument("--snp_list", help = "text filename with list of SNP IDs")
 parser$add_argument("-o", "--out", default = "tmp/", 
                     help = "prefix for output SnpMatrix .Rds objects")
 parser$add_argument("-c", "--cores", default = 64, 
                     help = "number of cores for parallel computing")
 args <- parser$parse_args()
-
+print(args)
 if (is.null(args$bim) | is.null(args$snp_set_size)) {
   stop("BIM file and SNP set size are required arguments.")
 }
 
-if (!dir.exists(args$out)) {
-  dir.create(args$out)
+if (!dir.exists(dirname(args$out))) {
+  dir.create(dirname(args$out), recursive = TRUE)
 }
 
 plink <- stri_replace_last_fixed(args$bim, ".bim", "")
-m <- as.integer(args$snp_set_size)
-cores <- as.integer(args$cores)
 
-# get total number of SNPs in .bim file
-n_snp <- as.integer(
-  stri_split_fixed(
-    system(paste0("wc -l ", plink, ".bim"), intern = TRUE), 
-    pattern=" ")[[1]][1]
-)
-
-n_groups <- (n_snp %/% m)
-groups <- c(seq(0, m * n_groups, m), n_snp)
-
-# use multicore, set to the number of our cores
-registerDoParallel(cores = cores)  
-out_list <- 
-  foreach (i=2:length(groups), .combine = c) %dopar% {
-    snp_start <- groups[i - 1] + 1
-    snp_stop <- groups[i]
-    mat <- read.plink(bed = paste0(plink, '.bed'), 
-                      bim = paste0(plink, '.bim'), 
-                      fam = paste0(plink, '.fam'),
-                      select.snps = snp_start:snp_stop)
-    out_file <- paste0(args$out, "SnpMatrix_", i - 1,".Rds")
-    saveRDS(mat, file = out_file)
-    rm(mat)
-    out_file
-  }
-stopImplicitCluster()
-
-saveRDS(out_list, file = paste0(args$out, "SnpMatrix_list.Rds"))
+if (is.null(args$snp_list)) {
+  m <- as.integer(args$snp_set_size)
[…]

### 2022-08-16 code/gwas/ordinal/subset_snplist_chr19.R

@@ -1,0 +1,7 @@
+#===========================================================================
+# select only Chromosome 19 SNP Matrices (95-97) for APOE-adjusted analysis
+#===========================================================================
+
+snpmat_list <- readRDS("data/mega/SnpMatrix/SnpMatrix_list.Rds")
+snpmat_list <- snpmat_list[95:97]
+saveRDS(snpmat_list, file = "data/mega/SnpMatrix/SnpMatrix_list95-97.Rds")

### 2022-08-16 code/gwas/rds_to_text.R

@@ -1,0 +1,14 @@
+#===================================
+# write text files of POLMM results
+#===================================
+
+library(data.table)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("-r", "--rds", help = "results .Rds file")
+parser$add_argument("-o", "--out", help = "output file name")
+args <- parser$parse_args()
+
+results <- readRDS(args$rds)
+fwrite(results, file = args$out, sep = " ", quote = FALSE)

### 2022-08-16 code/magma/01_snp_gene_annot.sh

@@ -10,5 +10,5 @@ cat raw_data/NCBI38.gene.loc | \
 magma \
   --snp-loc data/mega/mega_np.bim \
   --gene-loc tmp/hg38.gene.loc \
-  --annotate window=100 \
+  --annotate window=10 \
   --out data/mega/snp_gene_annot_hg38

### 2022-08-16 code/manuscript/02_table2_mega_analysis_results.R

@@ -1,0 +1,199 @@
+#==========================================
+# Produce Table 2: GWAS results summary
+#==========================================
+
+library(data.table)
+library(magrittr)
+library(rtracklayer)
+library(flextable)
+source("code/functions/find_closest_gene.R")
+source("code/functions/make_or_95_ci.R")
+
+bim <- fread("data/mega/mega_np.bim")
+setnames(bim, 
+         c("V1", "V2", "V4", "V5", "V6"), 
+         c("CHR", "SNP", "BP", "A1", "A2"))
+pheno_ord <- fread("data/mega/mega_np_ord.pheno")
+pheno_bin <- fread("data/mega/mega_np.pheno")
+
+results_files_ord <- paste0("output/gwas/mega/polmm/", 
+                            colnames(pheno_ord)[3:ncol(pheno_ord)],
+                            "_polmm_results.txt")
+results_files_bin <- paste0("output/gwas/mega/saige/",
+                            colnames(pheno_bin)[3:ncol(pheno_bin)],
+                            "_saige_results.txt")
+clump_files_ord <- paste0("output/gwas/mega/polmm/", 
+                          colnames(pheno_ord)[3:ncol(pheno_ord)],
+                          "_polmm_results.clumped")
+clump_files_bin <- paste0("output/gwas/mega/saige/",
+                          colnames(pheno_bin)[3:ncol(pheno_bin)],
+                          "_saige_results.clumped")
+
+get_results_list <- function(files) {
+  nfiles <- length(files)
+  file_list <- vector(mode = "list", length = nfiles)
+  for (i in 1:nfiles) {
+    file_list[[i]] <- fread(files[i])
+  }
+  return(file_list)
+}
+
+# collate binary and ordinal outcome results
+results_ord <- get_results_list(results_files_ord)
+for (i in 1:(ncol(pheno_ord) - 2)) {
+  results_ord[[i]][, Phenotype := colnames(pheno_ord)[i + 2]]
+}
+results_ord <- rbindlist(results_ord)
+
+results_bin <- get_results_list(results_files_bin)
+for (i in 1:(ncol(pheno_bin) - 2)) {
+  results_bin[[i]][, Phenotype := colnames(pheno_bin)[i + 2]]
+}
+results_bin <- rbindlist(results_bin)
+
+# add significant locus from APOE-adjusted CAA analysis
+caa_apoe_results <- 
+  fread("output/gwas/mega/polmm/caa_ord_apoe_polmm_results.txt")
+
+# collate clumped results
[…]

### 2022-08-16 code/manuscript/locuszoom_atheroscler.sh

@@ -1,0 +1,22 @@
+cat output/gwas/mega/polmm/atheroscler_polmm_results.txt | \
+  ~/software/locuszoom/bin/locuszoom \
+    --metal - \
+    --markercol SNPID \
+    --pvalcol pval.spa \
+    --chr 13 \
+    --refsnp rs2000660 \
+    --flank 200000 \
+    --source 1000G_Nov2014 \
+    --build hg38 \
+    --pop EUR \
+    --delim space \
+    --no-date \
+    ymax=10 \
+    legend="none" \
+    geneFontSize=1.18 \
+    axisTextSize=1.4 \
+    refsnpTextSize=1.2 \
+    showRecomb=FALSE \
+    rfrows=2	\
+    --prefix doc/ \
+    --rundir ~/software/locuszoom

### 2022-08-16 code/multivar/pheno_cors.R

@@ -1,64 +1,0 @@
-#==============================
-# get list of phenotypes with
-# sample sizes >5000
-# for genetic correlation
-#==============================
-
-library(data.table)
-library(magrittr)
-library(polycor)
-
-bin_pheno <- fread("data/mega/mega_np.pheno", na.strings = "-1")
-ord_pheno <- fread("data/mega/mega_np_ord.pheno", na.strings = "-1")
-ord_pheno[, `:=`(vcid  = NULL, part_def = NULL)]
-pheno <- merge(ord_pheno, bin_pheno[, .(FID, IID, hs, grossinf, microinf)],
-               by = c('FID', 'IID'))
-# pheno[, hs := NULL]
-
-for (i in colnames(pheno[, 3:ncol(pheno)])) {
-  if (is.numeric(pheno[[i]])) {
-    if (pheno[!(is.na(get(i))), 
-                  sum(get(..i))] %in% c(0, .N)) {
-      pheno[, (i) := NULL]
-    }
-  }
-}
-
-
-pheno_4k <- character()
-for (i in colnames(pheno)) {
-  if (pheno[!(is.na(get(i))), .N] >= 2000) {
-    pheno_4k <- c(pheno_4k, i)
-  }
-}
-pheno_4k <- pheno[, ..pheno_4k]
-pheno_4k <- pheno_4k[complete.cases(pheno_4k)]
-
-n_cols <- ncol(pheno_4k) - 2L
-cor_mat <- matrix(nrow = n_cols, ncol = n_cols)
-colnames(cor_mat) <- 
-  rownames(cor_mat) <- 
-  colnames(pheno_4k[, -c("FID", "IID")])
-for (i in 1L:n_cols) {
-  pheno1 <- colnames(pheno_4k)[i + 2L]
-  for (j in i:n_cols) {
-    pheno2 <- colnames(pheno_4k)[j + 2L]
-    cor_mat[i, j] <- cor_mat[j, i] <- 
-      pheno_4k[, polychor(get(..pheno1), 
-                          get(..pheno2))]
-  }
-}
-
-cor_mat[is.na(cor_mat)] <- 0
-cor_mat <- round(cor_mat, 2)
-
-pca1 <- prcomp(pheno_4k[, -c("FID", "IID")], scale. = T, center = T)
-pca2 <- princomp(cor_mat)
-pcs <- data.table(FID = pheno_4k$FID, 
[…]

### 2022-08-16 code/multivar/saige_analysis.sh

@@ -1,0 +1,66 @@
+cluster=vcid
+pheno=PC1
+
+covars=$(cat ./data/mega/mega_np.covar | \
+         cut -d " " -f 3- | \
+         awk 'BEGIN {FS=" "; OFS = ",";} NR==1 {$1=$1; print $0}')
+
+cat_covars=$(cat ./data/mega/mega_np.covar | \
+         cut -d " " -f 3,5-21 | \
+         awk 'BEGIN {FS=" "; OFS = ",";} NR==1 {$1=$1; print $0}')
+
+null_model_stage1=./output/multivar/"$cluster"_"$pheno"_stage1
+Rscript --vanilla ./code/SAIGE/step1_fitNULLGLMM.R \
+    --plinkFile=./tmp/mega_np_prune \
+    --phenoFile=./data/mega/multivar/pcs_"$cluster".txt \
+    --phenoCol="$pheno" \
+    --covarColList="$covars" \
+    --qCovarColList="$cat_covars" \
+    --sampleIDColinphenoFile=IID \
+    --nThreads=100 \
+    --outputPrefix="$null_model_stage1" \
+    --IsOverwriteVarianceRatioFile=TRUE \
+    --traitType=quantitative
+
+res_colname="$pheno"_res
+res_pheno_file=./tmp/"$cluster"_"$res_colname".pheno
+Rscript --vanilla ./code/SAIGE/saige_residuals.R \
+  --model "$null_model_stage1".rda \
+  --covar data/mega/mega_np.covar \
+  --colname $res_colname \
+  --sampleID_col IID \
+  --output $res_pheno_file
+  
+null_model_stage2=./output/multivar/"$cluster"_"$pheno"_stage2
+Rscript --vanilla ./code/SAIGE/step1_fitNULLGLMM.R \
+    --plinkFile=./tmp/mega_np_prune \
+    --phenoFile="$res_pheno_file" \
+    --phenoCol="$res_colname" \
+    --covarColList="$covars" \
+    --qCovarColList="$cat_covars" \
+    --sampleIDColinphenoFile=IID \
+    --nThreads=100 \
+    --outputPrefix="$null_model_stage2" \
+    --IsOverwriteVarianceRatioFile=TRUE \
+    --traitType=quantitative \
+    --invNormalize=TRUE
+
+saige_results=./output/multivar/"$cluster"_"$pheno"_saige_results_chr
+for chr in $(seq 1 22)
+    do
[…]

### 2022-08-16 code/np_coloc/loci.R

@@ -1,31 +1,60 @@
 library(data.table)
 library(magrittr)
-source('code/functions/plink2_to_plink_format.R')
 `%!in%` <- function(x, v) {!(x %in% v)}
 
-pheno <- fread("data/mega/mega_np.pheno", na.strings = '-1')
-phenotypes <- colnames(pheno)[3:length(colnames(pheno))]
-phenotypes <- phenotypes[phenotypes %!in% c("lewy")]
-file_path <- function(phenotype, prefix="output/gwas/mega") {
-  paste0(prefix, '/mega.', phenotype, '.glm.logistic')
+pheno_ord <- fread("data/mega/mega_np_ord.pheno", na.strings = '-1')
+pheno_bin <- fread("data/mega/mega_np.pheno")
+phenotypes_ord <- colnames(pheno_ord)[3:ncol(pheno_ord)]
+phenotypes_bin <- colnames(pheno_bin)[3:ncol(pheno_bin)]
+phenotypes <- c(phenotypes_ord, phenotypes_bin)
+
+results_files_ord <- paste0("output/gwas/mega/polmm/", 
+                            colnames(pheno_ord)[3:ncol(pheno_ord)],
+                            "_polmm_results.txt")
+results_files_bin <- paste0("output/gwas/mega/saige/",
+                            colnames(pheno_bin)[3:ncol(pheno_bin)],
+                            "_saige_results.txt")
+clump_files_ord <- paste0("output/gwas/mega/polmm/", 
+                          colnames(pheno_ord)[3:ncol(pheno_ord)],
+                          "_polmm_results.clumped")
+clump_files_bin <- paste0("output/gwas/mega/saige/",
+                          colnames(pheno_bin)[3:ncol(pheno_bin)],
+                          "_saige_results.clumped")
+
+get_results_list <- function(files) {
+  nfiles <- length(files)
+  file_list <- vector(mode = "list", length = nfiles)
+  for (i in 1:nfiles) {
+    file_list[[i]] <- fread(files[i])
+  }
+  return(file_list)
 }
 
-format_results <- function(file_name) {
-  results <- fread(file_name)
-  results <-plink2_to_plink_format(results) %>% 
-  .[!(is.na(P))]
+# 1. collate binary and ordinal outcome results
+results_ord <- get_results_list(results_files_ord)
[…]

### 2022-08-16 code/prep/mega/02_make_pheno.R

@@ -211,4 +211,68 @@ fwrite(mega_np[, .(FID, IID, braak, cerad, diffuse_abeta, arteriol,
        na = -1)
 saveRDS(mega_np, file = "data/mega/mega_np.Rds")
 
+# cohort-stratified pheno files
+## nacc/adc
+fwrite(mega_np[FID == "0", ..plink_pheno_vars], 
+       file = "data/mega/nacc_np.pheno", 
+       quote = FALSE, 
+       sep = ' ', 
+       na = -1)
+
+fwrite(mega_np[FID == "0", 
+               .(FID, IID, braak, cerad, diffuse_abeta, arteriol, 
+                 atheroscler, late, lewy_body, caa_ord, part_def,
+                 vcid)],
+       file = "data/mega/nacc_np_ord.pheno",
+       quote = FALSE,
+       sep = ' ',
+       na = -1)
+
+## rosmap
+fwrite(mega_np[FID == "1", ..plink_pheno_vars], 
+       file = "data/mega/rosmap_np.pheno", 
+       quote = FALSE, 
+       sep = ' ', 
+       na = -1)
+
+fwrite(mega_np[FID == "1", 
+               .(FID, IID, braak, cerad, diffuse_abeta, arteriol, 
+                 atheroscler, late, lewy_body, caa_ord, part_def,
+                 vcid)],
+       file = "data/mega/rosmap_np_ord.pheno",
+       quote = FALSE,
+       sep = ' ',
+       na = -1)
+
+## act (no LATE data available)
+fwrite(mega_np[FID == "2", ..plink_pheno_vars], 
+       file = "data/mega/act_np.pheno", 
+       quote = FALSE, 
+       sep = ' ', 
+       na = -1)
+
+fwrite(mega_np[FID == "2", 
+               .(FID, IID, braak, cerad, diffuse_abeta, arteriol, 
+                 atheroscler, lewy_body, caa_ord, part_def,
+                 vcid)],
+       file = "data/mega/act_np_ord.pheno",
+       quote = FALSE,
+       sep = ' ',
+       na = -1)
+
+## adni
+fwrite(mega_np[FID == "ADNI", ..plink_pheno_vars], 
+       file = "data/mega/adni_np.pheno", 
+       quote = FALSE, 
+       sep = ' ', 
+       na = -1)
+
+fwrite(mega_np[FID == "ADNI", 
+               .(FID, IID, braak, cerad, diffuse_abeta, arteriol, 
[…]

### 2023-01-05 manuscript.Rmd
update manuscript for initial submission

--- 2022-07-27 manuscript.Rmd
@@
 bibliography: doc/references.bib
-csl: doc/nature.csl
+csl: doc/springer_basic.csl
 link-citations: true
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
 ```
 
-# Introduction
+```{r}
+library(data.table)
+library(flextable)
+```
 
-Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinical AD; however, clinical AD does not always correspond with AD neuropathology and vice-versa.[@nelson2012] We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:
+Proposed journal: ***Molecular Neurodegeneration***
 
-1.  identify novel genetic risk loci for individual NPE
-2.  identify common sources of genetic risk (pleiotropy) between NPE
-3.  provide evidence that the effects of clinical AD genetic risk loci may be mediated by specific NPE.
+**Title**
+
+Genome-wide association study of multiple neuropathology endophenotypes identifies novel risk loci and provides insights into known Alzheimer's risk loci
+
+**Authors**
+
+Lincoln M.P. Shade^1^
+
+Yuriko Katsumata^1,2^
+
+Steven A. Claas^1^
+
+Mark T. W. Ebbert^2,3,4^
+
+Erin L. Abner^2,5^
+
+Timothy J. Hohman^6^
+
+Shubhabrata Mukherjee^7^
+
+Kwagsik Nho^8^
+
+Andrew J. Saykin^8^
+
+David A. Bennett^9^
+
+Julie A. Schneider^10^
+
+Peter T. Nelson^2,11^
+
+David W. Fardo^1,2^
+
+for the Alzheimer's Disease Neuroimaging Initiative^\*^
+
+**Author Affiliations**
+
+^1^Department of Biostatistics, College of Public Health, University of Kentucky, Lexington, KY
+
+^2^Sanders-Brown Center on Aging and Alzheimer's Disease Research Center, University of Kentucky, Lexington, KY
+
+^3^Division of Biomedical Informatics, Department of Internal Medicine, University of Kentucky, Lexington, KY
+
+^4^Department of Neuroscience, University of Kentucky College of Medicine, Lexington, KY
+
+^5^Department of Epidemiology and Environmental Health, College of Public Health, University of Kentucky, Lexington, KY
+
+^6^Vanderbilt Memory & Alzheimer's Center, Department of Neurology, Vanderbilt University Medical Center, Nashville, Tennessee
+
+^7^Department of Medicine, University of Washington, Seattle, WA
+
+^8^Department of Radiology & Imaging Sciences, Indiana University School of Medicine, Indianapolis, IN
+
+^9^Rush Alzheimer's Disease Center and Department of Neurological Sciences, Rush Medical College, Chicago, IL
+
+^10^Rush Alzheimer's Disease Center and Departments of Neurology and Pathology, Rush University Medical Center, Chicago, IL
+
+^11^Department of Pathology and Laboratory Medicine, University of Kentucky, Lexington, KY
+
+^\*^Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
+A complete listing of ADNI investigators can be found at: <http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf>
+
+**Corresponding Author**
+
+David W. Fardo
+
+Healthy Kentucky Research Building
+
+760 Press Ave, Room 372
+
+Lexington, KY 40536-0679
+
+[david.fardo\@uky.edu](mailto:david.fardo@uky.edu)
+
+859-218-2070
+
+**Abstract**
+
+***Background***
+
+Alzheimer's disease is highly heritable and exhibits neuropathological hallmarks of neurofibrillary tau tangles and neuritic amyloid plaques.
+However, the reality of the relationship between clinically diagnosed Alzheimer's disease and neuropathology is more complex.
+Upon autopsy, many Alzheimer's disease patients have multiple comorbid neuropathologies other than plaques and tangles which may have independent or pleiotropic genomic risk factors.
+Autopsy data combined with genome-wide association study (GWAS) provides the opportunity to study the genetic risk factors of multiple neuropathologies that contribute to dementia.
+
+***Methods***
+
+We studied the genome-wide risk factors of eleven Alzheimer's disease-related neuropathology endophenotypes.
+We used four sources of neuropathological data: National Alzheimer's Coordinating Center, Religious Orders Study and Rush Memory and Aging Project (ROSMAP), Adult Changes in Thought study, and Alzheimer's Disease Neuroimaging Initiative.
+We used generalized linear mixed models to identify risk loci, followed by Bayesian colocalization analyses to identify potential molecular functions of risk loci.
+We then analyzed associations between four CpG methylation sites, identified through colocalization analysis, with cerebral amyloid angiopathy (CAA) pathology in ROSMAP.
+
+***Results***
+
+After adjusting for *APOE* genotype, we identified a locus near *APOC2* (lead variant rs4803778) associated with CAA that colocalizes with DNA methylation at four nearby CpG sites in the cerebral cortex.
+Two of these sites, cg0955818 ($p = 0.004$) and cg13119609 ($p = 0.0007$), were themselves significantly associated with CAA in ROSMAP.
+We identified two other novel neuropathology risk loci: one *PIK3R5* locus (lead variant rs72807981) with neurofibrillary pathology, and one *COL4A1* locus (lead variant rs2000660) with cerebral atherosclerosis.
+We also confirmed associations between known Alzheimer's genes and multiple neuropathology endophenotypes, including *APOE* (neurofibrillary tangles, neuritic plaques, diffuse plaques, cerebral amyloid angiopathy, and TDP-43 pathology); *BIN1* (neurofibrillary tangles and neuritic plaques); and *TMEM106B* (TDP-43 pathology and hippocampal sclerosis).
+
+***Conclusions***
+
+Associations between the Chromosome 19 CAA risk locus, CpG methylation, and CAA pathology are consistent with the effect of this locus on disease being mediated by DNA methylation.
+Furthermore, the same two CpG sites are associated with dementia in an independent cohort, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk.
+The atherosclerosis risk variant rs2000660 is in strong linkage disequilibrium with a synonymous coding variant (rs650724) of *COL4A1*, providing a candidate functional variant.
+*BIN1* is associated with neurofibrillary tangles and neuritic plaques but not with diffuse amyloid pathology.
+*TMEM106B* is associated with hippocampal sclerosis and TDP-43 pathology but not the canonical Alzheimer's disease pathologies.
+Our findings provide novel insights into neuropathology risk factors and parse clinical Alzheimer's genetic risk into specific key neuropathological domains.
+
+# Key words
+
+GWAS, Alzheimer's, neuropathology, atherosclerosis, neurofibrillary tangles, cerebral amyloid angiopathy, TDP-43
+
+# Background
+
+Aged individuals often accumulate multiple brain pathologies that contribute individually and synergistically to cognitive decline and dementia.
+Alzheimer's disease (AD) is the most commonly diagnosed form of dementia and poses an enormous burden on human health and well-being.
+In the United States alone, over six million individuals are living with AD, and the disease imposes financial costs of treatment and care of over \$300 billion annually [@2021alz2021].
+
+The most common form of AD, late-onset AD (LOAD), presents in individuals aged 65 years and older and is highly heritable, with twin and family studies estimating its heritability at \~60% [@ridge2013; @gatz1997; @gatz2006].
+Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, immune response, and extracellular matrix regulation [@jansen2019; @wightman2021; @bellenguez2022].
+Most dementia GWAS to date have focused on clinically diagnosed LOAD or proxy outcomes based on family history of dementia [@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019]; however, late-onset dementia diagnosed as LOAD is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors [@farfel2016; @katsumata2022; @karanth2020].
+
+While neuritic plaques and neurofibrillary tangles (NFT) are present in the brains of the majority of people diagnosed with clinical AD, many people either do not have substantial AD neuropathology or have both AD and comorbid non-AD pathologies [@nelson2009; @karanth2020].
+For example, TDP-43 pathology contributes to an amnestic syndrome, limbic-predominant age-related TDP-43 encephalopathy (LATE), that presents similarly to LOAD [@nelson2019].
+However, LATE and LOAD exhibit differing patterns of pathogenesis in aged autopsy cohorts.
+LOAD pathology peaks in prevalence in individuals who die at age 95 years and decreases thereafter [@farfel2019].
+In contrast, prevalence of LATE neuropathologic change (LATE-NC) increases monotonically with age of death, with TDP-43 pathology often first appearing in the limbic region or entorhinal cortex and progressing to the neocortex [@farfel2019; @nelson2019].
+Hippocampal sclerosis (HS) of aging, a pathology characterized by uni- or bilateral neuronal death, gliosis, and atrophy of the hippocampus beyond normal ranges based on levels of AD pathology, commonly co-occurs with LATE-NC and is associated with severe cognitive impairment [@brenowitz2014; @nelson2019].
+
+In addition to proteinopathies, non-stroke vascular pathologies are common in elderly autopsied research participants and contribute to cognitive decline and dementia [@skrobot2016].
+Cerebral amyloid angiopathy (CAA) is one such vascular pathology characterized by amyloid-beta deposition in cerebral blood vessels [@weber2018].
+CAA often co-occurs with LOAD, but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma [@weber2018; @skrobot2016].
+Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline [@smith2012; @skrobot2016].
+Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and hippocampal sclerosis [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies [@arvanitakis2016; @ighodaro2017].
+Collectively, these factors reveal an increasingly complex and synergistic web of pathologies contributing to cognitive impairment and dementia, and demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.
+
+The investigation of genetic risk of neuropathologies is an attractive approach for uncovering biological pathways involved in the development of LOAD and related dementias.
+Neuropathologies are commonly operationalized in aged autopsy cohorts as binary or semi-quantitative (or ordinal) neuropathology endophenotypes (NPE).
+Prior GWAS of neuropathology endophenotypes have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022].
+However, continued recruitment in ongoing autopsy cohorts and methodological development for GWAS and downstream functional analysis provides additional opportunity for in-depth investigation into multiple neuropathologies studied across multiple cohorts.
+
+In this study, we perform GWAS on eleven neuropathology endophenotypes collected in four high-quality studies with autopsy and genotype data available.
+Endophenotypes studied include AD-associated pathologies; LATE-NC; hippocampal sclerosis; vascular NPE including cerebral amyloid angiopathy, gross infarcts, microinfarcts, atherosclerosis, and arteriolosclerosis; and Lewy body pathology.
+We also perform downstream functional analyses to explore potential biological functional mechanisms of identified risk loci and provide insight into the shared genomic risk of neuropathologies.
 
@@
 
-We used genotype and NPE data from four aged autopsy cohorts: National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Memory and Aging Project (ROSMAP), Adult Changes in Thought (ACT), and the AD Neuroimaging Initiative (ADNI).
+## Participants
 
-## Quality Control
+```{r}
+np <- readRDS("data/mega/mega_np.Rds")
+nadrc <- np[, length(table(NACCADC))]
+```
 
-Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS,[@shade2022] with the exception that all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021]
+We used genotype and neuropathology data from four high-quality data sources: the National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP), the Adult Changes in Thought (ACT) study, and the AD Neuroimaging Initiative (ADNI).
+Some participants in ROSMAP and ADNI also had neuropathology and genotype data available in NACC.
+In these cases, records in the NACC were preferentially kept in order to maximize sample size of the initial analysis using only NACC participants (see Statistical Analyses subsection).
 
-1.  Retrieve participants labeled as non-Hispanic white or similar (European, etc.) if available.
-2.  Perform imputation on TOPMed Imputation Server (ROSMAP and ADNI had already been imputed using same reference panel by collaborators, so this step was skipped for those cohorts).
-3.  Covert VCF files to PLINK file set, keeping variants will minor allele frequencies (MAF) of 0.1% or greater.
-4.  Merge sub-cohorts of same study (e.g. NACC cohorts 1-12 and ACT 1-3).
-5.  Remove duplicate participants in each study.
-6.  Remove participants without autopsy data available.
-7.  Iteratively remove variants and participants with excess missingness until no variant is missing in more than 5% of participants and no participant is missing more than 5% of variants.
-8.  Remove participants with unusually high/low heterozygosity (+/- 3 SD cutoff).
-9.  Merge with 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012] and remove participants without primarily European ancestry as determined by principal component analysis.
-10.
-Remove participants with rare neurological diseases that substantially raise the likelihood that present NPE could be due to unusual pathological processes (NACC only, as the other cohorts did not have this information available).
-11. Remove duplicate participants between studies, preferentially keeping the observation in the larger cohort (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
-12. Merge cohorts.
-13. Harmonize NPE variables to analyze.
-14. Create list of related ($\ge$ 2nd-degree relatives) individuals using KING[@manichaikul2010] to remove for each NPE variable, randomly keeping one participant in each related cluster.
-15. Create genetic principal components using the PC-AiR method in the GENESIS package in R 4.1.[@conomos2015]
+All study participants were deceased and the resulting data de-identified for University of Kentucky investigators, and we exclusively used archival samples.
+Therefore, our study does not fall under the definition of "Human Subjects Research" according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- "involving the collection/study of data or specimens if publicly available, or/or recorded such that subjects cannot be identified."
 
-We note that there are substantial differences in the way that NPE are collected in different cohorts,(cite) and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes and minimizing computational load for analysis.
-Thus, some ordinal variables were collapsed to binary variables, or synthetic NPE may have been created using categorical variables from one cohort and continuous variables from another.
-All code used for harmonization will be made available upon publication, and a brief overview of the harmonization strategy for each NPE will be included in the Supplemental Materials.
+The present study used NACC data from `r nadrc` National Institute on Aging-funded Alzheimer's Disease Research Centers (ADRCs).
+Individual ADRCs use different recruitment strategies and perform autopsies on-site, but neuropathology data at each ADRC are collected using a standard form (<https://files.alz.washington.edu/documentation/np11-form.pdf>) and submitted to NACC where they are aggregated and anonymized.
+The NACC Neuropathology (NP) data set based on the first version of this form was originally implemented in 2001 [@besser2018], and this analysis uses data from then through the December 2021 freeze.
+Participants were excluded if they did not have autopsy data available or if they were noted in the NP data set to have at least one of 19 conditions that could potentially bias results.
+These conditions include brain tumors, severe head trauma, and fronto-temporal lobar degeneration.
+(See **Supplementary Table S1** for full list of variables used for exclusion criteria.)
+
+ROSMAP has been previously described and consists of harmonized data from two longitudinal cohort studies: The Religious Orders Study (ROS) and the Rush Memory and Aging Project (MAP) [@bennett2018].
+ROS and MAP were both approved by an Institutional Review Board of Rush University Medical Center.
+All participants signed an Anatomic Gift Act, as well as informed and repository consents.
+ROS began in 1994 and has recruited over 1500 Catholic priests, nuns, and brothers across the United States.
+MAP started in 1997 and has enrolled more than 2300 community members in the greater Chicago area of northeastern Illinois.
+The ROSMAP NP data used in this study were received from Rush University Medical Center in January 2020.
+
+The ACT study began in 1994 and recruited residents in the greater Seattle area aged 65 years and older without dementia at time of enrollment [@kukull2002].
+The study has expanded to include three cohorts with continued enrollment using the original enrollment criteria and has a current total of 4,960 participants across all three cohorts.
+The ACT NP data used in this study were obtained from Kaiser Permanente in May 2021.
+
+The ADNI (<https://adni.loni.usc.edu>) was launched in 2003 as a public-private research partnership, led by Principal Investigator Michael W.
+Weiner, MD.
+The primary goal of ADNI has been to test whether serial magnetic resonance imaging, positron emission tomography (PET), other biological markers, and clinical and neuropsychological assessment can be combined to measure the progression of mild cognitive impairment and early AD.
+A subset of ADNI participants undergo autopsy and receive neuropathological phenotyping.
+The ADNI NP data used in this study was downloaded from the ADNI website in October 2021.
+
+## Genotype data and quality control
+
+Genotype data for all cohorts underwent imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel [@taliun2021].
+NACC and ACT raw genotype data were obtained from the September 2020 freeze of the Alzheimer's Disease Genetics Consortium (ADGC) and subsequently imputed, while pre-imputed ROSMAP and ADNI genotype data were received from collaborators in the Hohman lab at Vanderbilt University in March 2021 and December 2021, respectively.
+
+Quality control on genotype data was performed separately for each data source.
+Quality control and inclusion/exclusion criteria closely followed that used in our previous brain arteriolosclerosis GWAS [@shade2022].
+Briefly, we included only participants labeled as non-Hispanic white or similar (European, etc.) if available.
+After imputation, we first identified and subsequently removed duplicate samples using the KING software "\--duplicate" option [@manichaikul2010].
+We then removed participants without autopsy data available by merging genotype sample identification data with neuropathology data sets and retaining only samples present in both data sets.
+We then iteratively removed genetic variants and participants until no variants were missing in more than 5% of participants and no participants were missing more than 5% of variants (however, average genotype coverage was 99.7%, and no variants or participants were actually removed during this process).
+We then excluded participants with unusually high or low ($\pm$ 3 standard deviations from mean) genetic heterogeneity, as measured by the PLINK 1.9 "\--het" flag [@chang].
+Finally, we merged participants with the 1000 Genomes Phase 3 (1000 Genomes) cohorts [@the1000genomesprojectconsortium2012] and performed principal components analysis (PCA) on a subset of independent variants (measured by pairwise $r^2 < 0.2$).
+We excluded participants with substantial non-European ancestry, as determined by distance from the 1000 Genomes EUR superpopulation centroid using the first two principal components (PCs).
+
+## Defining and harmonizing neuropathology endophenotypes for analysis
+
+In total, we combined or harmonized 11 neuropathology endophenotypes for analysis across the four studies.
+We note that there are differences in the way that some neuropathological data were collected across studies, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
+Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another.
+A detailed listing of variables harmonized across cohorts to construct neuropathology endophenotypes for analysis is available in **Supplementary Table S3**.
+We briefly describe our harmonization methods below.
+
+Arteriolosclerosis, Braak NFT stage, CAA, atherosclerosis, Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, microinfarcts, and gross infarcts had variables in each cohort with directly comparable coding definitions and were straightforwardly renamed and combined with minimal recoding.
+Analyzed arteriolosclerosis, CAA, atherosclerosis, and CERAD score variables each had four stages with the following labels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe").
+Microinfarcts and gross infarcts were labeled either 0 ("absent") or 1 ("present").
+The Braak NFT Stage variable analyzed followed the staging criteria previously described in the literature and had seven levels, ranging from 0 (absent NFT) to six (diffuse NFT throughout cortex and large loss of neurons) [@braak1991].
+
+LATE-NC was recorded differently in several of the data sets and was harmonized to a four-level outcome variable following the simplified staging of TDP-43 pathology outlined in Figure 3B of the 2019 LATE working group report [@nelson2019].
+The following levels were used in the analyzed LATE-NC variable: 0, indicating lack of recorded TDP-43 proteinopathy; 1, indicating TDP-43 deposits in the amygdala only; 2, indicating deposits in the hippocampus or entorhinal cortex; and 3, indicating deposits in the neocortex.
+TDP-43 data was not available in ACT.
+In ROSMAP, TDP-43 pathology is recorded as a single variable following the same staging detailed above.
+In NACC and ADNI, the presence of TDP-43 pathology in each region of the brain is recorded as a separate binary indicator variable.
+To collapse TDP-43 pathology to a single ordinal variable in these data sets, we assigned a value based on the presence of the "highest" region where TDP-43 was present (*e.g.* a participant with TDP-43 pathology in both the hippocampus and the neocortex would be assigned a value of 3).
+Participants were labeled as 0 if they met two conditions: (1) they had recorded TDP-43 data available for at least one of the brain regions used for staging and (2) TDP-43 pathology was noted as absent in all the regions for which they had data available.
+
+Diffuse amyloid plaque pathology was recorded as a four-stage Thal phase of amyloid deposition in NACC, ACT, and ADNI with the following levels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe").
+In ROSMAP, diffuse plaques were examined and quantified in five regions (midfrontal cortex, entorhinal cortex, inferior parietal cortex, and hippocampus), then scaled by each region's standard deviation and averaged.
+To discretize this continuous variable in ROSMAP participants to a four-level variable, as recorded in the other data sets, we assigned participants a value of 0 ("none") if their averaged score was equal to 0, 1 ("mild") if their score was higher than 0 but $\le$ 0.5, 2 ("moderate") if their score was between 0.5 and one, and 3 ("severe") if their score was above 1.
+These labels roughly corresponded to score quartiles in ROSMAP.
+
+Hippocampal sclerosis is recorded as a binary indicator of the presence or absence of pathology in the NACC NP data set form version 1-9, ROSMAP, and ACT.
+In versions 10 and 11 of the NACC NP form and in ADNI, HS pathology is recorded as being absent, unilateral, or bilateral.
+To harmonize, we dichotomized HS pathology as being present if either unilateral or bilateral pathology was indicated.
+
+The Lewy body pathology variable we analyzed had four levels: 0, indicating absent Lewy body pathology in all regions examined or limited to the olfactory bulb; 1, indicating Lewy body pathology limited to the brainstem, including the substantia nigra; 2, indicating Lewy body pathology involving the limbic system or amygdala; and 3, indicating Lewy body involvement in the neocortex.
+In NACC and ROSMAP, Lewy body pathology was graded in ordinal variables with levels corresponding to the levels in the final harmonized outcome variable analyzed.
+In ADNI, a single variable was also used to record Lewy body pathology but included separate levels for (1) no pathology and olfactory-predominant and (2) limbic-predominant and amygdala-predominant pathology.
+These stages were collapsed into levels 0 and 2, respectively, for harmonization.
+In ACT, separate binary indicator variables were used to indicate presence of Lewy body pathology in each brain region checked.
+To harmonize, we created a new variable that coded Lewy body pathology stage according to the "highest" stage present in an individual (*e.g.* if pathology was present in both the amygdala and the neocortex, we assigned a value of 3).
 
@@
 
-All the analyses discussed below were done on the data set with the combined cohorts.
-
 ### Single-variant GWAS
 
-Single-variant logistic regression analyses were performed in PLINK 2.0.[@chang2015; @purcell] Covariates included age at death, sex, cohort, and the first 10 genetic principle components.
-Continuous covariates were standardized to a mean of 0 and a variance of 1.
+We analyzed ordinal endophenotypes using proportional odds logistic mixed-effects models implemented in the POLMM R package [@bi2021].
+We analyzed binary variables similarly with logistic mixed-effects models implemented in the SAIGE R package [@zhou2018].
+Covariates included age at death, sex, cohort, and the first 10 genetic PCs created using the PCA in Related Samples (PC-AiR) method in the GENESIS R package [@conomos2015].
+We included a dense genetic relationship matrix (GRM) as a random effect to account for relatedness between participants.
 An additive mode of inheritance was assumed in all analyses.
-Variants achieving a p-value of $< 5\times 10^{-8}$ were considered "genome-wide" significant.
-To identify independent genetic risk loci, variants were clumped using PLINK 1.9[@chang] using a linkage-disequilibrium threshold of $r^2 \le 0.05$.
 
-Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package.[@bi2021] Dense kinship matrices were used to account for relatedness.
+Analyses proceeded in two stages.
+In stage one, GRM were constructed using a pruned set of independent variants, defined as having a pairwise $r^2 < 0.2$ within moving windows of 15 kilobase pairs (kbp).
+Null models including fixed covariates and the GRM were then fitted using either the POLMM or SAIGE R packages.
+In stage 2, score tests were performed on each variant with a saddle-point approximation used to calculate p-values.
+We considered all variants with a p-value of $p < 5\times 10^{-8}$ to be genome-wide significant.
+To identify independent risk loci, we clumped results using the "\--clump" flag in PLINK 1.9 with the pairwise linkage-disequilibrium (LD) threshold set to $r^2 \le 0.05$ [@chang; @chang2015].
 
-### Conditional and mediation analyses
+We performed analyses both in individual cohorts separately and in a pooled data set with all participants.
+Single-variant analyses were first performed using NACC data.
+For all lead variants with a p-values $< 1 \times 10^{-6}$, we attempted to validate the association in ROSMAP, ACT, and ADNI separately.
+Participants were then pooled into a single data set and genome-wide association analyses were performed again using all participants.
+As noted in the section on harmonization above, LATE-NC data was not available for ACT participants, so validation of LATE-NC-associated variants was not attempted in ACT.
 
-For genetic loci associated with multiple NPE, we considered mediation hypotheses to test whether the association of the lead variant with one NPE was mediated by another NPE.
-Regression analyses were run in R using the `glm` or `polr` functions and mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.
+### Re-analysis of the *APOE* region
+
+The region surrounding the *APOE* gene on Chromosome 19 is consistently the single strongest genetic risk factor for LOAD in GWAS.
+Three common forms of the *APOE* gene -- $\epsilon 2$, $\epsilon 3$, and $\epsilon 4$ -- are present in our study populations, and the $\epsilon 2$ and $\epsilon 4$ alleles are associated with lower and higher risk of LOAD, respectively, relative to the $\epsilon 3$ allele.
+We therefore expected that variants in the *APOE* region, defined as the region within 200 kbp from the start and end transcription sites of *APOE*, would be associated with multiple NPE in our study.
+Moreover, we hypothesized that genetic variants in the *APOE* region may influence neuropathology risk independently of the effects of *APOE* $\epsilon$ alleles.
+
+To test the hypothesis that genetic loci in the *APOE* region may affect neuropathology risk independently from *APOE* $\epsilon$ alleles, we re-analyzed variants in Chromosome 19 while adjusting for *APOE* $\epsilon$ genotype.
+We limited re-analysis to endophenotypes with at least one significant association signal within the *APOE* locus in the pooled single-variant association analysis described in the previous section.
+*APOE* genotypes were determined using the rs7412 and rs429358 variants according to the SNPedia online reference [@apoe-s].
+The $\epsilon3/\epsilon3$ genotype was used as reference, and we included fixed-effect indicator variables to adjust for $\epsilon2/\epsilon2$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, and $\epsilon4/\epsilon4$ genotypes.
+We chose this approach rather than adjusting for counts of $\epsilon 2$ and $\epsilon 4$ alleles because it is robust to potential non-linear effects of genotypes.
+
+### Gene-based analyses
+
+Analyses that test for the aggregate effect of all variants in a gene region may increase power for detection of risk genes.
+We thus performed gene-based analyses for all neuropathology endophenotypes studied.
+Gene start and end sites were determined using GRCh38 gene regions.
+Variants were mapped to a gene if they were located within 10 kbp of the gene's start or end sites.
+Using an unrelated set of the pooled participant data set, gene-based analyses were performed using MAGMA v1.10 [@leeuw2015].
+The same fixed-effects covariates used in single-variant analyses were used, and the top variant PCs that accounted for 99.9% of variance in a gene's region were used to test for significance using an F-test.
+We considered genes with $p < 2.5 \times 10^{-6}$ to be significantly associated with endophenotypes.
+
+### Principal component outcome analyses
+
+Many neuropathology endophenotypes are highly correlated with one another, and this phenomenon is not reflected in analyses using individual neuropathology endophenotypes as outcomes.
+Thus, we sought to perform genetic association analyses that could better account for the common co-occurrence of NPE.
+We first assessed the phenotype correlation of neuropathology endophenotypes used in the pooled GWAS using polychoric correlation and grouped endophenotypes by visually examining a dendrogram based on phenotype-phenotype correlations.
+Using the psych R package [@revelle2022], for each group of endophenotypes, we then performed PCA on the matrix of neuropathology endophenotype outcomes in that group.
+The first PC (PC1) of each group was then used as the outcome variable for GWAS using SAIGE in a two-stage analysis.
+In the first stage, PC1 was regressed on fixed-effects covariates, with the GRM included as a random effect.
+Because the residuals of PC1 may not be normally distributed, the residuals of stage one were then rank inverse normalized and regressed on covariates and GRM again in stage two.
+Score tests were then performed on genetic variants and a saddle-point approximation used to estimate p-values.
 
@@
 
-All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues.
-Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs.
-A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) \> 80% was used as a threshold for evidence of colocalization.
+Colocalization analysis is a Bayesian analytical method that seeks to answer whether a single genetic locus drives genotypic-phenotypic associations in two or more traits.
+We used multiple sources of publicly available summary statistics from external studies as data sources for colocalization analyses.
+First, we downloaded Genotype-Tissue Expression Project (GTEx) v8 European ancestry quantitative trait loci (QTL) analysis summary statistics, which contains summary statistics for significant gene expression and splicing QTL variants (eQTL and sQTL, respectively) in 48 body tissues [@gtexconsortium2017].
+As an additional source of QTL data, we used gene expression and DNA methylation QTL (mQTL) analysis summary statistics from studies using tissue taken from the dorsolateral prefrontal cortex of ROSMAP participants [@ng2017].
+These studies examined the associations of genetic variants with molecular traits and provide curated lists of significant QTL variants.
+Finally, we downloaded the summary statistics from a recent GWAS of LOAD for a targeted *post hoc* colocalization analysis in the *TMEM106B* and *GRN* genes [@bellenguez2022].
+
+For each neuropathology endophenotype outcome in our study, we first compiled a list of genetic variants with p-values $\le 1 \times 10^{-5}$ in the pooled single-variant genome-wide association analysis.
+We then queried the lists of significant QTL variants in GTEx and ROSMAP to identify neuropathology-associated QTL variants.
+For each genetic locus associated with neuropathology endophenotypes that had at least one significant QTL in either GTEx or ROSMAP, we performed colocalization analysis using the "coloc.abf" function in the coloc R package [@giambartolomei2014].
+For ordinal variables, we chose dichotomizing cut points to determine case-control proportions.
+We used coloc's default prior probability of colocalization (PrC) of $\text{PrC} = 1\times 10^{-5}$ and considered a posterior $\text{PrC}> 80\%$ as a threshold for evidence of colocalization.
+
+To investigate whether shared GWAS signals drive association among multiple neuropathology endophenotypes, we also performed colocalization analysis on loci with variants exceeding a p-value threshold of $p < 1 \times 10^{-4}$ and concordant effect direction for at least two NPE in the pooled single-variant analysis.
+
+# Results
+
+```{r}
+covar <- fread("data/mega/mega_np.covar")
+nmega <- formatC(covar[, .N], big.mark = ",")
+nnacc <- formatC(covar[FID == "0", .N], big.mark = ",")
+nrosmap <- formatC(covar[FID == "1", .N], big.mark = ",")
+nact <- formatC(covar[FID == "2", .N], big.mark = ",")
+nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")
+```
+
+```{r}
+t1_env <- new.env()
+source("code/manuscript/01_table1_participant_demographics.R",
+       local = t1_env)
+t1dt <- t1_env$t1dt
+```
+
+```{r}
+t1_env$t1
+```
+
+In total, `r nmega` unique participants across all cohorts met our inclusion criteria and were included in our analyses; specifically, autopsy, and genotype data were available for these participants, and they passed our quality control measures for variant missingness, genetic heterogeneity, and ancestry.
+Included participants consisted of `r nnacc` participants from NACC; `r nrosmap` from ROSMAP; `r nact` from ACT; and `r nadni` from ADNI.
+Participant demographics are shown in **Table 1**.
+NACC participants died at a younger age (mean age at death 81 years) compared to ROSMAP (90 years) and ACT (88 years) participants, and had more balanced sex ratios, with 50% of NACC participants female vs.
+68% and 58% in ROSMAP and ACT, respectively.
+NACC participants were also more likely to have an *APOE* $\epsilon 4$ allele (55%) vs.
+ROSMAP (25%) or ACT (28%).
+Across the 11 NPE harmonized across multiple cohorts, NACC and ADNI tended to show more advanced neuropathology than ROSMAP and ACT.
+Using CERAD neuritic plaque score as an example, 65% of NACC and 59% of ADNI participants were rated as "none" for CERAD score, whereas 33% of ROSMAP and 27% of ACT participants were rated as "severe." This trend can be seen in most of the NPE harmonized across cohorts with the exceptions of CAA, LATE-NC, and microinfarcts (see **Supplementary Table S2** for summary demographics for all neuropathology endophenotypes).
+
+## Single-variant analysis identifies novel neuropathology-associated mapped to *COL4A1* and *PIK3R5*
+
+```{r}
+t2_env <- new.env()
+source("code/manuscript/02_table2_mega_analysis_results.R",
+       local = t2_env)
+```
+
+In the initial analysis using only NACC participants, variants within the *APOE* region were significantly associated with Braak NFT stage, CERAD score, CAA, and diffuse amyloid plaques.
+These associations were validated in all of the other data sets at a significance threshold of $p \le 0.05$.
+No other loci were significantly associated with any neuropathology endophenotypes.
+
+In total, 14 independent ($r^2 < 0.05$) loci had lead variants that met the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis.
+Of these, four were duplicate signals in the *APOE* region, leaving 10 significant signals across 11 endophenotypes.
+Novel variant-endophenotype associations included (1) a variant 12 kbp upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis (**Figure 1**) and (2) a *PIK3R5* intronic variant, rs72807981, associated with Braak NFT stage (**Figure 2**).
+Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3**).
+Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak NFT stage; and *TMEM106B* with HS and LATE-NC (**Table 2**).
+
+```{r}
+t2_env$t2
+```
+
+## Gene-based analysis confirms association between genes in *APOE* region and multiple neuropathology endophenotypes
+
+Gene-level analyses further corroborated our single-variant analyses, identifying gene associations with neuropathology endophenotypes in regions with significant variant-phenotype associations.
+*APOE* was significantly associated with NFT, diffuse plaques, CAA, neuritic plaques, and LATE-NC.
+*CYP27C1*, located approximately 50 kbp downstream of *BIN1*, was significantly associated with NFT.
+*TMEM106B* was associated with both HS and LATE-NC.
+
+## Analysis of neuropathology endophenotypic correlations identifies Alzheimer's-associated and vascular clusters
+
+Visual inspection of a dendrogram (**Supplementary Figure S1**) of neuropathology endophenotypes based on polychoric correlations reveals two primary clusters of NPE: (1) a cluster of AD- or amyloid-associated endophenotypes consisting of Braak NFT stage, CERAD score, diffuse amyloid, and CAA; and (2) a cluster of vascular endophenotypes consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts.
+PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster.
+Only the *APOE* region was significantly associated with the AD cluster PC1, while no loci were significantly associated with the vascular PC1.
+
+## Colocalization analyses identifies colocalization between CAA and DNA methylation in *APOC2* region
+
+Because CAA was the only NPE to have an associated locus in the *APOE* region with a signal independent of *APOE* $\epsilon$ diplotype, this region was excluded from endophenotype-endophenotype colocalization analysis.
+Pairs of endophenotypes with the same variant reaching a p-value threshold of $p < 1 \times 10^{-5}$ with concordant effect directions included (1) Braak NFT stage and CERAD score in the *BIN1* region on Chromosome 2 and (2) HS and LATE-NC in the *TMEM106B* region on Chromosome 7 (**Figure 4**).
+Both loci showed a high probability of colocalization, with both having $\text{PrC}> 90\%$.
+
+Multiple suggestive NPE loci showed evidence of colocalization with eQTL in GTEx.
+A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with *GRN* expression ($\text{PrC}>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL.
+The CAA-associated *APOC2* locus independent of *APOE* $\epsilon$ diplotype also colocalized with *APOE* and *APOC2* eQTL in GTEx ($\text{PrC}= 95\%$) and with four mQTL in ROSMAP ($\text{PrC}= 97\%$ for each of the four mQTL, see **Figure 3**).
+
+# Discussion
+
+We performed GWAS and downstream in silico functional analyses for eleven neuropathology endophenotypes across four high-quality neuropathology cohorts with a maximum sample size of 7,463 participants.
+Our work builds on previous genetic association studies of NPE, Alzheimer's, and related dementias and provides another attempt to better understand the complex associations between different neuropathologies and genetic risk factors [@katsumata2022; @shade2022; @beecham2014; @dugan2021; @bellenguez2022; @vattathil2021].
+Our study has several important limitations, including sample size and heterogeneity in study design.
+However, we discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively.
+We also investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
+
+## Limitations and conclusion
+
+There are limitations to the approach we took in this study.
+The four data sources we used, while all of high quality, used different study designs and recruitment strategies.
+NACC participant data arises from recruitment at each ADRC and likely consists of a more clinical population, whereas MAP and ACT participants may be more representative of the broader aged communities from which they are recruited.
+ROS participants include Catholic brothers, sisters, and priests and may represent a population more highly educated than the United States as a whole.
+These differences are reflected in the relative frequencies and distributions of NPE between cohorts, with NACC and ADNI tending to show more severe pathology than ACT and ROSMAP, and their mean ages of death were markedly lower (**Tables 1 and S2**).
+Additionally, there is likely substantial heterogeneity in the way that neuropathology data were collected and graded both within and between studies.
+We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis [@shade2022].
+
+Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is small relative to their clinical counterparts.
+A recent GWAS of AD included over 111,000 clinically diagnosed or "proxy" (based on family history) AD cases and nearly 700,000 controls.
+In contrast, our largest analyzed sample size for CERAD score totaled 7,447 participants, while LATE-NC totaled only 2,224.
+These sample reflect immense progress in autopsy cohorts from even eight years ago and will only continue to increase, but available sample sizes still pose limitations on power of GWAS and similar 'omics approaches to studying disease risk using neuropathology endophenotypes.
+A strength of our study is that data on multiple neuropathology is available for each participant, allowing for multivariate approaches to studying NPE.
+However, as our PCA-based multivariate analyses did not identify any risk loci not identified in single-outcome analyses, this approach alone is not sufficient overcome sample size limitations.
+
+## *COL4A1* and atherosclerosis
+
+One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis (odds ratio $[\text{OR}] =0.71$, $p = 3.9 \times 10^{-8}$).
+*COL4A1* in the brain is preferentially expressed in endothelial cells (**Figure 1)** and codes for a component of collagen IV, an important component of basal lamina.
+In previous GWAS, researchers have reported the *COL4A1/COL4A2* locus to be associated with numerous other vascular phenotypes, including peripheral artery disease, coronary artery disease, stroke, and arteriolar stiffness [@klarin2019, @nikpay2015, @steffensen2018].
+The *COL4A1/COL4A2* locus has also been found to be associated with rare familial cerebrovascular diseases and lacunar ischemic stroke[@blevins2021; @rannikmae2017].
+In a recent GWAS, rs2000660 in particular was a risk variant for migraines [@hautakangas2022].
+The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to vascular disease is possibly related to disruption of the extracellular matrix [@steffensen2018].
+
+rs2000660 was not a QTL in GTEx or ROSMAP.
+However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*.
+rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus [@vattathil2021].
+However, the sample size of the previous study was significantly smaller than the one used in the present study (1,325 vs 6,959).
+Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0079$).
+
+## *PIK3R5* and Braak NFT stage
+
+The other identified novel genome-wide significant locus located on Chromosome 17p13 was associated with Braak NFT stage and had a lead variant of rs72807981 ($\text{OR} =0.71$, $p = 3.9 \times 10^{-8}$, $\text{MAF}= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival.
+This variant was suggestively associated with Braak NFT stage in NACC ($\text{OR} =0.69$, $p = 1.8 \times 10^{-7}$) and was nominally validated in ACT ($\text{OR} =0.50$, $p = 0.028$).
+The same variant was also nominally associated with neuritic plaques in the pooled analysis ($\text{OR} =0.82$, $p = 0.0047$) but was not nominally associated with any other AD NPE.
+rs72807981 is not a QTL in GTEx or ROSMAP, nor were any of the variants in high LD with it in our study.
+There is previous research suggesting that *PIK3R5* may be involved in the development of NFT.
+One study compared differential mRNA expression in aged adults with Braak NFT stage VI vs.
+non-demented controls and found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $[FDR]= 0.01$) [@guennewig2021].
+Interestingly, *PIK3R5* is expressed preferentially in microglial cells in humans (see **Figure 2)**, suggesting that its association with neurofibrillary pathology may be immune-mediated [@zhang2016].
+
+## *APOE* and NPE
+
+Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
+Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE [@beecham2014; @dugan2021].
+While multiple loci in this region were significantly associated with each of these four NPE, only one CAA-associated locus with lead variant rs4803778 remained significant ($\text{OR} = 1.24$, $p = 5.8 \times 10^{-12}$) when adjusting for *APOE* $\epsilon$ diplotype.
+rs4803778 was in high LD ($r^2 = 0.97$) with rs4803779, which was also significantly associated with CAA ($\text{OR} = 1.24$, $p = 6.7 \times 10^{-12}$), is a lead mQTL in ROSMAP for five methylation sites ($p = 3.3 \times 10^{-91}$ for top site cg09555818; see **Figure 3**), and colocalizes with four of these mQTL ($\text{PrC} = 97\%$ for all four sites).
+Furthermore, two of these associated methylation sites, including cg09555818, were significantly associated with dementia in the Generation Scotland: Scottish Family Health Study [@walker2020].
+These results indicate that this locus may be associated with CAA through epigenetic regulation.
+However, this locus is located approximately 30 kbp downstream of the *APOE* coding end site and is situated between the *APOC4*, *APOC2*, and *CLPTM1*.
+While *APOE* may be the most likely target gene of this locus due to its known association with CAA, this inference is not straightforward.
+In GTEx, this locus is a stronger eQTL of *APOC2* than *APOE*, although it is strongly significant and colocalizes with expression of both genes in multiple tissues.
+Future work should examine the regulatory role of the identified methylation sites and test them for association with CAA and other AD NPE.
+
+## HS and LATE-NC
+
+One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $p = 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $p = 1.5 \times 10^{-11}$).
+Additionally, a locus within *GRN* was suggestively associated with HS ($\text{OR} = 0.72$, $p = 6.9 \times 10^{-8}$).
+Both of these genes have been found previously to be associated with frontotemporal lobar degeneration [@rollinson2011; @ciani2019], and in a recent GWAS by Bellenguez et al., both of these genes were identified as risk factors for LOAD [@bellenguez2022a].
+We downloaded available summary statistics from the GWAS Catalog and performed additional colocalization analyses at these loci to investigate if these loci were shared between clinical LOAD phenotypes and NPE [@buniello2019].
+We found that HS ($\text{PrC}= 91\%$) and LATE-NC ($\text{PrC}= 91\%$) colocalized with ADRD at both the *TMEM106B* (HS and LATE-NC both $\text{PrC}= 91\%$) and *GRN* (HS $\text{PrC}\gt 99.9\%$, LATE-NC $\text{PrC}= 90\%$) loci (**Figure 4**).
+HS and LATE-NC also colocalized with each other at these loci, as discussed above.
+These results indicate that HS, LATE-NC, and LOAD likely share causal loci at these genes.
+
+## *BIN1*, neurofibrillary tangles, and neuritic plaques
+
+A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $p = 6.6 \times 10^{-8}$).
+In our colocalization analysis, we verified that the same locus drives association signals with Braak NFT stage and CERAD score.
+In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD [@bellenguez2022].
+However, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $p = 0.15$) nor CAA ($\text{OR} = 1.02$, $p = 0.49$).
+Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology.
+In one study, researchers identified that *BIN1* variant rs744373 is associated with tau-PET but not amyloid-PET levels [@franzmeier2019].
+In another study, authors found that *BIN1* protein expression is correlated with NFT pathology but not with diffuse or neuritic plaques [@holler2014].
+As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles [@nelson2009], our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.
+
+# Conclusions
+
+In conclusion, we identified several promising novel loci associated with NPE and validated multiple known risk loci for AD using NPE.
+We also provided additional context and consideration for relationships between specific risk loci and different NPE.
+Overall, our study provides additional potential avenues of investigation into the relationship between genomics, Alzheimer's disease, and neuropathology.
+
+# List of abbreviations
+
+AD, Alzheimer's disease; LOAD, late-onset Alzheimer's disease; GWAS, genome-wide association study; NFT, neurofibrillary tangle; LATE, limbic-predominant age-related TDP-43 encephalopathy; LATE-NC, LATE neuropathologic change; HS, hippocampal sclerosis; CAA, cerebral amyloid angiopathy; NPE, neuropathology endophenotype; NACC, National Alzheimer's Coordinating Center; ROSMAP, the Religious Orders Study and the Rush Memory and Aging Project; ACT, the Adult Change in Thought Study; ADNI, AD Neuroimaging Initiative; ADRC, Alzheimer's Disease Research Center; NP, neuropathology; ROS, Religious Orders Study; MAP, Rush Memory and Aging Project; PET, positron emission tomography; TOPMed, Trans-'Omics for Precision Medicine; ADGC, Alzheimer's Disease Genetics Consortium; PCA, principal components analysis; GRM, genetic relationship matrix; PC-AiR, PCA in Related Samples; kbp, kilobase pairs; LD, linkage disequilibrium; PC1, first principal component; GTEx, Genotype-Tissue Expression Project; QTL, quantitative trait locus; eQTL, expression QTL; sQTL, splicing QTL; mQTL, methylation QTL; OR, odds ratio; PrC, probability of colocalization.
+
+\newpage
+
+# Declarations
+
+## Ethics approval and consent to participate
+
+All study participants were deceased and the resulting data de-identified, and we exclusively used archival samples.
+Therefore, our study does not fall under the definition of "Human Subjects Research" according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- "involving the collection/study of data or specimens if publicly available, or/or recorded such that subjects cannot be identified."
+
+## Consent for publication
+
+Not applicable.
+
+## Availability of data and materials
+
+All code used for data preparation and analysis is available at <https://www.github.com/lincoln-shade/np_phewas>.
+ROSMAP data can be requested at <https://www.radc.rush.edu> and <https://www.synapse.org>.
+ADGC data is can be requested from NIAGADS at <https://www.niagads.org/resources/related-projects/alzheimers-disease-genetics-consortium-adgc-collection>.
+NACC neuropathology data can be requested at <https://naccdata.org/>.
+ACT data can be requested at <https://actagingresearch.org/>.
+ADNI data can be downloaded at <https://adni.loni.usc.edu/>.
+
+## Competing interests
+
+JAS reported personal fees from Observational Study Monitoring Board Framingham, Observational Study Monitoring Board Discovery (National Institute of Neurological Disorders and Stroke), and Takeda Pharma.
+AJS reported support from Avid Radiopharmaceuticals, a subsidiary of Eli Lilly (in kind contribution of PET tracer precursor); Bayer Oncology (Scientific Advisory Board); Eisai (Scientific Advisory Board); Siemens Medical Solutions USA, Inc.
+(Dementia Advisory Board); NIH NHLBI (MESA Observational Study Monitoring Board); and Springer-Nature Publishing (Editorial Office Support as Editor-in-Chief, Brain Imaging and Behavior).
+All other authors declare that they have no competing interests.
+
+## Funding
+
+**The author(s) disclosed receipt of the following financial support for the research, authorship, and/or publication of this article:** R56AG057191; F30NS124136; P30AG028383; the University of Kentucky Center for Clinical and Translational Science TL-1 Fellowship [grant number TL1TR001997]; the National Center for Advancing Translational Sciences [grant number UL1TR001998]; and the Dean of the College of Medicine, University of Kentucky.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH or the University of Kentucky.
+
+Genotyping was supported by the Alzheimer's Disease Genetics Consortium through the National Institute of Aging [grant numbers U01 AG032984, RC2AG036528].
+
+Samples from the National Cell Repository for Alzheimer's Disease (NCRAD), which receives government support under a cooperative agreement grant (U24 AG21886) awarded by the National Institute on Aging (NIA), were used in this study.
+We thank contributors who collected samples used in this study, as well as patients and their families, whose help and participation made this work possible;
+
+Data for this study were prepared, archived, and distributed by the National Institute on Aging Alzheimer's Disease Data Storage Site (NIAGADS) at the University of Pennsylvania (U24-AG041689-01).
+
+We thank the study participants and staff of the Rush Alzheimer's Disease Center.
+The Religious Orders Study and the Rush Memory and Aging Project are supported by grants from the National Institutes of Health: [grant number P30AG10161, P30AG72975, R01AG15819, R01AG17917, R01AG22018, R01AG33678, R01AG34374, R01AG36042, R01AG40039, R01AG042210, U01AG46152, U01AG61356, R01AG47976, R01AG43379, RF1AG54057, R01AG56352, R01NS78009, and UH2NS100599], and the Illinois Department of Public Health.
+
+The Adult Changes in Thought Study is funded through the National Institute on Aging [grant number U19AG066567].
+
+Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
+ADNI is funded by the National Institute on Aging, the National Institute of Biomedical Imaging and Bioengineering, and through generous contributions from the following: AbbVie, Alzheimer's Association; Alzheimer's Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F.
+Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research & Development, LLC.; Johnson & Johnson Pharmaceutical Research & Development LLC.; Lumosity; Lundbeck; Merck & Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
+The Canadian Institutes of Health Research is providing funds to support ADNI clinical sites in Canada.
+Private sector contributions are facilitated by the Foundation for the National Institutes of Health ([www.fnih.org](https://fnih.org/)).
+The grantee organization is the Northern California Institute for Research and Education, and the study is coordinated by the Alzheimer's Therapeutic Research Institute at the University of Southern California.
+ADNI data are disseminated by the Laboratory for Neuro Imaging at the University of Southern California.
+
+## Authors' contributions
+
+L.M.P.S.
+conceptualized study design, prepared data, performed analyses, and was a major contributor in writing the manuscript.
+Y.K.
+provided feedback on analyses and contributed to the manuscript.
+SC and ME helped prepare figures and provided extensive feedback on manuscript preparation.
+E.L.A.
+provided guidance on interpretation of *BIN1* results and provided extensive feedback on manuscript preparation.
+T.J.H.
+performed imputation and quality control on ROSMAP genotype data.
+K.N.
+and A.J.S.
+provided imputed and quality-controlled ADNI genotype data.
+D.A.B.
+and J.A.S.
+provided ROSMAP neuropathology dataand made critical revisions to the manuscript.
+P.T.N.
+provided guidance on defining neuropathology endophenotypes and contributed to the manuscript.
+D.W.F.
+conceptualized study design and provided feedback on manuscript preparation.
+All authors read and approved the final manuscript.
+
+## Acknowledgements
+
+The NACC database is funded by NIA/NIH Grant U01 AG016976.
+NACC data are contributed by the NIA-funded ADRCs: P30 AG019610 (PI Eric Reiman, MD), P30 AG013846 (PI Neil Kowall, MD), P50 AG008702 (PI Scott Small, MD), P50 AG025688 (PI Allan Levey, MD, PhD), P50 AG047266 (PI Todd Golde, MD, PhD), P30 AG010133 (PI Andrew Saykin, PsyD), P50 AG005146 (PI Marilyn Albert, PhD), P50 AG005134 (PI Bradley Hyman, MD, PhD), P50 AG016574 (PI Ronald Petersen, MD, PhD), P50 AG005138 (PI Mary Sano, PhD), P30 AG008051 (PI Thomas Wisniewski, MD), P30 AG013854 (PI Robert Vassar, PhD), P30 AG008017 (PI Jeffrey Kaye, MD), P30 AG010161 (PI David Bennett, MD), P50 AG047366 (PI Victor Henderson, MD, MS), P30 AG010129 (PI Charles DeCarli, MD), P50 AG016573 (PI Frank LaFerla, PhD), P50 AG005131 (PI James Brewer, MD, PhD), P50 AG023501 (PI Bruce Miller, MD), P30 AG035982 (PI Russell Swerdlow, MD), P30 AG028383 (PI Linda Van Eldik, PhD), P30 AG053760 (PI Henry Paulson, MD, PhD), P30 AG010124 (PI John Trojanowski, MD, PhD), P50 AG005133 (PI Oscar Lopez, MD), P50 AG005142 (PI Helena Chui, MD), P30 AG012300 (PI Roger Rosenberg, MD), P30 AG049638 (PI Suzanne Craft, PhD), P50 AG005136 (PI Thomas Grabowski, MD), P50 AG033514 (PI Sanjay Asthana, MD, FRCP), P50 AG005681 (PI John Morris, MD), P50 AG047270 (PI Stephen Strittmatter, MD, PhD).
+
+\newpage
+
+# References
+
+::: {#refs}
+:::
+
+\newpage
+
+# Figure captions
+
+**Figure 1: Locus near *COL4A1* associates with cerebral atherosclerosis.** **A)** Regional LocusZoom plot of associated locus [@rj2010].
+Mb, megabase; uses hg19.
+**B)** Forest plot for individual and pooled cohorts for lead variant rs2000660 odds ratio and 95% confidence intervals.
+Lead variant reached a suggestive level of significance in NACC analysis ($\text{OR} = 0.72$, $p = 5.8\times10^{-7}$) and was nominally validated in ROSMAP (OR = 0.68, p = 0.0079).
+ADNI results are not shown due to wide confidence interval (OR = 1.25, 95% CI = 0.08-18.84).
+**C)** Human brain cell-type expression profile of *COL4A1* in Zang et al.
+(2016) [@zhang2016].
+*COL4A1* is preferentially expressed in fetal astrocytes and endothelial cells with lower expression in neurons.
+FPKM, Fragments Per Kilobase of transcript per Million mapped reads.
+
+**Figure 2: Intronic *PIK3R5* locus associates with Braak stage.** **A)** Regional LocusZoom plot of Braak-associated locus.
+Mb, megabase; uses hg19.
+**B)** Forest plot for individual and pooled cohorts for rs72807981 odds ratio and 95% confidence intervals.
+Lead variant rs72807981 reached a suggestive level of significance in NACC analysis ($\text{OR} = 0.69$, $p = 1.8\times10^{-7}$) and was nominally validated in ACT ($\text{OR} = 0.50$, $p = 0.0028$).
+**C)** Human brain cell-type expression profile of *PIK3R5* in Zang et al.
+(2016) [@zhang2016].
+*PIK3R5* is primarily expressed in microglia.
+FPKM, Fragments Per Kilobase of transcript per Million mapped reads.
+
+**Figure 3: Locus in *APOE* region associates with CAA and colocalizes with mQTL in ROSMAP.
+A)** Regional LocusZoom of associated locus.
+Mb, megabase; uses hg19.
+Eighteen genes are not shown due to space considerations.
+**B)** Regional plot showing CAA associations and colocalization with four mQTL in ROSMAP with CpG sites: cg04401876, cg09555818, cg10169327, and cg13119609.
+Posterior probability of colocalization equals 97% with each of the four mQTL.
+The darkly bordered box indicates the region with SNPs most associated with CAA and the four colocalizing mQTL.
+
+**Figure 4: HS, LATE-NC, and AD colocalize at *TMEM106B* and *GRN* loci.** Genome positions aligned to hg38.
+**A)** *TMEM106B* locus.
+Posterior probability of colocalization (PrC) equals 91% for each pair of phenotypes.
+**B)** *GRN* locus.
+PrC is greater than 99.9% between HS and AD.
+PrC equals 90% between LATE-NC and AD.
+PrC equals 89% between HS and LATE-NC.
+Key: MB, mega-basepairs; HS, hippocampal sclerosis; AD, clinical or proxy Alzheimer's disease; LATE-NC, limbic-predominant age-related TDP-43 encephalopathy neuropathologic change.

### 2023-01-05 supplementary_materials.Rmd
update manuscript for initial submission

---
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(data.table)
library(flextable)
library(magrittr)
library(table1)
```

# Supplementary Material

## NACC Exclusion Criteria

National Alzheimer's Coordinating Center (NACC) participants were excluded if they had any of the following conditions noted in the NACC Neuropathology Data Set. Variable names and descriptions are taken from <https://files.alz.washington.edu/documentation/rdd-np.pdf>. Variable descriptions may be lightly edited. Participants were not excluding for missing data in any of these fields.

```{r}
table_s1 <- fread("doc/nacc_exclusion_variables.txt", sep = "|") %>% 
  flextable() %>% 
  add_header_lines("Table S1: NACC Exclusion Criteria")

table_s1
```

\newpage

```{r}

pord <- fread("data/mega/mega_np_ord.pheno", na.strings = '-1')
pbin <- fread("data/mega/mega_np.pheno", na.strings = '-1')
pheno <- merge(pord, pbin, c("FID", "IID"))
covar <- fread("data/mega/mega_np.covar")
covar_apoe <- fread("data/mega/mega_np_apoe.covar", stringsAsFactors = TRUE)
covar <- merge(covar, 
               covar_apoe[, .(FID, IID, apoe)], 
               c("FID", "IID"), 
               all.x = TRUE)
covar <- covar[, .(FID, IID, msex, age_death, apoe)]
covar[, Sex := factor(msex, labels = c("Female", "Male"))]

dt <- merge(pheno, covar, by = c("FID", "IID"))
setnames(dt, "age_death", "Age of Death")
dt[, Study := factor(FID, labels = c("NACC", "ROSMAP", "ACT", "ADNI"))]
dt[, CERAD := ordered(cerad, 
                        labels = c("None", "Mild", "Moderate", "Severe"))]

# Display apoe variable as number of APOE e4 alleles for table
dt[apoe %in% c("apoe22", "apoe23", "apoe33"), 
     `APOE e4 alleles` := 0]
dt[apoe %in% c("apoe24", "apoe34"), 
     `APOE e4 alleles` := 1]
dt[apoe %in% c("apoe44"), 
     `APOE e4 alleles` := 2]
dt[, `APOE e4 alleles` := ordered(`APOE e4 alleles`)]

# add all other NPE used as outcomes
dt[, `Braak Stage` := ordered(braak)]
dt[, `Diffuse Amyloid` := 
     ordered(diffuse_abeta, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, Arteriolosclerosis := 
     ordered(arteriol, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, Atherosclerosis := 
     ordered(atheroscler, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, CAA := 
     ordered(caa_ord, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, `LATE-NC` := 
     ordered(late, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, `Lewy Body` := 
     ordered(lewy_body, 
             labels = c("None", "Mild", "Moderate", "Severe"))]
dt[, HS := factor(hs, labels = c("Absent", "Present"))]
dt[, Microinfarcts := factor(microinf, labels = c("Absent", "Present"))]
dt[, `Gross infarcts` := factor(grossinf, labels = c("Absent", "Present"))]

t1dt <- 
  as.data.frame(
    table1(~ Sex + `Age of Death` + `APOE e4 alleles` + CERAD + 
             `Braak Stage`+ `Diffuse Amyloid` + Arteriolosclerosis +
             Atherosclerosis + CAA + `LATE-NC` + `Lewy Body` + HS +
             Microinfarcts + `Gross infarcts`| Study,
           data = dt
           )
    )

t1 <- flextable(t1dt) %>% 
  autofit() %>% 
  padding(padding = 2, part = "all") %>%
  add_header_lines("Table S2: Participant Demographics") %>% 
  add_footer_lines(paste0("Key: SD, standard deviation; Min, minimum; ",
                          "Max, maximum. LATE-NC, LATE neuropathologic change",
                          " HS, Hippocampal sclerosis",
                          "Age of Death variable is integer ",
                          "in NACC, ADNI, and ACT but continuous in ROSMAP."))

# saveRDS(t1, 
#         file = "output/manuscript/tableS1_participant_demographics.Rds")
t1
```

\newpage

![Figure S1: Dendrogram of Neuropathology Endophenotype Polychoric Correlations](doc/figure_S1_dendrogram.png)

### 2023-01-05 code/gwas/write_snp_list.R

@@ -12,9 +12,14 @@ parser$add_argument("-f", "--file", help = "text file SNPs as a column")
 parser$add_argument("--snp_col", default = "SNP", help = "SNP column name")
 parser$add_argument("-o", "--out", default = "./snp_list.txt",
                     help = "output filename")
+parser$add_argument("--pval_col", help = "p-value column for use with --max_pval")
+parser$add_argument("--max_pval", help = "optional pval threshold for SNPs")
 args <- parser$parse_args()
 
 snps <- fread(args$file)
+if (!is.null(args$max_pval) & !is.null(args$pval_col)) {
+  snps <- snps[get(args$pval_col) <= as.numeric(args$max_pval)]
+}
 fwrite(snps[, .(get(args$snp_col))],
        file = args$out,
        col.names = FALSE,

### 2023-01-05 code/manuscript/forestplots.R

@@ -1,0 +1,115 @@
+library(data.table)
+library(ggplot2)
+library(forestplot)
+
+make_95CI <- function(or, p, ci=0.95) {
+  beta <- log(or)
+  lt_zero <- beta < 0
+  z <- qnorm(p/2, lower.tail = TRUE)
+  sd <- beta / z
+  zrad <- numeric(length = length(p))
+  for (i in 1:length(p)){
+  zrad[i] <- qnorm((1 - ci) / 2, lower.tail = lt_zero[i])
+  }
+  lower <- round(exp(beta + zrad * sd), 2)
+  upper <- round(exp(beta - zrad * sd), 2)
+  return(list(lower, upper))
+}
+
+#-----------------------------
+# Braak stage PIK3R5 locus
+#-----------------------------
+
+braak_mega <- fread("output/gwas/mega/polmm/braak_polmm_results.txt")
+braak_nacc <- fread("output/gwas/adc/polmm/braak_polmm_results.txt")
+braak_rosmap <- fread("output/gwas/rosmap/polmm/braak_polmm_results.txt")
+braak_act <- fread("output/gwas/act/polmm/braak_polmm_results.txt")
+braak_adni <- fread("output/gwas/adni/polmm/braak_polmm_results.txt")
+pik3r5_snp <- "rs72807981"
+dt <- rbind(braak_nacc[chr == 17][SNPID == pik3r5_snp], 
+                              braak_rosmap[chr == 17][SNPID == pik3r5_snp], 
+                              braak_act[chr == 17][SNPID == pik3r5_snp], 
+                              braak_adni[chr == 17][SNPID == pik3r5_snp],
+                              braak_mega[chr == 17][SNPID == pik3r5_snp])
+dt[, Study := c("NACC", "ROSMAP", "ACT", "ADNI", "Pooled")]
+dt[, Lower := make_95CI(exp(beta), pval.spa)[[1]]]
+dt[, Upper := make_95CI(exp(beta), pval.spa)[[2]]]
+dt <- dt[Study != "ADNI"]
+dt[, index := .N:1]
+
+# 331x940 px
+dt[, ggplot(data = .SD, aes(exp(beta), index, xmin=Lower, xmax=Upper))] +
+  geom_point() +
+  theme_minimal() +
+  theme(text = element_text(size = 16)) +
+  geom_errorbarh(height=0.15) +
[…]

### 2023-01-05 code/rnaseq/analysis.R

@@ -1,0 +1,49 @@
+#=============================================================================
+# Compare gene expression of COL4A1 and PIK3R5 with atherosclerosis and Braak
+#=============================================================================
+
+library(data.table)
+library(ordinal)
+library(readxl)
+library(ggplot2)
+rnaseq <- fread("data/rnaseq/rnaseq.txt.gz")
+rnaseq[, projid := as.character(projid)]
+# pheno <- fread("data/mega/mega_np_ord.pheno")
+pheno <- read_xlsx("/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx")
+setDT(pheno)
+# covar <- fread("data/mega/mega_np.covar")
+# dt <- merge(pheno, covar)
+dt <- pheno[, .(projid, braaksc, cvda_4gp2, pmi, study, age_death, msex)]
+col4a1 <- "ENSG00000187498"
+pik3r5 <- "ENSG00000141506"
+col4a1_col <- colnames(rnaseq)[grep(col4a1, colnames(rnaseq))]
+pik3r5_col <- colnames(rnaseq)[grep(pik3r5, colnames(rnaseq))]
+dt <- merge(dt, rnaseq[, c("projid", "RIN", ..col4a1_col, ..pik3r5_col)], by = "projid")
+
+col4a1_lm_null <- dt[!is.na(cvda_4gp2), lm(log(ENSG00000187498.9) ~ msex + age_death + RIN +pmi,
+                     data = .SD)]
+col4a1_lm <- dt[, lm(log(ENSG00000187498.9) ~ ordered(cvda_4gp2) + msex + age_death + RIN +pmi,
+                     data = .SD)]
+anova(col4a1_lm_null, col4a1_lm)
+
+pik3r5_lm_null <- dt[!is.na(braaksc), 
+                     lm(log(ENSG00000141506.9) ~ msex + age_death + RIN +pmi,
+                        data = .SD)]
+pik3r5_lm <- dt[, lm(log(ENSG00000141506.9) ~ (braaksc > 3) + msex + age_death + RIN +pmi,
+                     data = .SD)]
+anova(pik3r5_lm_null, pik3r5_lm)
+
+col4a1_clm <- dt[, clm(ordered(cvda_4gp2) ~ log(ENSG00000187498.9)+ study + msex + age_death + RIN +pmi,
+                       data = .SD)]
[…]

### 2023-01-05 code/tmp_table.R

@@ -1,0 +1,46 @@
+
+library(data.table)
+library(magrittr)
+library(flextable)
+library(table1)
+
+pord <- fread("data/mega/mega_np_ord.pheno", na.strings = '-1')
+pbin <- fread("data/mega/mega_np.pheno", na.strings = '-1')
+pheno <- merge(pord, pbin, c("FID", "IID"))
+covar <- fread("data/mega/mega_np.covar")
+covar_apoe <- fread("data/mega/mega_np_apoe.covar", stringsAsFactors = TRUE)
+covar <- merge(covar, 
+               covar_apoe[, .(FID, IID, apoe)], 
+               c("FID", "IID"), 
+               all.x = TRUE)
+covar <- covar[, .(FID, IID, msex, age_death, apoe)]
+covar[, Sex := factor(msex, labels = c("Female", "Male"))]
+
+dt <- merge(pheno, covar, by = c("FID", "IID"))
+setnames(dt, "age_death", "Age of Death")
+dt[, Study := factor(FID, labels = c("NACC", "ROSMAP", "ACT", "ADNI"))]
+dt[, CAA := ordered(caa_ord, 
+                      labels = c("None", "Mild", "Moderate", "Severe"))]
+
+# Display apoe variable as number of APOE e4 alleles for table
+dt[apoe %in% c("apoe22", "apoe23", "apoe33"), 
+   `APOE e4 alleles` := 0]
+dt[apoe %in% c("apoe24", "apoe34"), 
+   `APOE e4 alleles` := 1]
+dt[apoe %in% c("apoe44"), 
+   `APOE e4 alleles` := 2]
+dt[, `APOE e4 alleles` := ordered(`APOE e4 alleles`)]
+
+t1dt <- 
+  as.data.frame(
+    table1(~ Sex + `Age of Death` + `APOE e4 alleles` + CAA | Study,
+           data = dt
+    )
+  ) 
+
+t1 <- flextable(t1dt) %>% 
+  autofit() %>% 
+  padding(padding = 2, part = "all") %>%
+  add_header_lines("Table 1: Participant Demographics")
+
+t1

### 2023-02-02 manuscript.Rmd
add Mayo BB data

@@
 
-Proposed journal: ***Molecular Neurodegeneration***
+Proposed journal: **???**
 
@@
 Associations between the Chromosome 19 CAA risk locus, CpG methylation, and CAA pathology are consistent with the effect of this locus on disease being mediated by DNA methylation.
-Furthermore, the same two CpG sites are associated with dementia in an independent cohort, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk.
+Furthermore, the same two CpG sites are associated with dementia risk factors in independent cohorts, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk.
 The atherosclerosis risk variant rs2000660 is in strong linkage disequilibrium with a synonymous coding variant (rs650724) of *COL4A1*, providing a candidate functional variant.
@@
 The most common form of AD, late-onset AD (LOAD), presents in individuals aged 65 years and older and is highly heritable, with twin and family studies estimating its heritability at \~60% [@ridge2013; @gatz1997; @gatz2006].
-Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, immune response, and extracellular matrix regulation [@jansen2019; @wightman2021; @bellenguez2022].
+Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, tau, and immune response [@jansen2019; @wightman2021; @bellenguez2022].
 Most dementia GWAS to date have focused on clinically diagnosed LOAD or proxy outcomes based on family history of dementia [@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019]; however, late-onset dementia diagnosed as LOAD is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors [@farfel2016; @katsumata2022; @karanth2020].
@@
 Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and hippocampal sclerosis [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies [@arvanitakis2016; @ighodaro2017].
-Collectively, these factors reveal an increasingly complex and synergistic web of pathologies contributing to cognitive impairment and dementia, and demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.
+Collectively, these factors reveal an increasingly complex and synergistic web of pathologies which contribute to cognitive impairment and dementia.
 
-The investigation of genetic risk of neuropathologies is an attractive approach for uncovering biological pathways involved in the development of LOAD and related dementias.
-Neuropathologies are commonly operationalized in aged autopsy cohorts as binary or semi-quantitative (or ordinal) neuropathology endophenotypes (NPE).
-Prior GWAS of neuropathology endophenotypes have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022].
+```{=html}
+<!-- Insert paragraph on how genetic risk factors of clinical alzheimer's may be associated with specific neuropathologies and pathways, and how studying the genomic risk factors of individual neuropathologies may provide a finer grained understanding of the genetic risk of AD
+
+and they demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.-->
+```
+
+The reality of clinically diagnosed AD being a heterogeneous catch-all clinical syndrome with multiple and varied pathologies found in the brains of those diagnosed has implications for the interpretation of AD GWAS (which can often be more accurately described as dementia GWAS).
+Each individual risk loci identified in large AD GWAS may be associated with some brain pathologies but not others.
+If this is so, we would expect that investigation into the pathologies themselves may provide a reasonable alternative approach to discovering relevant risk genes and biological pathways involved in the development of LOAD and related dementias.
+Prior GWAS of neuropathology endophenotypes (NPE) have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022].
 However, continued recruitment in ongoing autopsy cohorts and methodological development for GWAS and downstream functional analysis provides additional opportunity for in-depth investigation into multiple neuropathologies studied across multiple cohorts.
@@
 In these cases, records in the NACC were preferentially kept in order to maximize sample size of the initial analysis using only NACC participants (see Statistical Analyses subsection).
+An overview of our study design is presented in **Supplementary Figure S1**.
 
@@
 
+## DNA Methylation Data
+
+Pre-processed and quality-controlled DNA methylation data for 740 ROSMAP participants were downloaded from Synapse.org (Synapse IDs: syn3157275 and syn3191087).
+DNA methylation preparation and quality control measures have been previously described [@yu2015; @dejager2018].
+Briefly, approximately 50mg of frozen gray matter tissue from the dorsolateral prefrontal cortex were sampled from each participant.
+DNA was then extracted and processed using the Illumina Infinium HumanMethylation450 BeadChip.
+Quality control measured included remove low-quality probes; removing participants with poor bisulfite-conversion efficiency; and adjusting methylation levels by age, sex, and batch, which adequately controlled for batch effects.
+Missing methylation levels were imputed using 100-nearest neighbors [@yu2015; @dejager2018].
+
 ## Statistical Analyses
@@
 
+## Association between CpG site methylation and cerebral amyloid angiopathy
+
+In our *APOE* $\epsilon$-adjusted genetic association analysis, one locus near *APOE* remained significantly associated with cerebral amyloid angiopathy.
+We found that this locus colocalized with DNA methylation levels at four CpG sites in ROSMAP.
+To investigate whether these CpG sites were themselves association with CAA pathology, we combined connected individual-level DNA methylation data and neuropathological data in ROSMAP for analysis.
+We used cumulative logit models using the "clm" function implemented in the "ordinal" R package with the semi-quantitative CAA variable described above as the outcome for analysis.
+We performed four analyses, with one of each of the four CpG sites tested as the independent variable of interest in each analsyis.
+We adjusted for age, sex, ROS vs MAP study, bisulfite conversion efficiency, post-mortem interval, and *APOE* $\epsilon$ genotype in each analysis.
+Z-values from the resulting parameter estimates were used as test statistics for statistical significance.
+
 # Results
@@
 
-## Single-variant analysis identifies novel neuropathology-associated mapped to *COL4A1* and *PIK3R5*
+## GWAS identifies novel neuropathology-associated variants mapped to *APOC2,* *COL4A1,* and *PIK3R5*
 
@@
 Novel variant-endophenotype associations included (1) a variant 12 kbp upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis (**Figure 1**) and (2) a *PIK3R5* intronic variant, rs72807981, associated with Braak NFT stage (**Figure 2**).
-Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3**).
+Additionally, one locus between *APOC2* and *CLPTM1* with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3a**).
 Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak NFT stage; and *TMEM106B* with HS and LATE-NC (**Table 2**).
@@
 
-Visual inspection of a dendrogram (**Supplementary Figure S1**) of neuropathology endophenotypes based on polychoric correlations reveals two primary clusters of NPE: (1) a cluster of AD- or amyloid-associated endophenotypes consisting of Braak NFT stage, CERAD score, diffuse amyloid, and CAA; and (2) a cluster of vascular endophenotypes consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts.
+Visual inspection of a dendrogram (**Supplementary Figure S2**) of neuropathology endophenotypes based on polychoric correlations reveals two primary clusters of NPE: (1) a cluster of AD- or amyloid-associated endophenotypes consisting of Braak NFT stage, CERAD score, diffuse amyloid, and CAA; and (2) a cluster of vascular endophenotypes consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts.
 PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster.
@@
 
-## Colocalization analyses identifies colocalization between CAA and DNA methylation in *APOC2* region
+## Colocalization analyses identifies colocalization between cerebral amyloid angiopathy, *APOC2* expression, and CpG methylation near *APOC2* in prefrontal cortex
 
@@
 A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with *GRN* expression ($\text{PrC}>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL.
-The CAA-associated *APOC2* locus independent of *APOE* $\epsilon$ diplotype also colocalized with *APOE* and *APOC2* eQTL in GTEx ($\text{PrC}= 95\%$) and with four mQTL in ROSMAP ($\text{PrC}= 97\%$ for each of the four mQTL, see **Figure 3**).
+In total, 42 NPE loci colocalized with QTL with $\text{PrC}= \ge80\%$, with 271 total colocalizing NPE-QTL pairs across 43 tissues (**Supplementary Table S4**).
+
+The *APOE* $\epsilon$-independent Chromosome 19 CAA-associated locus colocalized with expression of multiple genes in GTEx: *APOE* expression in the wall of the aorta; *8CLPTM1* expression in the skin of the leg and suprapubic region; and *APOC2* expression in 14 different tissues, including the brain cortex, caudate, nuclues accumbens, and cerebellum ($\text{PrC}= 90-97\%$).
+Additionally, the same locus colocalized with mQTL for four CpG sites in ROSMAP, cg09555818, cg04401876, cg10169327, and cg13119609 ($\text{PrC}= 97\%$, see **Figure 3b**).
+
+## DNA methylation levels at CpG sites cg09555818 and cg13119609 in the cerebral cortex are associated with cerebral amyloid angiopathy
+
+In total, 708 ROSMAP participants had DNA methylation and CAA data available for analysis.
+We test methylation levels at each of the four CpG sites that colocalize with the *APOE* $\epsilon$-independent CAA risk locus on Chromosome 19 -- cg09555818, cg04401876, cg10169327, and cg13119609 -- for association with CAA pathology.
+Of these, hypomethylation of cg09555818 (odds ratio $[\text{OR}] =0.82$, $p = 0.004$) and cg13119609 ($\text{OR} =0.78$, $p = 0.0007$) were significantly associated with worse CAA pathology (see **Figure 3c**).
 
@@
 Our study has several important limitations, including sample size and heterogeneity in study design.
-However, we discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively.
-We also investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
+However, we identified a novel APOE $\epsilon$-independent CAA risk locus which also affects *APOC2* brain expression and dementia-associated CpG methylation sites, which in turn were also associated with CAA risk.
+We also discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively.
+Lastly, we investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
 
-## Limitations and conclusion
+## Limitations
 
@@
 We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis [@shade2022].
+Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is modest relative to their clinical and proxy outcomes.
 
-Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is small relative to their clinical counterparts.
-A recent GWAS of AD included over 111,000 clinically diagnosed or "proxy" (based on family history) AD cases and nearly 700,000 controls.
-In contrast, our largest analyzed sample size for CERAD score totaled 7,447 participants, while LATE-NC totaled only 2,224.
-These sample reflect immense progress in autopsy cohorts from even eight years ago and will only continue to increase, but available sample sizes still pose limitations on power of GWAS and similar 'omics approaches to studying disease risk using neuropathology endophenotypes.
-A strength of our study is that data on multiple neuropathology is available for each participant, allowing for multivariate approaches to studying NPE.
-However, as our PCA-based multivariate analyses did not identify any risk loci not identified in single-outcome analyses, this approach alone is not sufficient overcome sample size limitations.
+## Known AD risk loci are associated with specific neuropathological features
+
+Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
+Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE [@beecham2014; @dugan2021].
+Notably, *APOE* was not associated with vascular pathology except for CAA.
+
+A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $p = 6.6 \times 10^{-8}$).
+We verified through colocalization analysis that the same locus drives association signals with Braak NFT stage and CERAD score.
+In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD [@bellenguez2022].
+Interestingly, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $p = 0.15$) nor CAA ($\text{OR} = 1.02$, $p = 0.49$).
+Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology [@holler2014; @franzmeier2019].
+As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles [@nelson2009], our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.
+
+One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $p = 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $p = 1.5 \times 10^{-11}$).
+Additionally, a locus within *GRN* was suggestively associated with HS ($\text{OR} = 0.72$, $p = 6.9 \times 10^{-8}$).
+Both of these genes are associated with frontotemporal lobar degeneration [@rollinson2011; @ciani2019], and with LOAD [@bellenguez2022a].
+We downloaded available summary statistics from the GWAS Catalog and performed additional colocalization analyses at these loci to investigate if these loci were shared between clinical LOAD phenotypes and NPE [@buniello2019].
+We found that HS, LATE-NC, and LOAD all colocalize at these two loci ($\text{PrC}\ge 90\%$, **Figure 4**).
+These results indicate that HS, LATE-NC, and LOAD likely share causal loci for these genes.
+
+# *APOE* $\epsilon$-independent *APOC2* is associated with CAA
+
+When we re-ran Chromosome 19 analyses on these phenotypes while adjusting for *APOE* $\epsilon$ genotype, only one CAA-associated locus with lead variant rs4803778 remained significant ($\text{OR} = 1.24$, $p = 5.8 \times 10^{-12}$).
+Several variants in this locus were lead eQTL for *APOC2* brain expression in GTEx and mQTL in ROSMAP for four methylation sites ($p = 3.3 \times 10^{-91}$ for top site cg09555818; see **Figure 3**).
+Colocalization analysis confirmed that the CAA risk locus shares a functional variants with both *APOC2* eQTL ($\text{PrC} = 89-97\%$ in brain tissues) and nearby mQTL ($\text{PrC} = 97\%$ for all four CpG sites).
+We confirmed that two of the CpG sites affected by the CAA risk locus, cg09555818 ($\text{OR} =0.82$, $p = 0.004$) and cg13119609 ($\text{OR} =0.78$, $p = 0.0007$), were in turn significantly associated with CAA pathology.
+Both of these CpG sites are located within the *APOC4-APOC2* readthrough transcript region overlapping *APOC4* and *APOC2*.
+We then confirmed that the directions of effects for rs4803779 risk allele, CpG methylation, and CAA risk were consistent with the hypothesis that CpG methylation mediates the genotype-phenotype association.
+Our results are consistent with the hypothesis that rs4803779 locus affects CAA risk through hypomethylation of CpG sites in the *APOC4-APOC2* region.
+These results are also consistent with previous studies in other human cohorts that implicate hypomethylation at cg09555818 and cg13119609 as potentially associated with Alzheimer's disease [@walker2021; @walker2020; @shao2018].
+Circumstantial evidence indicates that *APOC2* may be the target gene of the rs4803779 risk locus as it colocalizes with *APOC2* expression in multiple brain tissues in GTEx and the associated CpG sites are located in exon 3 of *APOC4-APOC2*.
+However, more research will need to be done to verify that *APOC2* is the target gene.
 
@@
 
-One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis (odds ratio $[\text{OR}] =0.71$, $p = 3.9 \times 10^{-8}$).
+One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis ( $\text{OR} =0.71$, $p = 3.9 \times 10^{-8}$).
 *COL4A1* in the brain is preferentially expressed in endothelial cells (**Figure 1)** and codes for a component of collagen IV, an important component of basal lamina.
+rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus [@vattathil2021].
+However, the sample size of the previous study was significantly smaller than the one used in the present study (1,325 vs 6,959).
+Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0079$).
+
+rs2000660 was not a QTL in GTEx or ROSMAP.
+However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*.
+As rs2000660 does not have an obvious functional role and synonymous variants in some genes have been shown to alter mRNA stability, protein conformation, and other regulatory functions [@sauna2011], rs650724 presents as a possible functional variant driving genotype-atheroscerosis assosiation in this locus.
 In previous GWAS, researchers have reported the *COL4A1/COL4A2* locus to be associated with numerous other vascular phenotypes, including peripheral artery disease, coronary artery disease, stroke, and arteriolar stiffness [@klarin2019, @nikpay2015, @steffensen2018].
@@
 The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to vascular disease is possibly related to disruption of the extracellular matrix [@steffensen2018].
-
-rs2000660 was not a QTL in GTEx or ROSMAP.
-However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*.
-rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus [@vattathil2021].
-However, the sample size of the previous study was significantly smaller than the one used in the present study (1,325 vs 6,959).
-Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0079$).
 
@@
 non-demented controls and found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $[FDR]= 0.01$) [@guennewig2021].
-Interestingly, *PIK3R5* is expressed preferentially in microglial cells in humans (see **Figure 2)**, suggesting that its association with neurofibrillary pathology may be immune-mediated [@zhang2016].
-
-## *APOE* and NPE
-
-Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
-Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE [@beecham2014; @dugan2021].
-While multiple loci in this region were significantly associated with each of these four NPE, only one CAA-associated locus with lead variant rs4803778 remained significant ($\text{OR} = 1.24$, $p = 5.8 \times 10^{-12}$) when adjusting for *APOE* $\epsilon$ diplotype.
-rs4803778 was in high LD ($r^2 = 0.97$) with rs4803779, which was also significantly associated with CAA ($\text{OR} = 1.24$, $p = 6.7 \times 10^{-12}$), is a lead mQTL in ROSMAP for five methylation sites ($p = 3.3 \times 10^{-91}$ for top site cg09555818; see **Figure 3**), and colocalizes with four of these mQTL ($\text{PrC} = 97\%$ for all four sites).
-Furthermore, two of these associated methylation sites, including cg09555818, were significantly associated with dementia in the Generation Scotland: Scottish Family Health Study [@walker2020].
-These results indicate that this locus may be associated with CAA through epigenetic regulation.
-However, this locus is located approximately 30 kbp downstream of the *APOE* coding end site and is situated between the *APOC4*, *APOC2*, and *CLPTM1*.
-While *APOE* may be the most likely target gene of this locus due to its known association with CAA, this inference is not straightforward.
-In GTEx, this locus is a stronger eQTL of *APOC2* than *APOE*, although it is strongly significant and colocalizes with expression of both genes in multiple tissues.
-Future work should examine the regulatory role of the identified methylation sites and test them for association with CAA and other AD NPE.
-
-## HS and LATE-NC
-
-One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $p = 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $p = 1.5 \times 10^{-11}$).
-Additionally, a locus within *GRN* was suggestively associated with HS ($\text{OR} = 0.72$, $p = 6.9 \times 10^{-8}$).
-Both of these genes have been found previously to be associated with frontotemporal lobar degeneration [@rollinson2011; @ciani2019], and in a recent GWAS by Bellenguez et al., both of these genes were identified as risk factors for LOAD [@bellenguez2022a].
-We downloaded available summary statistics from the GWAS Catalog and performed additional colocalization analyses at these loci to investigate if these loci were shared between clinical LOAD phenotypes and NPE [@buniello2019].
-We found that HS ($\text{PrC}= 91\%$) and LATE-NC ($\text{PrC}= 91\%$) colocalized with ADRD at both the *TMEM106B* (HS and LATE-NC both $\text{PrC}= 91\%$) and *GRN* (HS $\text{PrC}\gt 99.9\%$, LATE-NC $\text{PrC}= 90\%$) loci (**Figure 4**).
-HS and LATE-NC also colocalized with each other at these loci, as discussed above.
-These results indicate that HS, LATE-NC, and LOAD likely share causal loci at these genes.
-
-## *BIN1*, neurofibrillary tangles, and neuritic plaques
-
-A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $p = 6.6 \times 10^{-8}$).
-In our colocalization analysis, we verified that the same locus drives association signals with Braak NFT stage and CERAD score.
-In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD [@bellenguez2022].
-However, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $p = 0.15$) nor CAA ($\text{OR} = 1.02$, $p = 0.49$).
-Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology.
-In one study, researchers identified that *BIN1* variant rs744373 is associated with tau-PET but not amyloid-PET levels [@franzmeier2019].
-In another study, authors found that *BIN1* protein expression is correlated with NFT pathology but not with diffuse or neuritic plaques [@holler2014].
-As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles [@nelson2009], our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.
+Interestingly, *PIK3R5* is expressed preferentially in microglial cells in humans (see **Figure 2**), suggesting that its association with neurofibrillary pathology may be immune-mediated [@zhang2016].
 
@@
 ADNI data can be downloaded at <https://adni.loni.usc.edu/>.
+The results published here are in whole or in part based on data obtained from the AD Knowledge Portal.
 
@@
 
+The results published here are in whole or in part based on data obtained from the AD Knowledge Portal.
+
 \newpage
@@
 
-**Figure 3: Locus in *APOE* region associates with CAA and colocalizes with mQTL in ROSMAP.
+**Figure 3: Locus in *APOE/APOC2* region associates with CAA and colocalizes with mQTL in ROSMAP.
 A)** Regional LocusZoom of associated locus.
@@
 The darkly bordered box indicates the region with SNPs most associated with CAA and the four colocalizing mQTL.
+**C)** Associations between the same four CpG sites and CAA pathology in ROSMAP (N = 708).
+Two sites, cg09555818 (*p* = 0.004) and cg13119609 (*p* = 0.0007) were significantly associated with CAA pathology.
 

### 2023-02-02 code/coloc/00_coloc_pipeline.sh

@@ -1,22 +1,0 @@
-#!/bin/bash
-phenotype=$1
-
-# the plink2 glm results for the phenotype
-results=output/gwas/mega/mega."$phenotype".glm.logistic
-# the prefix for the plink2 binary fileset used for analysis
-data_prefix=data/mega/mega_np
-
-bash code/coloc/01_qtls.sh \
-  $phenotype \
-  $results
-
-top_qtls_file=tmp/top_qtls_"$phenotype".tmp
-
-if test -f $top_qtls_file; then
-    bash code/coloc/02_coloc.sh \
-      $data_prefix \
-      "$data_prefix".pheno \
-      $results \
-      output/coloc/mega/ \
-      $phenotype
-fi

### 2023-02-02 code/coloc/02_coloc.sh

@@ -1,55 +1,0 @@
-#!/bin/bash
-
-#====================================
-# Run colocalization analysis
-#====================================
-
-# command-line arguments
-# 1: PLINK binary fileset prefix
-# 2: PLINK .pheno file name
-# 3: PLINK .assoc.logistic file name
-# 4: relative path for colocalization results output directory 
-# (append with "/" e.g. ./ or data/)
-
-# calculate minor allele frequencies for each chromosome 
-# that has at least 1 qtl
-pfile=$1
-pheno=$2
-regression=$3
-out_folder=$4
-gwas_pheno=$5
-
-chrs=tmp/chrs_"$gwas_pheno".tmp
-maf_out=tmp/chr%_maf_"$gwas_pheno"
-cat $chrs | xargs -n 1 -P 10 -I % \
-  plink2 \
-    --pfile $pfile \
-    --chr % \
-    --freq \
-    --out "$maf_out"
-
-# prepare gwas summary stats data
-
-
-cat $chrs | xargs -n 1 -P 10 -I % \
-  Rscript code/coloc/02a_gwas_sumstats.R \
-    -p $pheno \
-    -f "$maf_out".afreq \
-    -r $regression \
-    --chr % \
-    --phenotype $gwas_pheno
-
-# merge gwas and qtl data
-cat tmp/top_qtls_"$gwas_pheno".tmp | \
-  awk 'NR > 1 {print$1,$2,$3,$4}' | \
-  xargs -P 22 -L 1 bash -c \
-  'code/coloc/02b_gwas_qtl.R -g $0 -q $1 -t $2 --chr $3 --qtl $4' $gwas_pheno
-
-
-# run coloc
-p12=0.00001
-cat tmp/top_qtls_"$gwas_pheno".tmp | \
-  awk 'NR>1 {print$1,$2,$3}' | \
-  xargs -L 1 bash -c \
-  './code/coloc/02c_coloc.R -f tmp/chr"$5"_"$3"_"$4"_gwas_qtl_"$2".tmp --p12 $0 -o "$1""$2"_coloc_results.txt -q $3 -t $4 --chr $5' \
-  $p12 $out_folder $gwas_pheno

### 2023-02-02 code/coloc/coloc_all_pheno.sh

@@ -1,3 +1,0 @@
-#!/bin/bash
-
-cat data/mega/mega_np.pheno | awk 'NR==1{for (i=3; i<=NF; i++) print $i}' | xargs -n 1 bash code/coloc/00_coloc_pipeline.sh

### 2023-02-02 code/coloc/coloc_binary_phenos.py

@@ -1,0 +1,24 @@
+
+# run in coloc directory
+import subprocess
+
+for pheno in ['grossinf', 'microinf']:
+    cmd = [
+        "python", "./coloc_qtl.py",
+        "--out_pref", "../np_phewas/output/coloc/mega/" + pheno + "/",
+        "--gwas_sumstats", 
+        "../np_phewas/output/gwas/mega/saige/" + pheno + "_saige_results.txt",
+        "--max_pval", "1e-5", 
+        "--pval_col", "p.value",
+        "--chrom_col", "CHR",
+        "--snp_col", "MarkerID",
+        "--maf_col", "AF_Allele2",
+        "--phenotype_type", "cc",
+        "--phenotype_col", pheno,
+        "--phenotype_file", "../np_phewas/data/mega/mega_np.pheno",
+        "--phenotype_na_string", "-1", 
+        "--gtex", 
+        "--rosmap"
+    ]
+    
+    subprocess.run(cmd)

### 2023-02-02 code/coloc/coloc_braak_bin1_susie.R

@@ -1,0 +1,69 @@
+library(data.table)
+library(gwasglue)
+library(coloc)
+
+calc_betavar_from_beta_pval <- function(beta, pval) {
+  z <- qnorm(pval / 2, lower.tail = FALSE)
+  se <- beta / z
+  betavar <- se^2
+  betavar
+}
+calc_betavar_from_beta_pval <- Vectorize(calc_betavar_from_beta_pval)
+
+make_coloc_dataset <- function(l) {
+  ds <- c(l$dataset, l$metadata)
+}
+
+braak_bin1 <- readRDS("output/coloc/mega/braak/GWAS/gwas_chr2_127133851_A_C_b38.Rds")
+gtex_bin1 <- readRDS("output/coloc/mega/braak/GTEx/Brain_Cerebellum_ENSG00000136717.14_chr2_127133851_A_C_b38.Rds")
+braak_results <- fread("output/gwas/mega/polmm/braak_polmm_results.txt")
+braak_results <- braak_results[chr == 2][SNPID %in% braak_bin1$dataset$rsID]
+braak_bin1$dataset <- merge(braak_bin1$dataset,
+                            braak_results[, .(SNPID, beta)],
+                            by.x = "rsID",
+                            by.y = "SNPID")
+gtex_bin1$dataset <- gtex_bin1$dataset[snp %in% braak_bin1$dataset$chr_bp_ref_alt_b38]
+braak_bin1$dataset <- braak_bin1$dataset[chr_bp_ref_alt_b38 %in% gtex_bin1$dataset$snp]
+snps <- braak_bin1$dataset$rsID
+
+ld_mat <- ld_matrix_local(snps, 
+                with_alleles = FALSE, 
+                bfile = "/home/lmsh224/1000g_plink/EUR",
+                plink_bin = genetics.binaRies::get_plink_binary())
+keep_snps <- colnames(ld_mat)
+ld_mat_gwas <- ld_matrix_local(keep_snps, 
+                          with_alleles = FALSE, 
+                          bfile = "data/mega/mega_np",
+                          plink_bin = genetics.binaRies::get_plink_binary())
+
+braak_bin1$dataset <- braak_bin1$dataset[rsID %in% keep_snps]
+gtex_bin1$dataset <- gtex_bin1$dataset[snp %in% braak_bin1$dataset$chr_bp_ref_alt_b38]
+
[…]

### 2023-02-02 code/coloc/coloc_caa_ord_apoe.py

@@ -1,0 +1,27 @@
+# run in coloc directory
+import subprocess
+import os
+import pandas as pd
+
+pheno_file = "../np_phewas/data/mega/mega_np_ord_dichotomized.pheno"
+pheno_data = pd.read_csv(pheno_file)
+phenos = ["caa_ord"]
+
+for pheno in phenos:
+    cmd = [
+        "python", "./coloc_qtl.py",
+        "--out_pref", "../np_phewas/output/coloc/mega/" + pheno + "_apoe/",
+        "--gwas_sumstats",
+        "../np_phewas/output/gwas/mega/polmm/caa_ord_apoe_polmm_results.txt",
+        "--max_pval", "1e-5",
+        "--pval_col", "pval.spa",
+        "--phenotype_type", "cc",
+        "--phenotype_col", pheno,
+        "--phenotype_file", pheno_file,
+        "--phenotype_na_string", "NA",
+        "--gtex",
+        "--rosmap"
+    ]
+
+    subprocess.run(cmd)
+

### 2023-02-02 code/coloc/coloc_ordinal_phenos.py

@@ -1,0 +1,26 @@
+# run in coloc directory
+import subprocess
+import os
+import pandas as pd
+
+pheno_file = "../np_phewas/data/mega/mega_np_ord_dichotomized.pheno"
+pheno_data = pd.read_csv(pheno_file)
+phenos = pheno_data.columns[2:]
+
+for pheno in phenos:
+    cmd = [
+        "python", "./src/coloc_qtl.py",
+        "--out_pref", "../np_phewas/output/coloc/mega/" + pheno + "/",
+        "--gwas_sumstats", 
+        "../np_phewas/output/gwas/mega/polmm/" + pheno + "_polmm_results.txt",
+        "--max_pval", "1e-5",
+        "--pval_col", "pval.spa",
+        "--phenotype_type", "cc",
+        "--phenotype_col", pheno,
+        "--phenotype_file", pheno_file,
+        "--phenotype_na_string", "NA", 
+        "--gtex", 
+        "--rosmap"
+    ]
+    
+    subprocess.Popen(cmd)

### 2023-02-02 code/coloc/coloc_results.R

@@ -1,0 +1,36 @@
+library(data.table)
+library(rtracklayer)
+library(stringi)
+# colocalization results
+pheno_ord <- fread("data/mega/mega_np_ord.pheno")
+pheno_bin <- fread("data/mega/mega_np.pheno")
+pheno <- merge(pheno_ord, pheno_bin, c("FID", "IID"))
+phenos <- colnames(pheno)[3:ncol(pheno)]
+coloc_list <- vector(mode = "list", length = length(phenos))
+names(coloc_list) <- phenos
+for (i in 1:length(phenos)) {
+  p <- phenos[i]
+  coloc_path <- paste0("output/coloc/mega/", p, "/coloc_results.txt")
+  coloc_list[[i]] <- fread(coloc_path)
+}
+
+coloc <- rbindlist(coloc_list)
+coloc[, gene_id := stri_replace_all_regex(QTL_Phenotype, '.*:', '')]
+coloc[, gene_id := stri_replace_all_regex(gene_id, '\\..*', '')]
+
+# gene names
+genes <- rtracklayer::import('/home/lmsh224/downloads/gencode.v40.annotation.gtf.gz')
+genes <- as.data.table(genes)
+genes[, gene_id_trunc := stri_replace_all_regex(gene_id, '\\..*', '')]
+
+coloc_genes <- genes[gene_id_trunc %in% coloc$gene_id
+][, .(gene_id_trunc, gene_name, gene_type)]
+coloc_genes <- coloc_genes[!duplicated(coloc_genes)]
+setnames(coloc_genes, "gene_id_trunc", "gene_id")
+
+coloc <- merge(coloc, coloc_genes, by = 'gene_id', all.x = TRUE)
+coloc[is.na(gene_name), gene_name := QTL_Phenotype]
+setnames(coloc, "gene_name", "QTL_Target")
+
+fwrite(coloc, "output/coloc/mega/coloc_results.txt")
+fwrite(coloc[PPH4 > 0.8], "output/coloc/mega/coloc_results_pph4_80.csv")

### 2023-02-02 code/manuscript/table_s3_coloc_results.R

@@ -1,0 +1,42 @@
+
+library(data.table)
+library(flextable)
+library(magrittr)
+library(stringi)
+coloc <- fread("output/coloc/mega/coloc_results.txt")
+coloc <- coloc[!duplicated(coloc)]
+
+coloc_ts3_dt <- coloc[
+  PPH4 >= 0.8,
+  .(GWAS_Phenotype, Chromosome, QTL_Target, Tissue, QTL_Type, QTL_Source,
+    P_GWAS, P_QTL, PPH4)]
+coloc_ts3_dt[, Chromosome := 
+               as.integer(stri_replace_first_fixed(Chromosome, "chr", ""))]
+setorder(coloc_ts3_dt, GWAS_Phenotype, Chromosome, QTL_Target, Tissue)
+coloc_ts3_dt[, Tissue := stri_replace_all_fixed(Tissue, "_", " ")]
+coloc_ts3_dt[, P_GWAS := formatC(P_GWAS, digits = 1, format = "e")]
+coloc_ts3_dt[, P_QTL := formatC(P_QTL, digits = 1, format = "e")]
+coloc_ts3_dt[, PPH4 := paste0(round(PPH4 * 100, 1), "%")]
+coloc_ts3_dt[PPH4 == "100%", PPH4 := ">99.9%"]
+coloc_ts3_dt[, GWAS_Phenotype := 
+               factor(
+                 GWAS_Phenotype, 
+                 labels = c("Arteriolosclerosis", "Atherosclerosis",
+                            "Braak NFT Stage", "CAA", "CERAD Score", 
+                            "Diffuse Amyloid Plaques", "Gross Infarction",
+                            "HS", "LATE-NC", "Lewy Body", "Microinfarct"))]
+setnames(coloc_ts3_dt, colnames(coloc_ts3_dt), 
+         c("NPE", "Chromosome", "QTL Target", "Tissue", "QTL Type", 
+           "QTL Source", "P NPE", "P QTL", "PrC"))
+ts3 <- flextable(coloc_ts3_dt) %>% 
+  autofit() %>% 
+  padding(padding = 2, part = "all") %>%
+  add_header_lines(paste0("Table S3: NPE-QTL Colocalizing Loci")) %>% 
+  footnote(i = nrow_part(., part = "header"), j = 2, 
+           value = 
+             as_paragraph(paste0("Closest protein-coding gene according to ",
+                                 "GENCODE release 40.")), 
+           ref_symbols = c("a"),
+           part = "header")
+ ts3
+ 

### 2023-02-02 code/prep/mcbb/rnaseq_tidy.R

@@ -1,0 +1,23 @@
+library(data.table)
+library(magrittr)
+library(stringi)
+rnaseq_t <- fread("data/mcbb/MayoRNAseq_RNAseq_TCX_geneCounts_normalized.tsv")
+rnaseq <- t(rnaseq_t)
+make_tidy <- function(m) {
+  gene_id <- as.vector(m[1, ])
+  specimenID <- rownames(m)[-1]
+  m <- m[-1, ] %>% 
+    apply(2, as.numeric) %>% 
+    as.data.table()
+  m[, specimenID := specimenID]
+  setcolorder(m, "specimenID")
+  
+  colnames(m)[2:ncol(m)] <- gene_id
+  m
+}
+
+rnaseq_tidy <- make_tidy(rnaseq)
+rnaseq_tidy[, specimenID := stri_replace_last_fixed(specimenID, "_TCX", "")]
+
+fwrite(rnaseq_tidy, file = "data/mcbb/rnaseq_gene_counts_normalized_tidy.csv",
+       quote = FALSE)

### 2023-02-02 code/prep/mega/02_make_pheno.R

@@ -9,7 +9,7 @@ adni <- fread("data/adni/adni_np.pheno", na.strings = '-1')
 fam <- fread("data/mega/mega_np.fam")
 
 act_np <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx",
-                               sheet = 2))
+                                  sheet = 2))
 adc_np <- setDT(readRDS("data/adc/np_qced.Rds")) # %>%
   # replace_with_na_all(condition = ~.x == -1)
 setDT(adc_np)

### 2024-04-30 manuscript.Rmd
Organize code before publication.

--- 2022-08-16 manuscript.Rmd
@@
 
-# Introduction
+# Title page
 
-Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinically diagnosed AD; however, clinically diagnosed AD is increasingly recognized as a heterogeneous disease that can reflect multiple independent or co-occurring pathological processes.
-Aged individuals in research autopsy cohorts often have more than one proteinopathy present in their brains at time of death.[@karanth2020].
-Vascular factors, especially small-vessel disease, also contributes to cognitive decline.
+**Title**
 
-does not always correspond with AD neuropathology and vice-versa.[@nelson2012]
+Genome-wide association study of multiple neuropathology endophenotypes identifies novel risk loci and provides insights into genetic risk of dementia
 
-We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:
+**Authors**
 
-1.  identify novel genetic risk loci for individual NPE
-2.  identify common sources of genetic risk (pleiotropy) between NPE
-3.  provide evidence that the effects of clinical AD genetic risk loci may be mediated by specific NPE.
+Lincoln M.P.
+Shade (1), Yuriko Katsumata (1, 2), Erin L.
+Abner (2, 3), Steven A.
+Claas (1), Timothy J.
+Hohman (4), Shubhabrata Mukherjee (5), Margaret A.
+Pericak-Vance (6), Richard P.
+Mayeux (7), Lindsay A.
+Farrer (8), Gerard D.
+Schellenberg (9), Jonathan L.
+Haines (10), Walter A.
+Kukull (11), Kwagsik Nho (12), Andrew J.
+Saykin (12), David A.
+Bennett (13), Julie A.
+Schneider (14), Mark T.
+W.
+Ebbert (2, 15, 16), Peter T.
+Nelson (16, 17), David W.
+Fardo (1, 2), the National Alzheimer's Coordinating Center, and the Alzheimer's Disease Genetics Consortium
+
+**Add Qi and Khine as coauthors**
+
+**Author Affiliations**
+
+(1) Department of Biostatistics, College of Public Health, University of Kentucky, Lexington, KY, USA;
+
+(2) Sanders-Brown Center on Aging and Alzheimer's Disease Research Center, University of Kentucky, Lexington, KY, USA;
+
+(3) Department of Epidemiology and Environmental Health, College of Public Health, University of Kentucky, Lexington, KY, USA;
+
+(4) Vanderbilt Memory & Alzheimer's Center, Department of Neurology, Vanderbilt University Medical Center, Nashville, TN, USA;
+
+(5) Department of Medicine, University of Washington, Seattle, WA, USA;
+
+(6) John P. Hussman Institute for Human Genomics, University of Miami Miller School of Medicine, Miami, FL, USA;
+
+(7) Department of Neurology, Columbia University, New York, NY, USA;
+
+(8) Department of Medicine, Boston University, Boston, MA, USA;
+
+(9) Penn Neurodegeneration Genomics Center, University of Pennsylvania Perelman School of Medicine, Philadelphia, PA, USA;
+
+(10) Cleveland Institute for Computational Biology, Case Western Reserve University, Cleveland, OH, USA;
+
+(11) Department of Epidemiology, University of Washington, Seattle, WA, USA;
+
+(12) Department of Radiology & Imaging Sciences, Indiana University School of Medicine, Indianapolis, IN, USA;
+
+(13) Rush Alzheimer's Disease Center and Department of Neurological Sciences, Rush Medical College, Chicago, IL, USA;
+
+(14) Rush Alzheimer's Disease Center and Departments of Neurology and Pathology, Rush University Medical Center, Chicago, IL, USA;
+
+(15) Division of Biomedical Informatics, Department of Internal Medicine, University of Kentucky, Lexington, KY, USA;
+
+(16) Department of Neuroscience, University of Kentucky College of Medicine, Lexington, KY, USA;
+
+(17) Department of Pathology and Laboratory Medicine, University of Kentucky, Lexington, KY, USA
+
+**Corresponding Author**
+
+David W. Fardo
+
+Healthy Kentucky Research Building
+
+760 Press Ave, Room 372
+
+Lexington, KY 40536-0679
+
+[david.fardo\@uky.edu](mailto:david.fardo@uky.edu)
+
+859-218-2070
+
+# Key words
+
+GWAS, Alzheimer's disease, neuropathology, atherosclerosis, neurofibrillary tangles, cerebral amyloid angiopathy, TDP-43
+
+# Main
+
+Late-onset amnestic dementia diagnosed as Alzheimer's Disease (AD/LOAD) is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors.[@farfel2016; @katsumata2022; @karanth2020] Because of the complex nature of amnestic dementia and the practical challenge of constructing large cohorts with harmonized pathology data, LOAD genome-wide association studies (GWAS) to date have primarily utilized clinical diagnosis or proxy phenotypes based on family history of dementia to help unravel the genetic etiology of disease.[@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019] While these LOAD GWAS have been immensely successful at identifying dementia-associated genetic loci, with over 80 such loci identified thus far,[@bellenguez2022] the clinical phenotypes used in most studies complicate the interpretation of their results and belies a more complex reality of mixed neuropathologies being the norm in aged individuals.[@karanth2020; @escott-price2022] To build on the successes of previous LOAD GWAS, and to begin decrypting the complexities of this heterogeneous clinical syndrome, we believe GWAS using neuropathology endophenotypes (NPE) is an essential next step--not only to broadly understand LOAD's genetic etiology, but to identify alleles that are drive specific LOAD-associated pathologies.[@karran2022]
+
+Neuritic plaques and neurofibrillary tangles (NFT) are present in the brains of the majority of people diagnosed with clinical LOAD, but \~20% of diagnosed do not have substantial AD neuropathologic change (ADNC), and another \~50% have both ADNC and comorbid non-AD pathologies.[@nelson2009; @karanth2020] For example, TAR DNA-binding protein 43-kDa (TDP-43) pathology is found in 40-50% of autopsied individuals with dementia and contributes to an amnestic syndrome, limbic-predominant age-related TDP-43 encephalopathy (LATE), with a characteristic pattern of neuropathological change called LATE-NC.[@karanth2020; @nelson2019; @nelson2022] Hippocampal sclerosis (HS) of aging, a pathology characterized by neuronal death, gliosis, and atrophy of the hippocampus beyond normal ranges based on levels of AD pathology, commonly co-occurs with LATE-NC and is associated with severe cognitive impairment.[@brenowitz2014; @nelson2019] It is already known that the genetic risk factors of ADNC and LATE-NC differ,[@dugan2021] but the genetic underpinnings of common comorbid (ADNC+LATE-NC) phenotypes are not well understood.
+
+Cerebrovascular pathologies, which contribute to cognitive decline and dementia, are also common in elderly autopsied research participants.[@skrobot2016] Cerebral amyloid angiopathy (CAA) is a cerebrovascular pathology characterized by amyloid-beta deposition in cerebral blood vessels.[@weber2018] CAA often co-occurs with LOAD but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma.[@weber2018; @skrobot2016] Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline.[@smith2012; @skrobot2016] Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and HS [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies.[@arvanitakis2016; @ighodaro2017] Collectively, these factors reveal an increasingly complex and synergistic web of pathologies which contribute to cognitive impairment and dementia and further demonstrate the need to understand how existing (and potentially new) GWAS loci are involved in LOAD-associated pathologies.
+
+Studying the genomic risk factors of individual underlying neuropathologies can provide a complementary approach to large GWAS of clinical- and family history-based outcomes for the study of AD and dementia risk.
+As proof of principle, prior GWAS of neuropathology endophenotypes (NPE) have confirmed known LOAD risk loci and have identified potential new neuropathology risk loci.[@beecham2014; @shade2022; @vattathil2021; @farrell2022; @reddy2021] Some NPE, particularly LATE-NC, have yet to be studied using GWAS; statistical methodological development for GWAS and downstream functional analyses along with continued recruitment in autopsy cohorts provide additional opportunity for in-depth investigation into multiple neuropathologies.
+Here, we perform GWAS on eleven neuropathology endophenotypes in the largest harmonized cohort, to date, using three high-quality data sources with autopsy and genotype data.
+For previously studied NPE, the sample sizes used here represent 50-150% increases relative to those used in prior studies.
+We also perform downstream functional analyses to explore potential biological functional mechanisms of newly identified risk loci and provide insight into known AD risk loci.
+
+# Results
+
+## Participants and distributions of neuropathology endophenotypes
+
+```{r}
+add_comma <- function(n) {
+  formatC(n, big.mark = ",")
+}
+nacc_covar <- fread("data/adc/adc_np.covar")
+rosmap_covar = fread("data/rosmap/rosmap_np.covar")
+act_covar = fread("data/act/act_np.covar")
+covar_vars = c("FID", "IID", "msex", "age_death", paste0("pc", 1:10))
+covar = rbindlist(
+  list(
+    nacc_covar[, ..covar_vars], 
+    rosmap_covar[, ..covar_vars], 
+    act_covar[, ..covar_vars]), 
+  use.names = TRUE
+)
+np_pheno <- fread("shared/nacc_rosmap_act_np.pheno")
+nmega <- formatC(np_pheno[, .N], big.mark = ",")
+nnacc <- formatC(np_pheno[FID == "0", .N], big.mark = ",")
+nrosmap <- formatC(np_pheno[FID == "1", .N], big.mark = ",")
+nact <- formatC(np_pheno[FID == "2", .N], big.mark = ",")
+# nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")
+
+# np_pheno_ord <- fread("data/mega/mega_np_ord.pheno", na.strings = "-1")
+# np_pheno_bin <- fread("data/mega/mega_np.pheno", na.strings = "-1")
+
+# rm(np_pheno_ord, np_pheno_bin)
+ncerad <- np_pheno[!is.na(cerad), add_comma(.N)]
+nbraak <- np_pheno[!is.na(braak), add_comma(.N)]
+# ndiffuse_abeta <- np_pheno[!is.na(diffuse_abeta), add_comma(.N)]
+narteriol <- np_pheno[!is.na(arteriol), add_comma(.N)]
+natheroscler <- np_pheno[!is.na(atheroscler), add_comma(.N)]
+ncaa <- np_pheno[!is.na(caa), add_comma(.N)]
+nlate <- np_pheno[!is.na(late), add_comma(.N)]
+nlewy = np_pheno[!is.na(lewy_body), add_comma(.N)]
+nhs = np_pheno[!is.na(hs), add_comma(.N)]
+nmicroinf = np_pheno[!is.na(microinf), add_comma(.N)]
+ngrossinf = np_pheno[!is.na(grossinf), add_comma(.N)]
+```
+
+We used genotype and neuropathology data from three autopsy data sources: (1) the National Alzheimer's Coordinating Center (NACC, N = `r nnacc`), (2) the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP, N = `r nrosmap`), and (3) the Adult Changes in Thought (ACT, N = `r nact`) study.
+In total, `r nmega` unique participants were included in our analyses.
+NACC participants were more likely to be diagnosed with either dementia (83%) or AD/mild cognitive impairment (MCI, 77.5%) than ROSMAP (55% and 66%) or ACT (13% and 10%) participants.
+NACC participants died at a younger age (mean age at death 81 years) compared to ROSMAP (90 years) and ACT (89 years) participants and had more balanced participation between the sexes, with 50% of NACC participants being female vs.
+68% and 58% in ROSMAP and ACT, respectively.
+NACC participants were also more likely to have an *APOE* $\epsilon 4$ allele (54%) vs.
+ROSMAP (25%) or ACT (28%; Chi-square test $P \text{-value}<0.001$ for all comparisons).
+These differences reflect heterogeneity in the recruitment for these studies.
+NACC is based on data collected from over 30 Alzheimer's Disease Research Centers which often recruit from clinic patients and their families.
+In contrast, ACT originally recruited aged participants in the Seattle, WA area who did not have dementia at time of enrollment, and ROS recruited members of the Catholic Church clergy.
+
+Available sample sizes differed among the eleven studied NPE, which included the AD-related pathologies of Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques (\$N = `r ncerad`, Braak staging of NFT (N = `r nbraak`), and diffuse amyloid-beta plaques; the non-AD proteinopathies LATE-NC (N = `r nlate`) and Lewy body pathology (N = `nlewy`); hippocampal sclerosis (N = `r nhs`); and the cerebrovascular pathologies CAA (N = `r ncaa`), gross infarcts (N = `r ngrossinf`), microinfarcts (N = `r nmicroinf`), atherosclerosis (N = `r natheroscler`), and arteriolosclerosis (N = `r narteriol`).
+Relative to the Beecham et al.
+(2014) GWAS of neuropathology endophenotypes, our updated sample represents an overall 52% increase in sample size.[@beecham2014] The sample size increase for individual neuropathologies is greater, ranging from 57% for Braak stage to 150% for CAA, demonstrating both an increased total number of autopsy participants and more complete neuropathological profiling for each participant.
+HS, microinfarcts, and gross infarcts were coded as binary case/control outcomes.
+All other phenotypes were ordinal variables with either four or seven levels (Methods and Supplementary Note describe our phenotype definitions and harmonization approach).
+
+To explore the co-occurrence of neuropathology endophenotypes in our pooled data, we estimated polychoric correlations (an approach which assumes that observed ordinal or binary variables reflect latent normally distributed variables) between neuropathology endophenotype pairs followed by hierarchical clustering using the polycor, psych, and pheatmap R packages.[@revelle2022; @fox2022; @kolde2019] Inspection of the resulting dendrogram identified three positively correlated clusters of endophenotypes: a "vascular" cluster consisting of gross infarcts, microinfarcts, arteriolosclerosis, and atherosclerosis; an "Alzheimer's disease" cluster consisting of Braak NFT stage, neuritic plaques, diffuse amyloid-beta plaques, and CAA; and a "LATE" cluster consisting of LATE-NC and HS (Supplementary Figure 1).
+
+## GWAS meta-analysis of eleven neuropathology endophenotypes (NPE)
+
+An overview of our study design is shown in **Figure 1**.
+Genetic association analyses were performed with logistic or proportional-odds logistic regression mixed-effects models as appropriate (Methods).
+We performed GWAS on 11 NPE using NACC participants, ROSMAP, and ACT separately.
+We then performed fixed-effect meta-analysis using METAL on variants with minor allele frequencies $\ge 1\%$ in each data source.[@metal:f]
+
+```{r}
+t2_env <- new.env()
+load("output/manuscript/table2_meta_analysis_results.Rdata",
+       envir = t2_env)
+```
+
+In total, 13 NPE-genomic locus associations had lead variants that met the genome-wide significance threshold of $P \text{-value}< 5 \times 10^{-8}$.
+Three NPE (Braak stage, CERAD score, and CAA) were associated with two or more independent loci ($r^2\le0.05$) in the *APOE* region, though we report only one locus in the *APOE* region for each phenotype unless secondary loci remained significant after conditioning on *APOE* $\epsilon$ diplotype (see conditional *APOE*-adjusted analysis below).
+The lead variant rs429358 in the *APOE* region was significantly associated with four NPE: Braak stage ($\text{OR} = 2.06$, $P \text{-value}= 9.7\times 10^{-89}$), CAA ($\text{OR} = 2.49$, $P \text{-value}= 4.4\times 10^{-138}$), CERAD score ($\text{OR} = 2.42$, $P \text{-value}= 4.7\times 10^{-103}$), and LATE-NC ( $\text{OR} = 1.70$, $P \text{-value}= 1.7\times 10^{-14}$).
+Variants within *TMEM106B* were associated with both HS (rs7805419; $\text{OR} = 0.65$, $P \text{-value}= 3.2\times 10^{-13}$) and LATE-NC (2043539; $\text{OR} = 0.70$, $P \text{-value}= 5.8\times 10^{-11}$).
+A *GRN* locus was associated with HS (rs5848; $\text{OR} = 1.40$, $P \text{-value}= 3.2\times 10^{-8}$).
+One *BIN1* locus was associated with Braak stage (rs6733839 $\text{OR} = 1.21$, $P \text{-value}= 1.6\times 10^{-9}$).
+
+Novel associations identified in the NPE GWAS meta-analysis included a variant 12 kbp upstream of *COL4A1* associated with cerebral atherosclerosis (rs2000660; $\text{OR} = 0.73$, $P \text{-value}= 2.7\times 10^{-8}$), a *PIK3R5* intronic locus associated with Braak stage (rs72844606 $\text{OR} = 0.69$, $P \text{-value}= 4.0\times 10^{-8}$), and an intronic *LZTS1* locus associated with arteriolosclerosis (rs78909048; $\text{OR} = 0.44$, $P \text{-value}= 5.7\times 10^{-10}$).
+One intronic locus of *SPATA48* was close to genome-wide signicantly associated with atherosclerosis (rs62447817; $\text{OR} = 1.35$, $P \text{-value}= 5.5\times 10^{-8}$)
+
+## Conditional *APOE*-adjusted analysis
+
+To check whether any loci in the *APOE* region (defined as less than 500 kbp from the start or end site of *APOE* transcription) were associated with NPE risk independently of the known effects of *APOE* $\epsilon 2/\epsilon 3/\epsilon 4$ alleles, we re-analyzed the *APOE* region of Chromosome 19 while adjusting for *APOE* $\epsilon$ diplotypes.
+We limited this conditional analysis to NPE with which variants within the *APOE* region met the genome-wide significance threshold of $P \text{-value}< 5\times 10^{-8}$ in Stage 3 (Braak NFT stage, neuritic plaque CERAD score, diffuse amyloid-beta plaques, cerebral amyloid angiopathy, and LATE-NC).
+One locus between *APOC4-APOC2* with lead variant rs7247551 remained significantly associated with CAA after adjusting for *APOE* $\epsilon 2/\epsilon 3/\epsilon 4$ diplotypes ($\text{OR} = 0.81$, $P \text{-value}= 8.0\times 10^{-12}$, Table 1 and Figure 3a).
+No variants remained genome-wide significantly associated with any other APOE-associated NPE, and the P-values of the lead variant rs429358 from genome-wide meta-analysis unadjusted for APOE were all above 0.05 in the re-analysis.
+Using independent data from a recent GWAS of CAA in 815 participants with dementia in the Mayo Clinic Brain Bank, we replicated the association of rs7247551 with CAA while adjusting for APOE $\epsilon$ diplotype ($\beta = -0.068$, $P \text{-value} = 0.0012$).
+These results provide evidence for a novel locus in the *APOE* region other than the known *APOE* $\epsilon$ haplotypes that affects CAA pathology burden, and that the genetic risk basis of CAA in the *APOE* region may differ from that AD neuropathologies of neuritic plaques and NFT.
+
+```{r}
+t2_env$t2
+```
+
+## Associations of clinical and proxy AD risk loci with NPE
+
+We checked whether AD-associated loci identified in the recent Bellenguez et al.
+(2022) GWAS of AD and related dementia (ADD) were associated with NPE.[@bellenguez2022] Bellenguez et al.
+identified a total of 83 non-*APOE* independent loci (39 previously identified, 44 novel) associated with ADD (hereafter, "ADD loci"), 80 of which had lead variants that were included in our study.
+Each of the 11 NPE used as outcomes in the present study was checked for association with variants in these 80 ADD loci.
+Benjamini-Hochberg adjustments were then applied to *P*-values for each NPE separately to control the false-discovery rate (FDR).
+In total, 22 NPE-locus associations had adjusted *P*-values (*Q*-values) $\le 0.05$ and were considered significant; of these 22 associations, 21 had concordant directions of effect with Bellenguez et al.
+(Table 2 and Supplementary Table 5).
+The pathognomonic AD pathologies, Braak stage and CERAD, had 70 and 60 of 80 concordant directions of effect with ADD loci, respectively, regardless of statistical significance.
+Three ADD loci (*BIN1, MME,* and *HLA*) were significantly associated with Braak NFT stage after multiple test correction.
+Ten AD loci (*CR1*, *BIN1, INPP5D,* ZCWPW1/NYAP1*, PTK2B, PICALM*, *SORL1*, *FERMT2*, *SNX1,* andABCA7) were significantly associated with CERAD score after multiple test correction.
+Four AD loci (*TMEM106B, IL34, GRN*, and *MAPT*) were significantly associated with HS, all of which except for *L34* (HS $\text{OR} = 0.74$, $P \text{-value}=0.0019$; AD $\text{OR} = 1.06$, $P \text{-value}=5.6\times 10^{-6}$) were concordant in effect direction.
+Two ADD loci (*TMEM106B* and *GRN*) were significantly associated with LATE-NC, both concordant.
+Finally, one ADD locus (*PLCG2*) was significantly associated with microinfarcts.
+These results indicate that NPE studies largely corroborate the findings of large AD GWAS based on clinical and proxy phenotypes.
+However, several ADD loci, particularly *TMEM106B* and *GRN*, are associated with non-AD pathology (HS and LATE-NC) but not neuritic plaques or NFT.
+
+```{r}
+load("output/manuscript/results_ad_snps_table.Rdata")
+results_ad_snps_table
+```
+
+## Gene-prioritization and enrichment analyses
+
+We used gene-based, pathway, and enrichment analyses implemented in FUMA to identify potential functional genes and biological pathways involved in neuropathology development (results available in Supplementary Data zip file).[@watanabe2017] We used a minimum threshold of association of $P \text{-value}\le 1 \times 10^{-5}$ in the Stage 3 analysis for variant annotation.
+Gene-level analyses implemented in MAGMA corroborated our single-variant analyses,[@leeuw2015] identifying gene associations with neuropathology endophenotypes in regions with significant variant-phenotype associations.
+*APOE* was significantly associated with NFT, diffuse plaques, CAA, neuritic plaques, and LATE-NC.
+*TMEM106B* was associated with both HS and LATE-NC.
+Gene-set enrichment analyses were performed using MAGMA and Gene Ontology (GO) and curated gene sets from MSigDB.[@liberzon2011] After Bonferroni correction genes involved in polysaccharide digestion were significantly associated with microinfarcts ($P \text{-value}_{\text{bon}} = 5.3\times10^{-5}$).
+We also found evidence that genes involved in substrate recognition for ubiquination were associated with cerebral atherosclerosis ($P \text{-value}_{\text{bon}} = 0.0054$), and genes overexpressed in developing B cells were associated with Braak NFT stage ($P \text{-value}_{\text{bon}} = 0.024$).
+FUMA variant annotation using CADD score, RegulomeDB score, eQTL, and chromatin interaction annotations indicated that the rs2000660 lead variant associated with cerebral atherosclerosis in Stage 3 is located within an enhancer region located 13 Kbp upstream of *COL4A1* transcription start site.
+
+## Colocalization analysis
+
+We investigated whether loci associated with multiple NPE colocalized using a Bayesian colocalization analysis approach implemented in the coloc R package.[@giambartolomei2014] Because CAA was the only NPE with an associated locus in the *APOE* region with a signal independent of *APOE* $\epsilon$ diplotype, we excluded this region from NPE-NPE colocalization analysis.
+Pairs of NPE with the same variant reaching a P-value threshold of $P \text{-value}< 1 \times 10^{-5}$ with concordant effect directions included (1) Braak NFT stage and CERAD score in the *BIN1* region on Chromosome 2 and (2) HS and LATE-NC in the *TMEM106B* region on Chromosome 7 (**Figure 4**).
+Both loci exhibited high posterior probability of colocalization ($\text{PrC}> 90\%$), indicating that the same genetic variants drive association with both endophenotypes.
+
+A locus within *GRN* associated with HS ($P \text{-value}= 3.2 \times 10^{-8}$) strongly colocalized with *GRN* expression in multiple tissues ($\text{PrC}>99.9\%$).
+The *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* expression in multiple tissues, including the cerebellar hemisphere ($\text{PrC}=90\%$).
+Two CpG sites located either within *TMEM106B* (cg09613507) or downstream (cg23422036) colocalized with both HS and late (cg09613507-HS $\text{PrC}=94\%$, cg09613507-LATE-NC $\text{PrC}=89\%$, cg23422036-HS $\text{PrC}=94\%$, cg23422036-LATE-NC $\text{PrC}=95\%$).
+
+The Chromosome 19 CAA-associated locus identified in the *APOE*-adjusted analysis colocalized with expression of multiple genes in GTEx: *APOE* expression in the wall of the aorta; *8CLPTM1* expression in the skin of the leg and suprapubic region; and *APOC2* expression in 14 different tissues, including the brain cortex, caudate, nucleus accumbens, and cerebellum ($\text{PrC}= 90\text{-}97\%$).
+The same locus colocalized with mQTL for four CpG sites in ROSMAP, cg09555818, cg04401876, cg10169327, and cg13119609 ($\text{PrC}= 96\text{-}99\%$, see **Figure 3b**).
+
+Multiple suggestive NPE loci showed evidence of colocalization with eQTL in GTEx.
+In total, 44 NPE loci (lead variant $P \text{-value} < 1 \times 10^{-5}$) colocalized with QTL with $\text{PrC}= \ge80\%$, with 533 total colocalizing NPE-QTL pairs across 44 tissues (Supplementary Table S4).
+
+## Multiple eQTL and mQTL colocalizing with NPE are validated with methylation and RNA-seq in ROSMAP
+
+Using ROSMAP participants with available DNA methylation and/or RNA-Seq data available from the DLPFC, we further investigated associations between QTL phenotypes found to colocalize with NPE in brain tissues of GTEx and/or ROSMAP.
+We tested methylation levels at each of the four CpG sites -- cg09555818, cg04401876, cg10169327, and cg13119609 -- that colocalized with the Chromosome 19 CAA risk locus for association with to confirm association with lead variant rs7247551.
+We first confirmed that all four CpG sites were significantly associated ($P \text{-value} < 0.0001$) with rs7247551 with directions of effect consistent with those previously reported for ROSMAP.[@ng2017] We then tested for association between methylation at these sites with CAA pathology.
+Hypomethylation at cg09555818 (odds ratio $[\text{OR}] =0.82$, $P \text{-value}= 0.004$) and cg13119609 ($\text{OR} =0.78$, $P \text{-value}= 0.0007$) were significantly associated with worse CAA pathology (Figure 3C).
+
+Next, as *APOC2* expression in multiple brain tissues colocalized with CAA in GTEx but not ROSMAP, we investigated to see whether there was a nominal association between *APOC2* expression in the DLPFC and rs7247551.
+We found that the G allele of rs7247551 was significantly associated with increased *APOC2* expression in the DLPFC ($\beta = 0.072$, $P \text{-value}= 0.00015$, Figure 3D); however, the direction of effect was opposite of that found in brain tissues in GTEx (i.e., the G allele of rs7247551 was associated with decreased *APOC2* expression in GTEx, Figure 3E).
+Expression of *APOC2* in the DLPFC was not associated with CAA in ROSMAP ($\text{OR} =0.98$, $P \text{-value}= 0.89$).
+We performed an additional *post-hoc* analysis for nominal *APOE* eQTL activity of rs7247551 in ROSMAP (rs7247551 was lead eQTL in colocalizing *APOE* expression loci in the aorta, thyroid, and prostate, though APOE expression in no brain tissues colocalized with CAA).
+We confirmed that rs7247551 was not associated with *APOE* expression in the DLPFC in ROSMAP ($P \text{-value}= 0.75$, Figure 3D).
+
+We then investigated the associations betwen HS and LATE-NC with the *GRN* and *TMEM106B* expression and methylation in the DLPFC.
+We first confirmed that the lead *TMEM106B* variants for HS (rs7805419) and LATE-NC (rs2043539) were in LD ($R^2 = 0.96$) and that rs2043539 was associated with DNA methylation levels at cg09613507 and cg23422036 as previously reported.
+We also found that Neither TMEM106B nor GR*N* expression were associated with HS ($P \text{-values} > 0.05$), while decreased *TMEM106B* expression was associated with more severe LATE-NC pathology ($P \text{-value}= 0.035$).
+Of the two CpG sites that colocalized with HS and LATE-NC, hypermethylation of cg09613507 was associated with more severe LATE-NC pathology ($P \text{-value}= 0.0093$).
+
+# Discussion
+
+The present study of eleven neuropathology endophenotypes with a maximum sample size of 7,786 participants serves as an autopsy-based complement to previous studies, expanding on the findings of previous genetic association studies of NPE, Alzheimer's, and related dementias to attempt to better understand the complex associations between different neuropathologies and genetic risk factors.[@katsumata2022; @shade2022; @beecham2014; @dugan2021; @bellenguez2022; @vattathil2021; @reddy2021] We identified a novel *APOE* $\epsilon$-independent CAA risk locus that also affects *APOC2* brain expression and dementia-associated CpG methylation sites.
+Two of these CpG sites were in turn also associated with CAA risk.
+We also discovered intriguing new loci mapped to *COL4A1*, *PIK3R5*, and *LZTS1* associated with atherosclerosis in the circle of Willis, Braak stage for neurofibrillary tangles, and brain arteriolosclerosis, respectively.
+Lastly, we investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
+
+Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
+Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE.[@beecham2014; @dugan2021] Notably, *APOE* variants were not significantly associated with vascular pathology except for CAA.
+When we re-ran Chromosome 19 analyses on these phenotypes while adjusting for *APOE* $\epsilon$ genotype, only one CAA-associated locus with lead variant rs7247551 remained significant ($\text{OR} = 0.81$, $P \text{-value}= 8.0 \times 10^{-12}$).
+We confirmed that this locus was indeed novel and not in LD with a recent study of CAA in *APOE* $\epsilon 4$ carriers by Reddy et al.
+(2021) by confirming that $R^2 <0.01$ between rs7247551 and Reddy et al.
+lead variant rs5117.[@reddy2021] Futhermore, we were able to replicate the associated between rs7247551 and CAA pathology while adjusting for *APOE* $\epsilon$ diplotype in 815 Mayo Clinic Brain Bank participants used in Reddy et al.
+but not used in the present CAA GWAS ($\beta = -0.068$, $P \text{-value} = 0.0012$), providing additional evidence for the APOE locus being important for CAA pathology beyond the known effects of APOE $\epsilon$ haplotypes.
+
+Several variants in the novel CAA locus were lead eQTL for *APOC2* and brain expression in GTEx Colocalization analysis confirmed that the CAA risk locus shares a functional variant with both *APOC2* eQTL ($\text{PrC} = 89-97\%$ in brain tissues) and nearby mQTL ($\text{PrC} = 97\text{-}99\%$).
+This locus also colocalized with *APOE* expression in the wall of the aorta ($\text{PrC} = 94\%$) and prostate ($\text{PrC} = 86\%$) but not in any brain tissues.
+Furthermore, a lead QTL in this locus, rs4803779, was associated with *APOC2* expression but not *APOE* expression in the DLPFC in ROSMAP.
+
+Variants in the same locus were also mQTL in the DLPFC in ROSMAP for four methylation sites (rs4803779 $P \text{-value}= 3.3 \times 10^{-91}$ for top site cg09555818; Figure 3).
+We confirmed that two of the CpG sites affected by the CAA risk locus, cg09555818 ($\text{OR} =0.82$, $P \text{-value}= 0.004$) and cg13119609 ($\text{OR} =0.78$, $P \text{-value}= 0.0007$), were in turn significantly associated with CAA pathology.
+Both of these CpG sites are located within the *APOC4-APOC2* readthrough transcript region overlapping *APOC4* and *APOC2*.
+Notably, we found that methylation levels at cg09555818 and cg13119609 are highly positively correlated ($\rho = 0.90$, $P \text{-value} < 2.2 \times 10^{-16}$) in ROSMAP.
+Our results are consistent with the hypothesis that rs7247551 locus affects CAA risk through hypomethylation of CpG sites in the *APOC4-APOC2* region.
+These results are also consistent with previous studies in other human cohorts that implicate hypomethylation at cg09555818 and cg13119609 as potentially associated with Alzheimer's disease.[@walker2021; @walker2020; @shao2018] Collectively, these results provide circumstantial evidence indicates that *APOC2* may be the target gene of the rs7247551 CAA risk locus as it colocalizes with *APOC2* expression in multiple brain tissues in GTEx and the associated CpG sites are located in exon 3 of *APOC4-APOC2*.
+However, more research must be done to verify that *APOC2* is the target gene.
+
+One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $P \text{-value}= 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $P \text{-value}= 1.5 \times 10^{-11}$).
+Additionally, a locus within *GRN* was associated with HS ($\text{OR} = 0.72$, $P \text{-value}= 3.8 \times 10^{-8}$).
+Both of these genes are associated with frontotemporal lobar degeneration,[@rollinson2011; @ciani2019] HS,[@katsumata2017; @nelson2015] and ADD.[@bellenguez2022; @wightman2021] We found that HS, LATE-NC, and ADD all colocalize at these two loci ($\text{PrC}\ge 90\%$, Figure 5A).
+These results indicate that HS, LATE-NC, and LOAD likely share causal loci for these genes.
+When NPE were regressed on known AD variants, Braak NFT stage was significantly associated with *TMEM106B* with a direction of effect opposite of both recent AD GWAS and other neuropathologies,[@bellenguez2022; @wightman2021] while neuritic plaques were not associated with *TMEM106B* in our study.
+Evidence from RNA-Seq analyses supported the hypothesis that *TMEM106B* may have discordant effects on TDP-43 and NFT pathology, and both the genetic and gene expression associations between *TMEM106B* and NFT were strengthened when adjusting for LATE-NC stage.
+While preliminary, these results support the possibility that the associations between *TMEM106B* and AD found in recent GWAS are driven by non-AD pathologies, namely TDP-43, in dementia cases, and may mask a more complex association between *TMEM106B* and NFT.
+
+A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $P \text{-value}= 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $P \text{-value}= 6.6 \times 10^{-8}$).
+We verified through colocalization analysis that the same locus drives association signals with Braak NFT stage and CERAD score.
+In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD.[@bellenguez2022] Interestingly, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $P \text{-value}= 0.15$) nor CAA ($\text{OR} = 1.02$, $P \text{-value}= 0.49$).
+Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology.[@holler2014; @franzmeier2019] As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles,[@nelson2009] our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.
+
+One locus on Chromosome 12q34 with lead variant rs2000660 (minor allele frequency $[\text{MAF}]= 9.9\%$ in pooled data set) located 12 kbp upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis ($\text{OR} =0.71$, $P \text{-value}= 3.9 \times 10^{-8}$).
+In previous GWAS, researchers have reported the *COL4A1/COL4A2* locus to be associated with numerous other vascular phenotypes, including peripheral artery disease, coronary artery disease, stroke, and arteriolar stiffness.[@klarin2019, @nikpay2015, @steffensen2018] The *COL4A1/COL4A2* locus has also been found to be associated with rare familial cerebrovascular diseases and lacunar ischemic stroke[@blevins2021; @rannikmae2017] In a recent GWAS, rs2000660 in particular was a lead risk variant for migraines.[@hautakangas2022] The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV in vascular disease is possibly related to disruption of the extracellular matrix.[@steffensen2018] *COL4A1* in the brain is preferentially expressed in endothelial cells (Figure 1**)** and codes for a component of collagen IV, an important component of basal lamina.
+rs2000660 was not nominally associated with any other vascular NPE in our study, and a previous GWAS of cerebral atherosclerosis using ROSMAP participants did not identify the *COL4A1* as a risk locus.[@vattathil2021] However, the sample size of the previous study was significantly smaller than the one used here (1,325 vs `r natheroscler`).
+Indeed, in the present study, rs2000660 reached genome-wide significance only in the pooled Stage 3 analysis, though it was nominally significant in the ROSMAP-only replication analysis in Stage 2 ($P \text{-value}=0.0079$).
+rs2000660 was not a QTL in GTEx or ROSMAP.
+However, rs650724, a variant in high LD with rs2000660 ($r^2=0.84$), is a synonymous coding variant (p.Ser1600Ser in ENST00000375820.10; p.Ser319Ser in ENST00000650424.1) for *COL4A1*, presenting a possible functional variant driving association in this locus.
+
+Another novel genome-wide significant locus located on Chromosome 17p13 was associated with Braak NFT stage and had a lead variant of rs72807981 ($\text{OR} =0.71$, $P \text{-value}= 3.9 \times 10^{-8}$, $\text{MAF}= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival.
+This variant was suggestively associated with Braak NFT stage in NACC ($\text{OR} =0.69$, $P \text{-value}= 1.8 \times 10^{-7}$) and was nominally replicated in ACT ($\text{OR} =0.50$, $P \text{-value}= 0.028$).
+The same variant was also nominally associated with neuritic plaques in the pooled analysis ($\text{OR} =0.82$, $P \text{-value}= 0.0047$) but was not nominally associated with any other AD NPE.
+rs72807981 is not a QTL in GTEx or ROSMAP, nor were any of the variants in high LD with it in our study.
+One variant in LD ($r^2= 0.92$) had a CADD score of 11.65, indicating a potential functional variant.
+There is previous research suggesting that *PIK3R5* is more highly expressed in aged adults with Braak NFT stages V and VI vs.
+non-demented controls (false-discovery rate $[FDR]= 0.01$).[@guennewig2021] *PIK3R5* is expressed preferentially in microglial cells in humans (Figure 3F),[@zhang2016] suggesting that its association with neurofibrillary pathology may be immune-mediated.
+
+Add discussion of novel arteriolosclerosis locus.
+
+In conclusion, we identified several promising novel loci associated with NPE and replicated multiple known risk loci for AD using NPE.
+We also provided additional context and consideration for relationships between specific risk loci and different NPE.
+Our study demonstrates the importance of studying genetic risk factors of neuropathology endophenotypes as a complement to studies of clinical and proxy phenotypes of Alzheimer's disease.
 
@@
 ```{r}
-np <- readRDS("data/mega/mega_np.Rds")
+np <- readRDS("data/adc/np_qced.Rds")
 nadrc <- np[, length(table(NACCADC))]
@@
 
-We used genotype and NPE data from four aged autopsy cohorts: National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Memory and Aging Project (ROSMAP), Adult Changes in Thought (ACT), and the AD Neuroimaging Initiative (ADNI).
-All study participants were de-identified and deceased, and archival samples were exclusively used.
-Therefore, our study does not fall under the definition of "Human Subjects Research," according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- involving the collection/study of data or specimens if publicly availablr, or/or recorded such that subjects cannot be identified.
+**NACC:** The present study used National Alzheimer's Coordinating Center (NACC) data from `r nadrc` National Institute on Aging-funded Alzheimer's Disease Research Centers (ADRCs).
+Individual ADRCs use different recruitment strategies and perform autopsies on-site, but neuropathology data at each ADRC are collected using a standard form (<https://files.alz.washington.edu/documentation/np11-form.pdf>) and submitted to NACC where they are aggregated and anonymized.
+The NACC Neuropathology (NP) data set based on the first version of this form was originally implemented in 2001,[@besser2018] and this analysis uses data from then through the December 2021 freeze.
+Participants were excluded if they did not have autopsy data available or if they were noted in the NP data set to have at least one of 19 conditions that could potentially bias results.
+These conditions include brain tumors, severe head trauma, and fronto-temporal lobar degeneration.
+(See **Supplementary Table S1** for full list of variables used for exclusion criteria.)
 
-This analysis used NACC data from `r nadrc` National Institute of Aging-funded Alzheimer's Disease Research Centers (ADRC).
-Individual ADRC use different recruitment methods and perform autopsies on-site, but participant data at each ADRC are collected using a standard form (<https://files.alz.washington.edu/documentation/np11-form.pdf>) and submitted to NACC where they are aggregated.
-The NACC Neuropathology (NP) data set based on the first version of this form was originally implemented in 2001[@besser2018], and this analysis uses data from then through the December 2021 freeze.
-Participants were excluded if they did not have autopsy data available or if they were noted in the NP data set to have at least one of 19 conditions that could potentially invalidate results.
-These conditions include brain tumors, severe head trauma, and fronto-temporal dementia (see **Supplementary Materials** for full list of variables used).
+**ROSMAP:** ROSMAP consists of harmonized data from two longitudinal cohort studies: The Religious Orders Study (ROS) and the Rush Memory and Aging Project (MAP).[@bennett2018] ROS and MAP were both approved by an Institutional Review Board of Rush University Medical Center.
+All participants signed an Anatomic Gift Act, as well as informed and repository consents.
+ROS began in 1994 and has recruited over 1500 Catholic priests, nuns, and brothers across the United States.
+MAP started in 1997 and has enrolled more than 2300 community members in the greater Chicago area of northeastern Illinois.
+The ROSMAP NP data used in this study were received from Rush University Medical Center in January 2020.
 
-ROSMAP has been previously described and consists of harmonized data from two longitudinal cohorts: The Religious Orders Study (ROS) and the Memory and Aging Project (MAP).[@bennett2018] ROS began in 1994 and has recruited over 1,100 Catholic priests, nuns, and brothers across the United States.
-MAP started in 1997 and enrolls community members in northeastern Illinois.
-The ROSMAP NP data used in this study was received from Rush University researchers in January 2020.
+**ACT:** The Adult Changes in Thought (ACT) study began in 1994 and recruited residents in the greater Seattle area aged 65 years and older without dementia at time of enrollment.[@kukull2002] The study has expanded to include three cohorts with continued enrollment using the original enrollment criteria and has a current total of 4,960 participants across all three cohorts.
+The ACT NP data used in this study were obtained from Kaiser Permanente in May 2021.
 
-The ACT study began in 1994 and recruited residents in the greater Seattle area aged 65 years and older without dementia at time of enrollment.21 The goal of the study has expanded to include three cohorts and continuous enrollment using the same enrollment criteria and has a current total of 4,960 participants across all three cohorts.
-The ACT NP data used in this study were obtained from Kaiser Permanente in May 2020.
-
-The ADNI (<https://adni.loni.usc.edu>) was launched in 2003 as a public-private research partnership, led by Principal Investigator Michael W.
+**ADNI:** The Alzheimer's Disease Neuroimaging Initiative (ADNI; <https://adni.loni.usc.edu>) was launched in 2003 as a public-private research partnership, led by Principal Investigator Michael W.
 Weiner, MD.
-The primary goal of ADNI has been to test whether serial magnetic resonance imaging (MRI), positron emission tomography (PET), other biological markers, and clinical and neuropsychological assessment can be combined to measure the progression of mild cognitive impairment (MCI) and early AD.
+The primary goal of ADNI has been to test whether serial magnetic resonance imaging, positron emission tomography (PET), other biological markers, and clinical and neuropsychological assessment can be combined to measure the progression of mild cognitive impairment and early AD.
 A subset of ADNI participants undergo autopsy and receive neuropathological phenotyping.
@@
 
+Some participants in ROSMAP and ADNI also had neuropathology and genotype data available in NACC.
+In these cases, records in the NACC were preferentially kept to maximize sample size of the initial Stage 1 analysis using only NACC participants (see Statistical Analyses subsection).
+An overview of our study design is presented in Figure 1.
+
+All study participants were deceased and the resulting data were de-identified for University of Kentucky investigators, and we exclusively used archival samples.
+Therefore, our study does not fall under the definition of "Human Subjects Research" according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- "involving the collection/study of data or specimens if publicly available, or/or recorded such that subjects cannot be identified."
+
 ## Genotype data and quality control
 
-Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS.[@shade2022] Genotype data for all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021] NACC and ACT raw genotype data were obtained from the September 2020 freeze of the Alzheimer's Disease Genetics Consortium (ADGC) and subsequently imputed, while pre-imputed ROSMAP and ADNI genotype data were received from collaborators in the Hohman lab at Vanderbilt University in March 2021 and December 2021, respectively.
+Genotype data for all cohorts underwent imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021] NACC and ACT raw genotype data were obtained from the September 2020 freeze of the Alzheimer's Disease Genetics Consortium (ADGC) and subsequently imputed, while pre-imputed ROSMAP and ADNI genotype data were received from collaborators in the Hohman lab at Vanderbilt University in March 2021 and December 2021, respectively.
+Genetic variants with MAF \< 0.1% and imputation quality scores of \<0.8 were removed prior to further quality control measures.
+Due to small samples sizes of participants with substantial non-European ancestry, especially in replication cohorts, these participants were excluded from analysis.
+Standard GWAS quality control procedures were followed for variant and participant inclusion (see Supplementary Note).
 
-Quality control on genotype data was performed separately for each cohort.
-Firstly, only participants labeled as non-Hispanic white or similar (European, etc.) were used if available.
-After imputation, duplicate samples and samples without autopsy data available were removed.
-Genetic variants and participants were iteratively removed under no variants were missing in more than 5% of participants and no participants were missing more than 5% of variants (however, average genotype coverage was 99.7%, and no variants or participants were actually removed during this process).
-Then, participants with unusually high or low (+/- 3 standard deviations from mean) were removed.
-Finally, participants were merged with the 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012], and participants with substantial non-European ancestry, as determined by distance from EUR superpopulation centroid in principal component analysis using the first two principal components, were removed.
-Some participants had NP and genotype data available in more than one cohort.
-In these cases, records in the larger cohort were preferentially kept and discarded in other cohorts (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
+## Defining and harmonizing neuropathology endophenotypes for analysis
 
-## Defining and harmonizing NPE for analysis
+In total, we combined and/or harmonized 11 neuropathology endophenotypes for analysis across the four studies.
+We note that there are differences in the way that some neuropathological data were collected across studies, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
+Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another.
+Hippocampal sclerosis (HS), microinfarcts, and gross infarcts were recorded as binary case-control phenotypes.
+Ateriolosclerosis, atherosclerosis, cerebral amyloid angiopathy (CAA), Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, diffuse amyloid plaques, limbic-predominant age-related TDP-43 encephalopathy neuropathologic change (LATE-NC), and Lewy body pathology were recorded as four-stage ordinal variables that either measured progressive severity of pathology ("none" \< "mild" \< "moderate" \< "severe") or progressing anatomical distribution of pathology.
+Braak NFT was recorded as a seven-stage ordinal variable that followed the anatomical distributional stages originally characterized by Braak and Braak.[@braak1991] We provide deeper description of our harmonization approach in the Supplementary Note, and a detailed listing of variables harmonized across data sources to construct neuropathology endophenotypes for analysis is available in **Supplementary Table 3**.
 
-We note that there are substantial differences in the way that some NPE are collected in different cohorts, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes.
-Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another.
-All code used for harmonization is available at <https://www.github.com/lincoln-shade/np_phewas>.
+## DNA methylation data
 
-In total, eleven NPE were harmonized across the four cohorts used.
-This included eight ordinal outcomes and three binary outcomes.
-Ordinal outcomes harmonized were brain arteriolosclerosis, atherosclerosis in the circle of Willis, Braak Stage for neurofibrillary tangles (NFT), cerebral amyloid angiopathy, Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, diffuse amyloid-beta plaques, limbic-predominant age-related TDP-43 encephalopathy neuropathologic change (LATE-NC), and Lewy body pathology.
-Braak Stage ranges from zero (absent NFT) to six (diffuse NFT throughout cortex and large loss of neurons).[@braak1991] LATE-NC had four levels, with "None" indicating lack of TDP-43 proteinopathy, "Mild" indicating TDP-43 deposits in the amygdala only, "Moderate" indicating deposits in the amygdala and limbic system, and "Severe" indicating deposits in amygdala, limbic system, and neocortex.
-Except where noted otherwise, other ordinal NPE were coded in four levels, "None," "Mild," "Moderate," or "Severe." Diffuse amyloid plaque pathology followed the Thal staging of amyloid deposition in NACC, ACT, and ADNI.
-In ROSMAP, diffuse plaques are examined and quantified in five regions (midfrontal cortex, entorhinal cortex, inferior parietal cortex, and hippocampus), then scaled by each regions standard deviation and then averaged.
-ROSMAP participants were labeled as "None" if their averaged score was equal to zero, "Mild" if their score was higher than zero but $\le$ 0.5, "Moderate" if their score was between 0.5 and one, and "Severe" if their score was above one.
-These labels roughly corresponded to score quartiles.
+Pre-processed and quality-controlled DNA methylation data for 740 ROSMAP participants were downloaded from Synapse.org (Synapse IDs: syn3157275 and syn3191087).
+DNA methylation preparation and quality control measures have been previously described.[@yu2015; @dejager2018] Briefly, approximately 50mg of frozen gray matter tissue from the dorsolateral prefrontal cortex (DLPFC) were sampled from each participant.
+DNA was then extracted and processed using the Illumina Infinium HumanMethylation450 BeadChip.
+Quality control measures included removing low-quality probes; removing participants with poor bisulfite-conversion efficiency; and adjusting methylation levels by age, sex, and batch, which adequately controlled for batch effects.[@dejager2018] Missing methylation levels were imputed using 100-nearest neighbors.[@yu2015; @dejager2018]
 
-The three binary NPE outcomes included hippocampal sclerosis (HS), gross infarctions, and microinfarcts.
-Participants were labeled as cases for HS if they have unilateral or bilateral HS, cases for gross infarctions in they have at least one chronic or acute gross infarction, and cases for microinfarcts if they had one or more microinfarcts in any region observed.
+## RNA-Seq data
+
+Pre-processed and quality-controlled bulk-tissue RNA-Seq data from the DLPFC of ROSMAP participants were downloaded from Synapse.org (Synapse IDs: syn21088596, syn21323366, syn3505732, syn3505724).
+As previously described, samples were prepared by sectioning approximately 100 mg of gray matter tissue from the DLPFC and RNA extracted using the Qiagen MiRNeasy Mini (cat no.
+217004) protocol and then submitted for transcriptome library construction using the dUTP protocol and Illumina sequencing.[\@dejager2018] A total of 634 participants in seven batches were sequenced with an average sequencing depth of 50 million paired reads per sample.[\@dejager2018] Subsequent quality control and batch corrections were performed, and the final output of the RNA-Seq pipeline was fragments per kilobase of transcript per million mapped reads (FPKM).[@dejager2018]
 
@@
 
-Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package,[@bi2021] while binary variables were analyzed with logistic mixed-effects models using the SAIGE R package.[@zhou2018] Covariates included age at death, sex, cohort, and the first 10 genetic principle components created using the PC-AiR method in the GENESIS R package.[@conomos2015].
-Dense genetic relationship matrices (GRM) created using a pruned set of independent ($r^2 < 0.2$) were used to account for relatedness between participants.
+We analyzed ordinal endophenotypes using proportional odds logistic mixed-effects models implemented in the POLMM R package[@bi2021] and analyzed binary variables similarly with logistic mixed-effects models implemented in the SAIGE R package.[@zhou2018] Fixed-effect covariates included age at death, sex, cohort, and the first 10 genetic principal components (PCs) created using the PCA in Related Samples (PC-AiR) method in the GENESIS R package.[@conomos2015] We included a dense genetic relationship matrix (GRM) as a random effect to account for relatedness between participants.
 An additive mode of inheritance was assumed in all analyses.
-Analyses proceeded in two stages: in stage one, null models with fixed covariates and GRM were fitted using either POLMM or SAIGE.
-In stage 2, score tests were performed on each variant with saddle-point approximation used to calculate p-values.
-Variants achieving a p-value of $p < 5\times 10^{-8}$ were considered "genome-wide" significant.
-To identify independent risk loci, variants were clumped using PLINK 1.9[@chang] with a linkage-disequilibrium threshold of $r^2 \le 0.05$.
 
-For NPE available in multiple cohorts, single-variant analyses were first performed in NACC participants.
-Lead variants with p-values exceeding a suggestive threshold of $p < 1 \times 10^{-6}$ were then tested in ROSMAP, ACT, and/or ADNI as available for attempted validation.
-Participants were then pooled for genome-wide mega-analysis.
+Analyses of individual data sources proceeded in two stages.
+In stage one, GRM were constructed using a pruned set of independent variants, defined as having a pairwise $r^2 <0.2$ within moving windows of 15 kilobase pairs (kbp).
+Null models, which including fixed covariates and the GRM, were then fitted using either the POLMM or SAIGE R packages.
+In stage 2, score tests were performed on each variant with a saddle-point approximation used to calculate p-values.
+We considered all variants with a $P \text{-value}< 5\times 10^{-8}$ to be genome-wide significant.
+To identify independent risk loci, we clumped results using the "\--clump" flag in PLINK 1.9 with the pairwise linkage-disequilibrium (LD) threshold set to $r^2 \le 0.05$.[@chang; @chang2015] Following analyses of individual cohorts, we performed fixed-effects meta-analyses using the METAL software using inverse-variance weighting on variants with MAF \>1% in each cohort.[@metal:f]
 
-To examine whether variants in the *APOE* region, defined as the region with 200 kilo-basepairs (KBP) from the start and end of the APOE coding region, independent of *APOE* $\epsilon$ haplotypes influence NPE risk, we re-analyzed Chromosome 19 using *APOE* $\epsilon$ diplotype as an additional covariate for NPE with significant association signals within the *APOE* locus in the main mega-analysis.
-*APOE* diplotypes were determined using the rs7412 and rs429358 variants according to the SNPedia online reference.[@apoe-s] Diplotype $\epsilon3/\epsilon3$ was used as the reference, while $\epsilon2/\epsilon2$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, and $\epsilon4/\epsilon4$ were adjusted for using fixed effect indicator variables.
-For NPE with multiple independent ($r^2 < 0.05$) association signals in the *APOE* region in the main analysis not adjusting for *APOE* $\epsilon$ diplotype, only the most significant signal was reported in our results.
-Additional signals were only reported if they were significant in the *APOE*-adjusted models.
+### Conditional analysis of the *APOE* region
 
-### Multivariate analyses
+The region surrounding the *APOE* gene on Chromosome 19 is consistently the single strongest genetic risk factor for LOAD in GWAS.
+Three common forms of the *APOE* gene -- $\epsilon 2$, $\epsilon 3$, and $\epsilon 4$ -- are present in our study populations (see Supplementary Table 2 for distribution of *APOE* alleles in participants), and the $\epsilon 2$ and $\epsilon 4$ alleles are associated with lower and higher risk of LOAD, respectively, relative to the $\epsilon 3$ allele.[@reiman2020] We therefore expected that variants in the *APOE* region, defined as the region within 200 kbp from the start and end transcription sites of *APOE*, would be associated with multiple NPE in our study.
+Moreover, we hypothesized that genetic variants in the *APOE* region may influence neuropathology risk independently of the effects of *APOE* $\epsilon$ alleles.
+To test this hypothesis, we re-analyzed variants in Chromosome 19 while adjusting for *APOE* $\epsilon$ genotype.
+We limited re-analysis to endophenotypes with at least one genome-wide significant association signal within the *APOE* locus in the Stage 3 GWAS.
+*APOE* genotypes were determined using the rs7412 and rs429358 variants according to the SNPedia online reference.[@apoe-s] Both variants had very high imputation quality scores in our data ($r^2 = 0.997 \text{ and } 0.975$, respectively).
+The $\epsilon3/\epsilon3$ genotype was used as reference, and we included fixed-effect indicator variables to adjust for $\epsilon2/\epsilon2$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, and $\epsilon4/\epsilon4$ genotypes.
+We chose this approach rather than adjusting for counts of $\epsilon 2$ and $\epsilon 4$ alleles because it is robust to potential non-linear effects of genotypes.
 
-Because many NPE have high phenotype correlation that single-outcome association analysis cannot take into account, we sought to perform genetic association analyses that could better account for the common co-occurrence of of NPE.
-We first assessed the phenotype correlation of NPE used in the pooled GWAS using polychoric correlation and grouped NPE visually using a dendrogram.
-Principal components analysis (PCA) was then performed using the psych R package on each group of NPE.
-The first PC (PC1) of each group was then used as the outcome variable for GWAS using SAIGE in a two-stage analysis.
-In the first stage, PC1 was regression on fixed-effects covariates using the GRM to account for relatedness.
-Because PC1 was not normally distributed, the residuals of stage one were then rank inverse normalized and regressed on covariates and GRM again in stage two.
-Score tests were then performed on genetic variants and saddle-point approximation used to estimate p-values.
+### Replication of known AD risk loci in NPE
 
-### Gene-based analyses
+We used the 83 AD and proxy-AD loci presented in a recent large GWAS to investigate whether AD-associated loci were associated with NPE.[@bellenguez2022] We restricted our comparison to AD loci with lead variants with MAF $\ge 0.01$, which excluded 3 loci, leaving 80 loci for comparison.
+We controlled the false-discovery rate for each NPE using the Benjamini-Hochberg procedure.[@benjamini1995] Variants with an adjusted Q-value $\le 0.05$ were considered significant.
 
-Gene start and end sites were determined using NCBI version 38 gene regions.
-Variants were mapped to a gene if they were located within 10 KBP of the gene's start or end sites.
-Using an unrelated set of the pooled participant data set, gene-based analyses were peformed using MAGMA v1.10.[@leeuw2015] The same fixed-effects covariates used in single-variant analyses were used, and the top variant principal components that accounted for 99.9% of variance in a gene's region were used to test for significance using an F-test.
-Genes with $p < 2.5 \times 10^{-6}$ were considered significantly associated with NPE.
+### FUMA annotation, gene-prioritization, and functional enrichment pipeline
+
+We mapped variants to genes and performed subsequent gene and gene-set analyses using the FUMA pipeline.[@watanabe2017] Variants were mapped to genes if they had $P \text{-values}\le 1\times10^{-5}$ in the Stage 3 analysis and were located within 10 Kbp of a protein-coding gene's transcription start or end sites.
+Gene-based analyses were performed using MAGMA.
+The top variant PCs that accounted for 99.9% of variance in a gene's region were used to test for significance using an F-test.
+We considered genes with resulting P-values of $P \text{-value}\le 2\times10^{-6}$ to be significantly associated with NPE.
+Gene-set enrichment analyses were performed using MAGMA gene-set analysis of Gene Ontology (GO) and curated gene sets from MSigDB.[@liberzon2011] Bonferroni P-value corrections were made for each NPE individually.
 
@@
 
-All variants in each logistic GWAS with p-values $< 1 \times 10^{-5}$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues and in ROSMAP bulk RNA-seq data in the dorsolateral prefrontal cortex.
-Colocalization analysis using the coloc R package was then performed on GWAS loci with significant QTLs.
-For ordinal variables, cut points widely used in previous studies were chosen to determine case-control proportions.
-A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) $> 80\%$ was used as a threshold for evidence of colocalization.
+We used multiple sources of publicly available summary statistics from external studies as data sources for Bayesian colocalization analyses.
+First, we downloaded Genotype-Tissue Expression Project (GTEx) v8 European ancestry quantitative trait loci (QTL) analysis summary statistics, which contains summary statistics for significant gene expression and splicing QTL variants (eQTL and sQTL, respectively) in 48 body tissues.[@gtexconsortium2017] We also used gene expression and DNA methylation QTL (mQTL) analysis summary statistics from studies using tissue from the dorsolateral prefrontal cortex of ROSMAP participants.[@ng2017] These studies examined the associations of genetic variants with molecular traits and provide curated lists of significant QTL variants.
+Finally, we downloaded the summary statistics from a recent GWAS of LOAD for a targeted *post hoc* colocalization analysis in the *TMEM106B* and *GRN* genes.[@bellenguez2022]
 
-To investigate whether shared GWAS signals drive association between multiple NPE, colocalization analysis was also performed on loci with variants exceeding a p-value threshold of $p < 1 \times 10^{-4}$ and concordant effect direction for more than one NPE in the pooled mega-analytic GWAS.
+For each neuropathology endophenotype outcome in our study, we first compiled a list of genetic variants with p-values $\le 1 \times 10^{-5}$ in our Stage 3 GWAS.
+We then queried the lists of significant QTL variants in GTEx and ROSMAP to identify neuropathology-associated QTL variants.
+For each genetic locus associated with neuropathology endophenotypes that had at least one significant QTL in either GTEx or ROSMAP, we performed colocalization analysis using the "coloc.abf" function in the coloc R package.[@giambartolomei2014] For ordinal variables, we chose dichotomizing cut points to determine case-control proportions.
+We used coloc's default prior probability of colocalization (PrC) of $\text{PrC} = 1\times 10^{-5}$ and considered a posterior $\text{PrC}> 80\%$ as a threshold for evidence of colocalization.
 
-### Conditional and mediation analyses
+To investigate whether shared GWAS signals drive association among multiple neuropathology endophenotypes, we also performed colocalization analysis on loci with variants exceeding a p-value threshold of $P \text{-value}< 1 \times 10^{-4}$ and concordant effect direction for at least two NPE in the Stage 3 analysis.
 
-For genetic loci associated with multiple NPE, we considered mediation hypotheses to test whether the association of the lead variant with one NPE was mediated by another NPE.
-Regression analyses were run in R using the `glm` function for binary outcome traits or `polr` function in the MASS[@venables2002] package for ordinal traits.
-Mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.
-To assess the possibility of reverse causation, mediation analyses were performed with both NPE in each pair as outcomes.
+### Association analyses using DLPFC DNA methylation and bulk RNA-seq data from ROSMAP
 
-# Results
+ROSMAP participants had post-mortem bulk tissue samples collected from the dorsolateral prefrontal cortex (DLPFC) which underwent DNA methylation quantification using Illumina DNAMethylation450 chip and gene expression and RNA-Seq using Illumina HiSeq2000.[@dejager2018] In total, 708 ROSMAP participants had DNA methylation data available for analysis.
+We restricted analyses involving DNA methylation or RNA-Seq data to NPE-associated loci that reached the genome-wide significance threshold in the meta-analysis and also colocalized with mQTL or eQTL in a brain tissue in either GTEx or ROSMAP.
 
-```{r}
-covar <- fread("data/mega/mega_np.covar")
-nmega <- formatC(covar[, .N], big.mark = ",")
-nnacc <- formatC(covar[FID == "0", .N], big.mark = ",")
-nrosmap <- formatC(covar[FID == "1", .N], big.mark = ",")
-nact <- formatC(covar[FID == "2", .N], big.mark = ",")
-nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")
-```
+In our *APOE* $\epsilon$-adjusted genetic association analysis, one locus near *APOE* remained significantly associated with cerebral amyloid angiopathy.
+This locus colocalized with DNA methylation levels at four CpG sites in ROSMAP.
+To investigate whether these CpG sites were in turn associated with CAA pathology, we combined individual-level DNA methylation and neuropathological data in ROSMAP for analysis.
+We used cumulative logit models using the "clm" function implemented in the "ordinal" R package with the semi-quantitative CAA variable described above as the outcome for analysis.
+We performed four analyses, with one of each of the four CpG sites tested as the independent variable of interest in each analysis.
+We adjusted for age, sex, ROS vs MAP study, bisulfite conversion efficiency, post-mortem interval, and *APOE* $\epsilon$ genotype in each analysis.
+Similar models were used to test associations between HS and LATE-NC and methylation levels at CpG sites cg09613507 and cg23422036.
+Wald tests were performed on the resulting parameter estimates to test for statistical significance.
 
-```{r}
-t1_env <- new.env()
-source("code/manuscript/01_table1_participant_demographics.R",
-       local = t1_env)
-t1dt <- t1_env$t1dt
-```
+For genes with significant eQTL in GTEx or ROSMAP that colocalized with NPE, we performed additional targeted analyses to assess the association between gene expression and NPE.
+We first assessed the association between NPE lead variants and gene expression in ROSMAP to confirm nominal eQTL status.
+We then performed generalized linear regression models between square-root or log-transformed mRNA expression and NPE outcomes adjusting for age at death, sex, PMI, and RNA integrity number.
 
-In total, `r nmega` participants across all cohorts underwent autopsy, had genotype data available, and passed quality control measures.
-This included `r nnacc` NACC, `r nrosmap` ROSMAP, `r nact` ACT, and `r nadni` ADNI participants.
-Participant demographics are shown in **Table 1**.
-NACC participants were younger (mean age-of-death 81 years) compared to ROSMAP (90 years) and ACT (88 years) participants, and had more balanced sex ratios, with 50% of NACC participants female vs.
-68% and 58% in ROSMAP and ACT, respectively.
-NACC participants were also more likely to have an APOE $\epsilon 4$ allele (55%) vs.
-ROSMAP (25%) or ACT (28%).
-Across the 13 NPE harmonized across multiple cohorts, NACC and ADNI tended to show more advanced neuropathology than ROSMAP and ACT.
-Using CERAD neuritic plaque score as an example, 9% of NACC participants were rated as "none" for CERAD score, whereas 24% and 23% of ROSMAP and ACT participants were rated as "none," respectively.
-At the other extreme, 65% of NACC participants were rated as having "severe" neuritic plaque pathology, whereas only 33% and 27% of ROSMAP and ACT participants, respectively, were rated as such.
-This trend can be seen across **XX** of the NPE harmonized across cohorts and is shown in **Supplementary Table S2**.
+### Replication of CAA locus Using Mayo Clinic Neuropathology GWAS
 
-## Single-variant analysis
+We used data from Mayo Clinic Brain Bank participants available from the Reddy et al.
+(2021) study of the genetic risk factors of CAA (data set heretofore referred to as MC-CAA) to attempt to replicate a novel CAA locus in the present study in an independent sample.[@reddy2021] Neuropathology and genetic variant data were downloaded from Synapse (Synapse IDs: syn10930250, syn21499318, syn21522653, and syn21547862).
+Eight participants were identified as duplicates between batches or with NACC participants and removed.
+Whereas CAA is graded on a four-level ordinal scale in the present study, CAA in MC-CAA is graded as an average of CAA burden across five brain regions.[@reddy2021] We therefore used linear regression with the outcome variable as sqrt(CAA) with the independent variable of interest being the number of G alleles of variant rs7247551.
+Covariates included *APOE* diplotype ($\epsilon3/\epsilon3$, $\epsilon2/\epsilon4$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, or $\epsilon4/\epsilon4$), sex, age at death (truncated at 90 years), and the first 3 genetic PCs.
 
-```{r}
-t2_env <- new.env()
-source("code/manuscript/02_table2_mega_analysis_results.R",
-       local = t2_env)
-```
+# List of abbreviations
 
-In the NACC analysis, no variants outside of the APOE region were significantly associated with any NPE.
-Variants within this region were significantly associated with Braak stage, CERAD score, CAA, and diffuse amyloid plaques.
-Variants reaching a suggestive significance threshold ($p < 1 \times 10^{-5}$ and replicating at a nominal significance threshold of $p < 0.05$ are shown in Table 2 below.
-Variants outside og the APOE region that reached a type-I error rate below $0.05$ using Benjamini-Hochberg adjustment included &&&.
+ACT, the Adult Change in Thought Study; AD, Alzheimer's disease; ADGC, Alzheimer's Disease Genetics Consortium; ADNI, AD Neuroimaging Initiative; ADRC, Alzheimer's Disease Research Center; CAA, cerebral amyloid angiopathy; eQTL, expression QTL; GRM, genetic relationship matrix; GTEx, Genotype-Tissue Expression Project; GWAS, genome-wide association study; HS, hippocampal sclerosis; kbp, kilobase pairs; LATE-NC, LATE neuropathologic change; LATE, limbic-predominant age-related TDP-43 encephalopathy; LD, linkage disequilibrium; LOAD, late-onset Alzheimer's disease; MAP, Rush Memory and Aging Project; mQTL, methylation QTL; NACC, National Alzheimer's Coordinating Center; NFT, neurofibrillary tangle; NP, neuropathology; NPE, neuropathology endophenotype; OR, odds ratio; PC-AiR, PCA in Related Samples; PC1, first principal component; PCA, principal components analysis; PET, positron emission tomography; PrC, probability of colocalizationQTL, quantitative trait locus; ROS, Religious Orders Study; ROSMAP, the Religious Orders Study and the Rush Memory and Aging Project; sQTL, splicing QTL; TOPMed, Trans-'Omics for Precision Medicine.
 
-In total, fourteen independent ($r^2 < 0.05$) loci had lead variants that exceeded the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis.
-Of these, four were duplicate signals in the *APOE* region, leaving ten significant signals across 13 NPE.
-Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE*-adjusted analysis.
-Novel variant-NPE associations included a variant 12 KBP upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis and a *PIK3R5* intronic variant, rs72807981, associated with Braak stage (**Table 3**).
-Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak stage; and *TMEM106B* with HS and LATE-NC.
+\newpage
 
-## Multivariate analysis
+# Declarations
 
-Visual inspection of a dendrogram (**Supplementary Figure S2**) of NPE based on polychoric correlations reveals two primary clusters of NPE: an AD NPE cluster consisting of Braak stage, CERAD score, diffuse amyloid, and CAA; and a vascular NPE cluster consisting of atherosclerosis, arteriolosclerosis, gross infarcts, and microinfarcts.
-PC1 in the AD cluster accounted for 66% of variance in the cluster, while PC1 in the vascular cluster accounted for 50% of the variance in that cluster.
-Only the *APOE* region was significantly associated with the AD cluster PC1, while no loci were significantly associated with the vascular PC1.
+## Ethics approval and consent to participate
 
-## Gene-based analysis
+All study participants were deceased and the resulting data de-identified, and we exclusively used archival samples.
+Therefore, our study does not fall under the definition of "Human Subjects Research" according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- "involving the collection/study of data or specimens if publicly available, or/or recorded such that subjects cannot be identified."
 
-## Colocalization analyses
+## Consent for publication
 
-Because CAA was the only NPE to have an associated locus in the *APOE* region with a signal independent of *APOE* $\epsilon$ diplotype, this region was exluded from NPE-NPE colocalization analysis.
-Pairs of NPE with the same variant reaching a p-value threshold of $p < 1 \times 10^{-5}$ with concordant effect directions included Braak stage and CERAD score in the *BIN1* region on Chromosome 2 and HS and LATE-NC in the *TMEM106B* region on Chromosome 7.
-Both loci showed a high probability of colocalization, with both having a PPH4 $> 99.9%$.
+Not applicable.
 
-Multiple suggestive NPE loci showed evidence of colocalization with gene expression QTL (eQTL) in GTEx.
-A locus within *GRN* suggestively associated with HS ($p = 6.9 \times 10^{-8}$) colocalized with GRN expression (PPH4 $>99.9\%$), and the *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* eQTL.
-The CAA-associated *APOE* locus indepedent of *APOE* $\epsilon$ diplotype also colocalized with *APOE* eQTL in GTEx (PPH4 $= 95\%$) and with a methylation QTL (mQTL) in ROSMAP (PPH4 $>99.9\%$).
+## Availability of data and materials
 
-## Conditional and mediation analyses
+All code used for data preparation and analysis is available at <https://www.github.com/lincoln-shade/np_phewas>.
+ROSMAP data can be requested at <https://www.radc.rush.edu> and <https://www.synapse.org>.
+ADGC data is can be requested from NIAGADS at <https://www.niagads.org/resources/related-projects/alzheimers-disease-genetics-consortium-adgc-collection>.
+NACC neuropathology data can be requested at <https://naccdata.org/>.
+ACT data can be requested at <https://actagingresearch.org/>.
+ADNI data can be downloaded at <https://adni.loni.usc.edu/>.
+The results published here are in whole or in part based on data obtained from the AD Knowledge Portal.
 
-We performed conditional and mediation analyses on both loci that colocalized between two NPE: *BIN1*, which colocalized with Braak stage and CERAD score; and *TMEM106B*, which colocalized with HS and LATE-NC.
-In unconditional analyses (the single-variant analyses discussed above), the lead *BIN1* variant, rs6733839, had similar effect sizes for both Braak stage ($p = 3.8 \times 10^{-10}$, OR $= 1.21$) and CERAD score ($p = 6.6 \times 10^{-8}$, OR $= 1.20$).
-However, when Braak stage was included as a covariate for CERAD, the association between rs6733839 and CERAD score was markedly reduced (OR $= 1.03$, $p = 0.47$).
-In contrast, the association between rs6733839 and Braak stage was less attenuated when adjusted for CERAD score (OR $= 1.14$, \$p = \$).
-In the mediation analysis, we found that Braak stage mediated 87% of the effect of rs633839 on CERAD score, and the post-mediation effect size was not significantly different than zero.
-For the *TMEM106B* analysis, we used the lead variant for HS, rs4721058, which was significantly associated with both HS (OR $= 0.67$, $p = 1.7 \times 10^{-11}$) and LATE-NC (OR $= 0.71$, $p = 1.7 \times 10^{-8}$).
-Mediation analysis resulted in moderate mediation (\~50%) for both HS and LATE-NC.
+## Competing interests
 
-# Discussion
+JAS reported personal fees from Observational Study Monitoring Board Framingham, Observational Study Monitoring Board Discovery (National Institute of Neurological Disorders and Stroke), and Takeda Pharma.
+AJS reported support from Avid Radiopharmaceuticals, a subsidiary of Eli Lilly (in kind contribution of PET tracer precursor); Bayer Oncology (Scientific Advisory Board); Eisai (Scientific Advisory Board); Siemens Medical Solutions USA, Inc.
+(Dementia Advisory Board); NIH NHLBI (MESA Observational Study Monitoring Board); and Springer-Nature Publishing (Editorial Office Support as Editor-in-Chief, Brain Imaging and Behavior).
+All other authors declare that they have no competing interests.
 
-We performed GWAS and functional analyses for eleven neuropathology endophenotypes across four high-quality neuropathology cohorts with a maximum sample size of 7,463 participants.
-Our work builds on previous genetic association studies of NPE, Alzheimer's, and related dementias and provides another attempt to better understand the complex associations between different neuropathologies and genetic risk factors.
-We discovered intriguing new loci mapped to *COL4A1* and *PIK3R5* associated with atherosclerosis in the circle of Willis and Braak stage for neurofibrillary tangles, respectively.
-We also investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.
-However, our study has several important limitations, including sample size and heterogeneity in design between cohorts.
+## Funding
 
-## *COL4A1* and atherosclerosis
+**The author(s) disclosed receipt of the following financial support for the research, authorship, and/or publication of this article:** R56AG057191; F30NS124136; P30AG028383; the University of Kentucky Center for Clinical and Translational Science TL-1 Fellowship [grant number TL1TR001997]; the National Center for Advancing Translational Sciences [grant number UL1TR001998]; and the Dean of the College of Medicine, University of Kentucky.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH or the University of Kentucky.
 
-One locus on Chromosome 12q34 with lead variant rs2000660 (MAF $= 9.9\%$ in study participants) located 12 KBP upstream of *COL4A1* was significantly associated with atherosclerosis in the circle of Willis.
-*COL4A1* codes for a component of collagen IV, an important component of basal lamina.
-The *COL4A1/COL4A2* locus has been previously found to be associated with numerous other vascular phenotypes, including peripheral artery disease[@klarin2019], coronary artery disease[@nikpay2015], stroke(cite), and arteriol stiffness(cite) through GWAS.
-Additionally, the locus has been associated with rare familial cerebrovascular diseases[@blevins2021] and lacunar ischemic stroke[@rannikmae2017] rs2000660 in particular has recently been found to be a risk variant for migraine.[@hautakangas2022] The relevance of the *COL4A1* locus to cerebral vascular traits is thus highly supported by previous research, and the biological role of collagen IV to cerebrovascular disease is likely related to disruption of the extracellular matrix.(cite) rs2000660 did not show evidence of colocalization in GTEx or ROSMAP, so the functional effect of this variant, if indeed it is has one, is not obvious from our results.
-rs2000660 was located just outside the 10 KBP window used for gene-based analysis, and the *COL4A1* and *COL4A2* genes were not nominally associated with atherosclerosis in our study ($p>0.05$).
-Furthermore, rs2000660 was not nominally associated with any other vascular NPE in our study.
-A previous GWAS in ROSMAP of cerebral atherosclerosis did not identify the *COL4A1* as a risk locus.[@vattathil2021] However, this sample size of this study was significantly smaller than the one used in the present study (1,325 vs 6,959).
-Indeed, rs2000660 only reached genome-wide significance in the pooled mega-analysis, though it was nominally significant in the ROSMAP-only validation analysis ($p=0.0081$).
+Genotyping was supported by the Alzheimer's Disease Genetics Consortium through the National Institute of Aging [grant numbers U01 AG032984, RC2AG036528]
 
-## *PIK3R5* and Braak stage
+Samples from the National Cell Repository for Alzheimer's Disease (NCRAD), which receives government support under a cooperative agreement grant (U24 AG21886) awarded by the National Institute on Aging (NIA), were used in this study.
+We thank contributors who collected samples used in this study, as well as patients and their families, whose help and participation made this work possible;
 
-The other identified novel genome-wide significant locus located on Chromosome 17p13 had a lead variant of rs72807981 (MAF $= 6.3\%$), an intronic variant within *PIK3R5*, which codes for a phosphatidylinositol 3-kinase involved in cell growth, motility, and survival.
-One study which compared differential mRNA expression in aged adults with Braak stage VI vs.
-non-demented controls found *PIK3R5* to be more highly expressed in cases than controls (false-discovery rate $= 0.01$).[@guennewig2021]
+Data for this study were prepared, archived, and distributed by the National Institute on Aging Alzheimer's Disease Data Storage Site (NIAGADS) at the University of Pennsylvania (U24-AG041689-01).
 
-## *APOE* and NPE
+We thank the study participants and staff of the Rush Alzheimer's Disease Center.
+The Religious Orders Study and the Rush Memory and Aging Project are supported by grants from the National Institutes of Health: [grant number P30AG10161, P30AG72975, R01AG15819, R01AG17917, R01AG22018, R01AG33678, R01AG34374, R01AG36042, R01AG40039, R01AG042210, U01AG46152, U01AG61356, R01AG47976, R01AG43379, RF1AG54057, R01AG56352, R01NS78009, and UH2NS100599], and the Illinois Department of Public Health.
 
-Multiple variants in the APOE region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA.
-Additionally, APOE was also associated with LATE-NC.
-This result was unsurprising and consistent with previous GWAS of NPE.[@beecham2014 @dugan2021] While multiple loci in this region were significantly associated with each of these four NPE, only one locus with lead variant rs4803778 remained significant (OR $= 1.24$, $p = 5.8 \times 10^{-12}$) when adjusting for *APOE* $\epsilon$ diplotype.
-rs4803778 was in high LD ($r^2 = 0.97$) with rs4803779, which was also significantly associated with CAA (OR $= 1.24$, $p = 6.7 \times 10^{-12}$) and is a lead mQTL in ROSMAP for site cg09555818 ($p = 3.3 \times 10^{-91}$), indicating that this locus may be associated with CAA through epigenetic regulation of *APOE*.
-However, this locus is located approximately 30 KBP downstream of the *APOE* coding end site and situated between the *APOC4*, *APOC2*, and *CLPTM1*, and is a stronger eQTL of *APOC2* than *APOE*, although it is strongly significant and colocalizes with expression of both genes in multiple tissues.
+This research was funded by the National Institute on Aging (U19AG066567).
+Data collection for this work was additionally supported, in part, by prior funding from the National Institute on Aging (U01AG006781).
+All statements in this report, including its findings and conclusions, are solely those of the authors and do not necessarily represent the views of the National Institute on Aging or the National Institutes of Health.
+We thank the participants of the Adult Changes in Thought (ACT) study for the data they have provided and the many ACT investigators and staff who steward that data.
+You can learn more about ACT at: [https://actagingstudy.org/](https://actagingstudy.org/%22).
 
-## *BIN1*, neurofibrillary tangles, and neuritic plaques
+Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
+ADNI is funded by the National Institute on Aging, the National Institute of Biomedical Imaging and Bioengineering, and through generous contributions from the following: AbbVie, Alzheimer's Association; Alzheimer's Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F.
+Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research & Development, LLC.; Johnson & Johnson Pharmaceutical Research & Development LLC.; Lumosity; Lundbeck; Merck & Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
+The Canadian Institutes of Health Research is providing funds to support ADNI clinical sites in Canada.
+Private sector contributions are facilitated by the Foundation for the National Institutes of Health ([www.fnih.org](https://fnih.org/)).
+The grantee organization is the Northern California Institute for Research and Education, and the study is coordinated by the Alzheimer's Therapeutic Research Institute at the University of Southern California.
+ADNI data are disseminated by the Laboratory for Neuro Imaging at the University of Southern California.
 
-An locus approximately 30 KBP downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage (OR $= 1.21$, $p = 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques (OR $= 1.20$, $p = 6.6 \times 10^{-8}$).
-This locus is a known AD risk locus and is second only to *APOE* for strength of association with late-onset AD.[@bellenguez2022] However, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques (OR $= 1.05$, $p = 0.15$) nor CAA (OR $= 1.02$, $p = 0.49$).
-Previous research has identified that *BIN1* variant rs744373 is associated with tau-PET but not amyliod-PET levels[@franzmeier2019] and that *BIN1* protein expression is correlated with neurofibrillary table pathology but not with diffuse or neuritic plaques.[@holler2014] While the *BIN1* locus was associated with both neurofibrillary tangle and neuritic plaque pathology in our sample, neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles[@nelson2009], indicating that pathological process related to tau could mediate the association.
-We therefore performed mediation analysis to test the hypothesis that the association between the *BIN1* locus and neuritic plaques is mediated by overall tau burden, using Braak staging of neurofibrillary tangles as a proxy measurement.
-We found that tau burden completely mediated the association between BIN1 and neuritic plaques, as the association between *BIN1* and neuritic plaques was not longer significant after adjusting for neurofibrillary tangle pathology (OR = 1.03, p = 0.48, see **Figure X**).
-Our findings are therefore consistent with the hypothesis that the relationship between *BIN1* and Alzheimer's disease is mediated by tau rather than amyloid pathological processes.
+## Authors' contributions
 
-## Limitations and conclusion
+L.M.P.S.
+conceptualized study design, prepared data, performed analyses, and was a major contributor in writing the manuscript.
+Y.K.
+provided feedback on analyses and contributed to the manuscript.
+SC and ME helped prepare figures and provided extensive feedback on manuscript preparation.
+E.L.A.
+provided guidance on interpretation of *BIN1* results and provided extensive feedback on manuscript preparation.
+T.J.H.
+performed imputation and quality control on ROSMAP genotype data.
+K.N.
+and A.J.S.
+provided imputed and quality-controlled ADNI genotype data used in an earlier version of the manuscript and provided feedback on manuscript preparation.
+D.A.B.
+and J.A.S.
+provided ROSMAP neuropathology data and made critical revisions to the manuscript.
+P.T.N.
+provided guidance on defining neuropathology endophenotypes and contributed to the manuscript.
+D.W.F.
+conceptualized study design and provided feedback on manuscript preparation.
+All authors read and approved the final manuscript.
 
-There are multiple limitations to the approach we took in this study.
-The four cohorts we used, while all high quality, used different recruitment and design considerations.
-NACC recruitment is handled independently at each ADRC and likely consists of a more clinically representative population, whereas MAP and ACT particpants may be more representative of the broader aged communities from which they are recruited.
-ROS recruits Catholic brothers and sisters and may represent a population more highly educated than the United States as a whole.
-These differences are reflected in the relative frequencies and distributions of NPE between cohorts, with NACC and ADNI tending to show more severe pathology than ACT and ROSMAP (**Tables 1 and S2**).
-Additionally, there is likely substantially heterogeneity in the way that NPE data are collected and graded both within and between studies.
-We review these potential limitations briefly in our previous GWAS of brain arteriolosclerosis.[@shade2022]
+**ADD QI AND KHINE CONTRIBUIONS**
 
-Additionally, while recruitment, autopsy, and genotyping of the cohorts used is ongoing, the sample sizes available for genome-wide investigation of NPE genetic risk is small relative to their clinical counterparts.
-A recent GWAS of AD included over 111,000 clinically diagnosed or "proxy" (based on family history) AD cases and nearly 700,000 controls.
-In contrast, our largest analyzed sample size for CERAD score totalled 7,447 participants, while LATE-NC totaled only 2,224.
-These sample reflect immense progress in autopsy cohorts from even eight years ago and will only continue to improve, but availabel sample sizes still pose limitations on power of GWAS and similar 'omics approaches to studying disease risk using NPE.
-A strength of our study is that data on multiple NPE is available for each participant, allowing for multivariate approaches to studying NPE.
-However, as our PCA-based multivariate analyses did not identify any risk loci not identified in single-outcome analyses, this approach cannot overcome sample size limitations.
+## Acknowledgements
 
-In conclusion, we have identified several intriguing novel loci associated with NPE and validated multiple known risk loci for AD and related dementias in NPE.
-We also provide additional context and consideration for relationships between specific risk loci and different NPE.
-Overall, our study provides additional potential avenues of investigation into the relationship between genomics, neuropathology, and dementia.
+The NACC database is funded by NIA/NIH Grant U24 AG072122.
+NACC data are contributed by the NIA-funded ADRCs: P30 AG062429 (PI James Brewer, MD, PhD), P30 AG066468 (PI Oscar Lopez, MD), P30 AG062421 (PI Bradley Hyman, MD, PhD), P30 AG066509 (PI Thomas Grabowski, MD), P30 AG066514 (PI Mary Sano, PhD), P30 AG066530 (PI Helena Chui, MD), P30 AG066507 (PI Marilyn Albert, PhD), P30 AG066444 (PI John Morris, MD), P30 AG066518 (PI Jeffrey Kaye, MD), P30 AG066512 (PI Thomas Wisniewski, MD), P30 AG066462 (PI Scott Small, MD), P30 AG072979 (PI David Wolk, MD), P30 AG072972 (PI Charles DeCarli, MD), P30 AG072976 (PI Andrew Saykin, PsyD), P30 AG072975 (PI David Bennett, MD), P30 AG072978 (PI Neil Kowall, MD), P30 AG072977 (PI Robert Vassar, PhD), P30 AG066519 (PI Frank LaFerla, PhD), P30 AG062677 (PI Ronald Petersen, MD, PhD), P30 AG079280 (PI Eric Reiman, MD), P30 AG062422 (PI Gil Rabinovici, MD), P30 AG066511 (PI Allan Levey, MD, PhD), P30 AG072946 (PI Linda Van Eldik, PhD), P30 AG062715 (PI Sanjay Asthana, MD, FRCP), P30 AG072973 (PI Russell Swerdlow, MD), P30 AG066506 (PI Todd Golde, MD, PhD), P30 AG066508 (PI Stephen Strittmatter, MD, PhD), P30 AG066515 (PI Victor Henderson, MD, MS), P30 AG072947 (PI Suzanne Craft, PhD), P30 AG072931 (PI Henry Paulson, MD, PhD), P30 AG066546 (PI Sudha Seshadri, MD), P20 AG068024 (PI Erik Roberson, MD, PhD), P20 AG068053 (PI Justin Miller, PhD), P20 AG068077 (PI Gary Rosenberg, MD), P20 AG068082 (PI Angela Jefferson, PhD), P30 AG072958 (PI Heather Whitson, MD), P30 AG072959 (PI James Leverenz, MD).
+
+The results published here are in part based on data obtained from the AD Knowledge Portal.
 
@@
 
-# Tables
+# Figure captions
 
-```{r}
-t1_env$t1
-```
+**Figure 1: Overview of study design.** White boxes represent sources of data or summary statistics.
+Purple boxes represent genetic association analysis, while green boxes represent downstream functional analyses.
+Stage 1 GWAS were performed using National Alzheimer's Coordinating Center (NACC) neuropathology (NP) data set.
+Variants reaching a suggestive threshold of association ($P \text{-value} \le 1 \times 10^{-5}$) in Stage 1 then underwent attempted nominal replication ($P \text{-value} < 0.05$) separately in three independent studies: Religious Orders Study and the Rush Memory and Aging Project (ROSMAP), the Adult Changes in Thought (ACT) study, and the Alzheimer's Disease Neuroimaging Initiative (ADNI).
+In Stage 3, data from NACC, ROSMAP, ACT, and ADNI were then merged for pooled GWAS.
+Stage 3 summary statistics were then used in additional analyses, including attempted replication of Alzheimer's disease risk loci in neuropathology endophenotypes, gene-based and pathway analysis, and colocalization analysis.
+Significant Stage 3 loci that showed evidence of colocalization were then investigated using DNA methylation and RNA-Seq data from ROSMAP.
 
-```{r}
-t2_env$t2
-```
+**Figure 2: Manhattan plots of Stage 3 results for LATE-NC, Braak stage, cerebral atherosclerosis, and HS.** *P*-values are derived using saddle-point approximation and are truncated at $P \text{-value} = 1 \times 10^{-15}$.
+Genome-wide significant loci are highlighted in orange and annotated with mapped gene (known loci) or nearest protein-coding gene (novel loci).
+The solid black line represents the genome-wide significance level ($P \text{-value} = 5 \times 10^{-8}$), and the black dotted line represents the suggestive significance level ($P \text{-value} = 1 \times 10^{-5}$).
+
+**Figure 3: Novel associations identified between *COL4A1* and *PIK3R5* loci with cerebral atherosclerosis and Braak stage, respectively.** **A)** Regional LocusZoom plot of atherosclerosis-associated *COL4A1* locus.[@rj2010] Mb, megabase; uses hg19.
+**B)** Forest plot for individual and pooled cohorts for lead variant rs2000660 odds ratio and 95% confidence intervals.
+Lead variant reached a suggestive level of significance in NACC analysis ($\text{OR} = 0.72$, $P \text{-value}= 5.8\times10^{-7}$) and was nominally replicated in ROSMAP (OR = 0.68, p = 0.0079).
+ADNI results are not shown due to wide confidence interval (OR = 1.25, 95% CI = 0.08-18.84).
+**C)** Human brain cell-type expression profile of *COL4A1* in Zang et al.
+(2016).[@zhang2016] *COL4A1* is preferentially expressed in fetal astrocytes and endothelial cells with lower expression in neurons.
+FPKM, Fragments Per Kilobase of transcript per Million mapped reads.
+**D)** Regional LocusZoom plot of Braak-associated *PIK3R5* locus.
+Mb, megabase; uses hg19.
+**E)** Forest plot for individual and pooled cohorts for rs72807981 odds ratio and 95% confidence intervals.
+Lead variant rs72807981 reached a suggestive level of significance in NACC analysis ($\text{OR} = 0.69$, $P \text{-value}= 1.8\times10^{-7}$) and was nominally replicated in ACT ($\text{OR} = 0.50$, $P \text{-value}= 0.0028$).
+**F)** Human brain cell-type expression profile of *PIK3R5* in Zang et al.
+(2016).[@zhang2016] *PIK3R5* is primarily expressed in microglia.
+FPKM, Fragments Per Kilobase of transcript per Million mapped reads.
+
+**Figure 4: Locus in *APOE/APOC2* region associates with CAA and colocalizes with mQTL in ROSMAP.
+A)** Regional LocusZoom of associated locus.
+Mb, megabase; uses hg19.
+Eighteen genes are not shown due to space considerations.
+**B)** Regional plot showing CAA associations and colocalization with four mQTL in ROSMAP with CpG sites: cg04401876, cg09555818, cg10169327, and cg13119609.
+Posterior probability of colocalization equals 97% with each of the four mQTL.
+The darkly bordered box indicates the region with SNPs most associated with CAA and the four colocalizing mQTL.
+**C)** Associations between the same four CpG sites and CAA pathology in ROSMAP (N = 708).
+Two sites, cg09555818 (*p* = 0.004) and cg13119609 (*p* = 0.0007) were significantly associated with CAA pathology.
+**D)** Schematic showing regression beta estimates between SNP and CpG sites, SNP and CAA, and CpG sites and CAA.

### 2024-04-30 supplementary_note.Rmd
Organize code before publication.

---
output: word_document
bibliography: doc/references.bib
csl: doc/springer_basic.csl
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(data.table)
library(flextable)
library(magrittr)
library(table1)
library(CMplot)
source("./code/functions/plot_manhattan.R")
source("./code/functions/find_closest_gene.R")
source("./code/functions/get_results_list.R")
```

# Supplementary Material

## Quality control

Quality control on genotype data was performed separately for each data source. Quality control and inclusion/exclusion criteria closely followed that used in our previous brain arteriolosclerosis GWAS [@shade2022]. Briefly, we included only participants labeled as non-Hispanic white or similar (European, etc.) if available. After imputation, we first identified and subsequently removed duplicate samples using the KING software "\--duplicate" option [@manichaikul2010]. We then removed participants without autopsy data available by merging genotype sample identification data with neuropathology data sets and retaining only samples present in both data sets. We then iteratively removed genetic variants and participants until no variants were missing in more than 5% of participants and no participants were missing more than 5% of variants (however, average genotype coverage was 99.7%, and no variants or participants were actually removed during this process). We then excluded participants with unusually high or low ($\pm$ 3 standard deviations from mean) genetic heterogeneity, as measured by the PLINK 1.9 "\--het" flag [@chang]. Finally, we merged participants with the 1000 Genomes Phase 3 (1000 Genomes) cohorts [@the1000genomesprojectconsortium2012] and performed principal components analysis (PCA) on a subset of independent variants (measured by pairwise $r^2 < 0.2$). We excluded participants with substantial non-European ancestry, as determined by distance from the 1000 Genomes EUR superpopulation centroid using the first two principal components (PCs).

## NACC Exclusion Criteria

National Alzheimer's Coordinating Center (NACC) participants were excluded if they had any of the following conditions noted in the NACC Neuropathology Data Set. Variable names and descriptions are taken from <https://files.alz.washington.edu/documentation/rdd-np.pdf>. Variable descriptions may be lightly edited. Participants were not excluding for missing data in any of these fields.

```{r}
table_s1 <- fread("doc/nacc_exclusion_variables.txt", sep = "|") %>% 
  flextable() %>% 
  add_header_lines("Supplementary Table 1: NACC Exclusion Criteria")

table_s1
```

\newpage

## Harmonization of NPE

We combined or harmonized 11 neuropathology endophenotypes across the four data sources used. Arteriolosclerosis, Braak NFT stage, CAA, atherosclerosis, Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, microinfarcts, and gross infarcts had variables in each cohort with directly comparable coding definitions and were straightforwardly renamed and combined with minimal recoding. Analyzed arteriolosclerosis, CAA, atherosclerosis, and CERAD score variables each had four stages with the following labels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe"). Microinfarcts and gross infarcts were labeled either 0 ("absent") or 1 ("present"). Braak NFT Stage followed the staging criteria previously described in the literature and had seven levels, ranging from 0 (absent NFT) to six (diffuse NFT throughout cortex and large loss of neurons) [@braak1991].

LATE-NC was recorded differently in several of the data sets and was harmonized to a four-level outcome variable following the simplified staging of TDP-43 pathology outlined in Figure 3B of the 2019 LATE working group report [@nelson2019]. The following levels were used in the analyzed LATE-NC variable: 0, indicating lack of recorded TDP-43 proteinopathy; 1, indicating TDP-43 deposits in the amygdala only; 2, indicating deposits in the hippocampus or entorhinal cortex; and 3, indicating deposits in the neocortex. TDP-43 data was not available in ACT. In ROSMAP, TDP-43 pathology is recorded as a single variable following the same staging detailed above. In NACC and ADNI, the presence of TDP-43 pathology in each region of the brain is recorded as a separate binary indicator variable. To collapse TDP-43 pathology to a single ordinal variable in these data sets, we assigned a value based on the presence of the "highest" region where TDP-43 was present (*e.g.* a participant with TDP-43 pathology in both the hippocampus and the neocortex would be assigned a value of 3). Participants were labeled as 0 if they met two conditions: (1) they had recorded TDP-43 data available for at least one of the brain regions used for staging and (2) TDP-43 pathology was noted as absent in all the regions for which they had data available.

Diffuse amyloid plaque pathology was recorded as a four-stage Thal phase of amyloid deposition in NACC, ACT, and ADNI with the following levels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe"). In ROSMAP, diffuse plaques were examined and quantified in five regions (midfrontal cortex, entorhinal cortex, inferior parietal cortex, and hippocampus), then scaled by each region's standard deviation and averaged. To discretize this continuous variable in ROSMAP participants to a four-level variable, as recorded in the other data sets, we assigned participants a value of 0 ("none") if their averaged score was equal to 0, 1 ("mild") if their score was higher than 0 but $\le$ 0.5, 2 ("moderate") if their score was between 0.5 and one, and 3 ("severe") if their score was above 1. These labels roughly corresponded to score quartiles in ROSMAP.

Hippocampal sclerosis is recorded as a binary indicator of the presence or absence of pathology in the NACC NP data set form version 1-9, ROSMAP, and ACT. In versions 10 and 11 of the NACC NP form and in ADNI, HS pathology is recorded as being absent, unilateral, or bilateral. To harmonize, we dichotomized HS pathology as being present if either unilateral or bilateral pathology was indicated.

The Lewy body pathology variable we analyzed had four levels: 0, indicating absent Lewy body pathology in all regions examined or limited to the olfactory bulb; 1, indicating Lewy body pathology limited to the brainstem, including the substantia nigra; 2, indicating Lewy body pathology involving the limbic system or amygdala; and 3, indicating Lewy body involvement in the neocortex. In NACC and ROSMAP, Lewy body pathology was graded in ordinal variables with levels corresponding to the levels in the final harmonized outcome variable analyzed. In ADNI, a single variable was also used to record Lewy body pathology but included separate levels for (1) no pathology and olfactory-predominant and (2) limbic-predominant and amygdala-predominant pathology. These stages were collapsed into levels 0 and 2, respectively, for harmonization. In ACT, separate binary indicator variables were used to indicate presence of Lewy body pathology in each brain region checked. To harmonize, we created a new variable that coded Lewy body pathology stage according to the "highest" stage present in an individual (*e.g.* if pathology was present in both the amygdala and the neocortex, we assigned a value of 3).

\newpage

## Study participant demographics and distributions of endophenotypes

\newpage

## Supplementary Figure 1: Dendrogram of Neuropathology Endophenotype Polychoric Correlations

## Supplementary Figures 2-12: Manhattan plots of Stage 3 NPE GWAS

```{r}
# genes <- 
#   rtracklayer::import("~/downloads/gencode.v40.annotation.gtf.gz",
#                       feature.type = "gene") %>% 
#   as.data.table() %>% 
#   .[gene_type == "protein_coding"]
# genes <- genes[, .(gene_name, seqnames, start, end)] %>% 
#   melt(., id.vars = c("gene_name", "seqnames"), 
#        measure.vars = c("start", "end"))
# genes[, chr := as.integer(seqnames)]
```

```{r}
# bim <- fread("./data/mega/mega_np.bim")
# pheno_ord <- fread("data/mega/mega_np_ord.pheno")
# pheno_bin <- fread("data/mega/mega_np.pheno")
# ord_phenos <- colnames(pheno_ord)[3:ncol(pheno_ord)]
# bin_phenos <- colnames(pheno_bin)[3:ncol(pheno_bin)]
# results_files_ord <- paste0("output/gwas/mega/polmm/", 
#                             ord_phenos,
#                             "_polmm_results.txt")
# results_files_bin <- paste0("output/gwas/mega/saige/",
#                             bin_phenos,
#                             "_saige_results.txt")
# clump_files_ord <- paste0("output/gwas/mega/polmm/", 
#                           ord_phenos,
#                           "_polmm_results.clumped")
# clump_files_bin <- paste0("output/gwas/mega/saige/",
#                           bin_phenos,
#                           "_saige_results.clumped")
# 
# results_ord <- get_results_list(results_files_ord)
# results_bin <- get_results_list(results_files_bin)
# clump_ord <- get_results_list(clump_files_ord)
# clump_bin <- get_results_list(clump_files_bin)
# 
# for (i in 1:length(clump_bin)) {
#   clump_bin[[i]][, Gene := find_closest_gene(CHR, BP, genes)]
#   clump_bin[[i]][CHR == 19, Gene := "APOE"]
#   setorder(clump_bin[[i]], P)
#   clump_bin[[i]] <- clump_bin[[i]][!duplicated(Gene)]
#   setorder(clump_bin[[i]], CHR, BP)
# }
# 
# for (i in 1:length(clump_ord)) {
#   clump_ord[[i]][, Gene := find_closest_gene(CHR, BP, genes)]
#   clump_ord[[i]][CHR == 19, Gene := "APOE"]
#   setorder(clump_ord[[i]], P)
#   clump_ord[[i]] <- clump_ord[[i]][!duplicated(Gene)]
#   setorder(clump_ord[[i]], CHR, BP)
# }
# 
# for (i in 1:length(results_ord)) {
#   results_ord[[i]] <- merge(results_ord[[i]], 
#                             bim[, .(V2, V4)], 
#                             by.x = "SNPID", 
#                             by.y = "V2")
# }
# results <- fread("output/gwas/mega/polmm/late_polmm_results.txt")
# clump <- fread("output/gwas/mega/polmm/late_polmm_results.clumped")
# results <- merge(results, bim[, .(V2, V4)], by.x = "SNPID", by.y = "V2")
```

```{r}
# for (i in 1:length(results_ord)) {
#   manplot <- plot_manhattan(results_ord[[i]], 
#                           pos_col = "V4", 
#                           annotate = clump_ord[[i]][P < 5e-8]$SNP,
#                           annotate_labels = clump_ord[[i]][P < 5e-8]$Gene,
#                           highlight = TRUE,
#                           max_yval = 15) +
#     ylim(0, 15)
#   ggsave(filename = paste0("Stage3_manhattan_", ord_phenos[i], ".png"),
#          path = "./doc/",
#          width = 12,
#          height = 6,
#          units = "in")
# }
# 
# for (i in 1:length(results_bin)) {
#   manplot <- plot_manhattan(results_bin[[i]], 
#                           chr_col = "CHR",
#                            id_col = "MarkerID",
#                            pval_col = "p.value",
#                           annotate = clump_bin[[i]][P < 5e-8]$SNP,
#                           annotate_labels = clump_bin[[i]][P < 5e-8]$Gene,
#                           highlight = TRUE,
#                           max_yval = 15) +
#     ylim(0, 15)
#   ggsave(filename = paste0("Stage3_manhattan_", bin_phenos[i], ".png"),
#          path = "./doc/",
#          width = 12,
#          height = 6,
#          units = "in")
# }

```

## Sensitivity analyses for novel APOE-adjusted CAA risk locus

<!-- analyses in code/conditional/caa_apoe_stratified_analysis.R -->

We performed several sensitivity analyses to further investigate the novel CAA-associated locus on Chromosome 19 that we identified when adjusting for *APOE* $\epsilon$ diplotype.
First, for each of the NACC, ROSMAP, and ACT data sources, we re-analyzed the association between CAA and lead variant rs7247551 from the meta-analysis while stratifying by *APOE* $\epsilon$ diplotype and visually compared effect sizes across groups. Due to low sample sizes prevent model convergence, APOE $\epsilon 4$ carriers (diplotypes $\epsilon 3$) Then, for each data source, we performed an an association analysis between CAA and rs7247551 that included interaction effects between rs7247551 and *APOE* $\epsilon$ diplotype. We then performed ANOVA between these models and nested models which did not include interaction effects and performed Chi-Square tests to test whether any interaction terms were significant.

Results from stratified analyses are shown in Supplementary Figures 13-15 below. No interaction effects were significant in any of the data sources (NACC Chi-Square P = 0.84; ROSMAP Chi-Square P = 0.13; ACT Chi-Square P = 0.91). These results show that the association between rs7247551 and CAA does not significantly differ based on  *APOE* $\epsilon$ diplotype in any of the three data sources used.

### 2024-04-30 code/SAIGE/saige_analysis_act_cerad_adj.sh

@@ -1,0 +1,61 @@
+#==================================================================
+# Run SAIGE GWAS for binary variables: hs, microinf, and grossinf
+#==================================================================
+study=act
+mkdir ./output/gwas/$study/saige/
+
+pheno_file=tmp/"$study"_cerad_adj_saige.pheno
+Rscript --vanilla ./code/SAIGE/make_pheno_file.R \
+  --pheno data/act/"$study"_np.pheno \
+  --covar data/act/"$study"_np.covar \
+  --out $pheno_file
+
+covars=$(cat $pheno_file | \
+         cut -d " " -f 4,14- | \
+         awk 'BEGIN {FS=" "; OFS = ",";} NR==1 {$1=$1; print $0}')
+
+cat_covars=$(cat $pheno_file | \
+         cut -d " " -f 4,14,16,17 | \
+         awk 'BEGIN {FS=" "; OFS = ",";} NR==1 {$1=$1; print $0}')
+
+phenos=$(cat $pheno_file | \
+         cut -d " " -f 11,12,13 | \
+         awk 'NR==1 {print $0}')
+         
+
+
+for pheno in grossinf microinf
+  do
+  Rscript --vanilla ./code/SAIGE/step1_fitNULLGLMM.R \
+    --plinkFile=./data/act/act_np_pruned \
+    --phenoFile="$pheno_file" \
+    --phenoCol="$pheno" \
+    --covarColList="$covars" \
+    --qCovarColList="$cat_covars" \
+    --sampleIDColinphenoFile=IID \
+    --nThreads=100 \
+    --outputPrefix=./output/gwas/"$study"/saige/"$pheno"_cerad_adj_saige \
+    --IsOverwriteVarianceRatioFile=TRUE
+
+  for chr in $(seq 1 22)
+    do
+    plink=./data/act/plink/act_np_chr"$chr"
+    Rscript --vanilla ./code/SAIGE/step2_SPAtests.R \
+      --bedFile="$plink".bed \
+      --bimFile="$plink".bim \
+      --famFile="$plink".fam \
+      --AlleleOrder=ref-first \
+      --SAIGEOutputFile=./output/gwas/"$study"/saige/"$pheno"_cerad_adj_saige_results_chr"$chr".txt \
+      --chrom="$chr" \
+      --GMMATmodelFile=./output/gwas/"$study"/saige/"$pheno"_cerad_adj_saige.rda \
[…]

### 2024-04-30 code/aux/removed_before_pub/caa_apoe_mediation.R

@@ -1,0 +1,22 @@
+library(data.table)
+library(magrittr)
+library(MASS)
+library(mediation)
+library(ordinal)
+
+pheno <- setDT(readRDS('data/mega/mega_np.Rds'))[
+  , .(FID, IID, caa_ord)
+]
+
+covar <- fread('data/mega/mega_np_apoe.covar')
+caa_snp <- fread("data/mega/conditional/caa_apoe_snp.raw")
+caa_snpid <- "rs4803779_C"
+covars <- colnames(covar)[3:ncol(covar)]
+dt <- Reduce(
+  merge, 
+  list(pheno, 
+       covar, 
+       caa_snp[, .(FID, IID, 'snp' = get(..caa_snpid))]
+  )
+)
+

### 2024-04-30 code/aux/removed_before_pub/heritability/format_sumstats_plink2.R

@@ -1,0 +1,96 @@
+#!/usr/bin/Rscript --vanilla
+
+#=====================================
+# Format plink .assoc.logistic file
+# into LDAK summary statistics format
+# Predictor A1 A2 Z
+#=====================================
+
+library(data.table)
+library(magrittr)
+library(stringi)
+library(rtracklayer)
+library(argparse)
+
+parser <- ArgumentParser()
+parser$add_argument("--sumstats")
+parser$add_argument("--N")
+parser$add_argument("--out")
+parser$add_argument("--build", default="hg38")
+parser$add_argument("--tags", default="raw_data/ldak.thin.hapmap.gbr.tagging",
+                    help="Tagging file for LDAK heritability model")
+args <- parser$parse_args()
+
+results_path <- args$sumstats
+out_path <- args$out
+tags_path <- args$tags
+
+if (args$build %in% c("hg19", "hg38")) {
+  build <- args$build
+  print(paste("Variant positions are assumed to map to build:", build))
+  if (build == "hg38") {
+    print("Variants will be remapped to build hg19.")
+  }
+} else {
+  stop("Build must be either: hg19, hg38")
+}
+
+results <- fread(results_path)
+# results[Allele1 != ALT, T_STAT := -T_STAT]
+results[, Z := Effect / StdErr]
+results <- results[, .(`chr`, bp, MarkerName, Allele2, Allele1, OBS_CT, Z)]
+setnames(results, colnames(results), c('CHR', 'BP', 'SNP', 'A2', 'A1', 'n', 'Z'))
+
+
+
+# LDAK only accepts SNPs for genetic correlation
+snp_alleles <- c("A", "C", "T", "G")
+sumstats <- results[A1 %in% snp_alleles & A2 %in% snp_alleles]
+sumstats[, ChrBP := paste0(CHR, ":", BP)]
+sumstats <- sumstats[!(duplicated(ChrBP))]
+sumstats[, seq_name := paste0("chr", CHR)]
+
+# convert to hg19 build if current build is hg38
+if (build == "hg38") {
+  setnames(sumstats, "BP", "BP_hg38", skip_absent = TRUE)
+  
+  # commented out because of issues installing GEOquery
+  # copy over chain file
+  if (!file.exists("raw_data/hg38ToHg19.over.chain")) {
[…]

### 2024-04-30 code/aux/removed_before_pub/mcbb/regression.R

@@ -1,0 +1,17 @@
+
+library(data.table)
+library(ordinal)
+mcbb <- fread("data/mcbb/MayoRNAseq_individual_metadata_031422.csv")
+snp <- fread("data/mcbb/tmem106b.raw", check.names = TRUE)
+rnaseq <- fread("data/mcbb/rnaseq_gene_counts_normalized_tidy.csv")
+tmem106b_col <- colnames(rnaseq)[grep("ENSG00000106460", colnames(rnaseq))]
+rnaseq_cols <- c("specimenID", tmem106b_col) 
+dt <- merge(mcbb, snp, by.x = "individualID", by.y = "IID")
+dt <- merge(dt, rnaseq[, ..rnaseq_cols], by.x = "individualID", by.y = "specimenID")
+dt[, Braak_rd := as.ordered(floor(Braak))]
+dt[ageDeath == "90_or_over", ageDeath := "90"]
+dt[, ageDeath := as.numeric(ageDeath)]
+m <- dt[race == "White", clm(as.ordered(Braak_rd) ~ X7.12269593_A + sex + ageDeath + factor(apoeGenotype))]
+summary(m)
+
+m1 <- dt[race == "White", clm(as.ordered(Braak_rd) ~ log(ENSG00000106460) + sex + ageDeath +pmi)]

### 2024-04-30 code/aux/removed_before_pub/mega/02_make_pheno.R

@@ -1,0 +1,232 @@
+library(data.table)
+library(stringi)
+library(readxl)
+
+adc <- fread("data/adc/adc_np.pheno", na.strings = '-1')
+ros <- fread("data/rosmap/rosmap_np.pheno", na.strings = '-1')
+act <- fread("data/act/act_np.pheno", na.strings = '-1')
+adni <- fread("data/adni/adni_np.pheno", na.strings = '-1')
+fam <- fread("data/mega/mega_np.fam")
+
+act_np <- as.data.table(read_xlsx("raw_data/ACT_235_provided_2023_06_22/np.xlsx",
+                                  sheet = 1))
+adc_np <- setDT(readRDS("data/adc/np_qced.Rds")) # %>%
+  # replace_with_na_all(condition = ~.x == -1)
+setDT(adc_np)
+ros_np <- setDT(read_xlsx(
+  "/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx"))
+adni_np <- fread("raw_data/NEUROPATH_07_06_21.csv")
+adni_fam <- fread("data/adni/adni_np.fam")
+adni_np <- merge(adni_fam, adni_np, by.x = 'V2', by.y = 'RID')
+rm(adni_fam)
+
+
+set_ids_as_char <- function(dt) {
+  dt[, IID := as.character(IID)]
+  dt[, FID := as.character(FID)]
+}
+
+set_ids_as_char(adc)
+set_ids_as_char(ros)
+set_ids_as_char(act)
+set_ids_as_char(adni)
+
+ros_np[, FID := '1']
+act_np[, FID := '2']
+setnames(ros_np, 'projid', 'IID')
+setnames(act_np, 'IDfromDave', 'IID')
+setnames(adni_np, c('V1', 'V2'), c('FID', 'IID'))
+set_ids_as_char(adc_np)
+set_ids_as_char(ros_np)
+set_ids_as_char(act_np)
+set_ids_as_char(adni_np)
+
+
+
+
+# rename ADC, ROSMAP, and ACT phenotypes to match those in ADNI
+# (since I renamed those myself already)
+phenotypes <- colnames(adni)[3:length(colnames(adni))]
+setnames(adc, c('LEWY', 'ATHCW', 'NACCINF', 'NACCMICR', 'HS', 'BRAAK', 
+                'NEUR', 'CAA', 'ASC'), 
+         phenotypes)
+setnames(ros, 
[…]

### 2024-04-30 code/coloc/coloc_ordinal_phenos.py

@@ -1,26 +1,0 @@
-# run in coloc directory
-import subprocess
-import os
-import pandas as pd
-
-pheno_file = "../np_phewas/data/mega/mega_np_ord_dichotomized.pheno"
-pheno_data = pd.read_csv(pheno_file)
-phenos = pheno_data.columns[2:]
-
-for pheno in phenos:
-    cmd = [
-        "python", "./src/coloc_qtl.py",
-        "--out_pref", "../np_phewas/output/coloc/mega/" + pheno + "/",
-        "--gwas_sumstats", 
-        "../np_phewas/output/gwas/mega/polmm/" + pheno + "_polmm_results.txt",
-        "--max_pval", "1e-5",
-        "--pval_col", "pval.spa",
-        "--phenotype_type", "cc",
-        "--phenotype_col", pheno,
-        "--phenotype_file", pheno_file,
-        "--phenotype_na_string", "NA", 
-        "--gtex", 
-        "--rosmap"
-    ]
-    
-    subprocess.Popen(cmd)

### 2024-04-30 code/coloc/coloc_results.R

@@ -2,18 +2,18 @@ library(data.table)
 library(rtracklayer)
 library(stringi)
 # colocalization results
-pheno_ord <- fread("data/mega/mega_np_ord.pheno")
-pheno_bin <- fread("data/mega/mega_np.pheno")
-pheno <- merge(pheno_ord, pheno_bin, c("FID", "IID"))
+pheno <- fread("shared/nacc_rosmap_act_np.pheno")
 phenos <- colnames(pheno)[3:ncol(pheno)]
 coloc_list <- vector(mode = "list", length = length(phenos))
 names(coloc_list) <- phenos
 for (i in 1:length(phenos)) {
   p <- phenos[i]
-  coloc_path <- paste0("output/coloc/mega/", p, "/coloc_results.txt")
+  coloc_path <- paste0("output/coloc/meta/", p, "/coloc_results.txt")
   coloc_list[[i]] <- fread(coloc_path)
 }
 
+coloc_list[[length(coloc_list) + 1]] = fread("output/coloc/meta/caa_apoe/coloc_results.txt")
+names(coloc_list)[[length(coloc_list)]] = "caa_apoe_adj"
 coloc <- rbindlist(coloc_list)
 coloc[, gene_id := stri_replace_all_regex(QTL_Phenotype, '.*:', '')]
 coloc[, gene_id := stri_replace_all_regex(gene_id, '\\..*', '')]
@@ -32,5 +32,5 @@ coloc <- merge(coloc, coloc_genes, by = 'gene_id', all.x = TRUE)
 coloc[is.na(gene_name), gene_name := QTL_Phenotype]
 setnames(coloc, "gene_name", "QTL_Target")
 
-fwrite(coloc, "output/coloc/mega/coloc_results.txt")
-fwrite(coloc[PPH4 > 0.8], "output/coloc/mega/coloc_results_pph4_80.csv")
+fwrite(coloc, "output/coloc/meta/coloc_results.txt")
+fwrite(coloc[PPH4 > 0.8], "output/coloc/meta/coloc_results_pph4_80.csv")

### 2024-04-30 code/conditional/apoe_haplotyping.R

@@ -6,7 +6,10 @@ nacc <- fread('data/adc/apoe_snps.raw')
 rosmap <- fread('data/rosmap/apoe_snps.raw')
 act <- fread('data/act/apoe_snps.raw')
 adni <- fread('data/adni/apoe_snps.raw')
-covar <- fread('data/mega/mega_np.covar')
+# covar <- fread('data/mega/mega_np.covar')
+nacc_covar = fread("data/adc/adc_np.covar")
+rosmap_covar = fread("data/rosmap/rosmap_np.covar")
+act_covar = fread("data/act/act_np.covar")
 
 nacc[, FID := as.character(FID)]
 nacc[, FID := '0']
@@ -20,9 +23,6 @@ act[, FID := as.character(FID)]
 act[, FID := '2']
 act[, IID := stri_replace_first_regex(IID, '.*_', '')]
 setnames(act, colnames(act)[7:8], c('rs429358_C', 'rs7412_T'))
-
-adni[, FID := as.character(FID)]
-adni[, FID := 'ADNI']
 
 apoe <- rbindlist(list(nacc, rosmap, act, adni), use.names = TRUE)
 apoe[, `:=`(PAT = NULL, MAT = NULL, SEX = NULL, PHENOTYPE = NULL)]
@@ -47,7 +47,24 @@ apoe <- apoe[!(FID == '0' & IID %in% IID[duplicated(IID)])]
 apoe <- apoe[apoe != 'apoe14']
 apoe[, `:=`(rs429358_C = NULL, rs7412_T = NULL)]
 
-apoe_covar <- merge(covar, apoe, c('FID', 'IID'))
+nacc_covar[, FID := as.character(FID)]
+nacc_covar[, IID := as.character(IID)]
+nacc_covar = merge(nacc_covar, apoe, c('FID', 'IID'))
 
-fwrite(apoe_covar, file = "data/mega/mega_np_apoe.covar", 
+rosmap_covar[, FID := as.character(FID)]
+rosmap_covar[, IID := as.character(IID)]
+rosmap_covar = merge(rosmap_covar, apoe, c('FID', 'IID'))
+
+act_covar[, FID := as.character(FID)]
+act_covar[, IID := as.character(IID)]
+act_covar = merge(act_covar, apoe, c('FID', 'IID'))
+act_covar = act_covar[apoe != "apoe22"] # only 3 individuals
+
[…]

### 2024-04-30 code/conditional/tmem.R

@@ -1,68 +1,0 @@
-
-library(data.table)
-library(magrittr)
-library(MASS)
-library(mediation)
-
-pheno <- setDT(readRDS('data/mega/mega_np.Rds'))[
-  , .(FID, IID, hs, late, late23)
-]
-covar <- fread('data/mega/mega_np.covar')
-related_rm <- fread('data/mega/related_rm/hs.remove')
-tmem_snps <- fread('data/mega/conditional/tmem106b.raw')
-tmem_top_late_snp <- 'rs1060700'
-tmem_top_snp_col <- colnames(tmem_snps)[grep(tmem_top_late_snp, colnames(tmem_snps))]
-
-covars <- colnames(covar)[3:ncol(covar)] %>% 
-  .[!(. %in% c('act', 'ACT2', 'ACT3', 'adni'))]
-
-dt <- Reduce(merge, 
-             list(pheno, 
-                  covar, 
-                  tmem_snps[, .(FID, IID, 'snp' = get(..tmem_top_snp_col))]))
-
-dt <- dt[!(IID %in% related_rm$V2)] %>% 
-  .[complete.cases(.)]
-
-# binary mediation analysis
-f1 <- formula(paste0(c('hs ~ snp', covars), collapse = ' + '))
-f2 <- formula(paste0(c('late23 ~ snp', covars), collapse = ' + '))
-f3 <- formula(paste0(c('hs ~ snp', 'late23', covars), collapse = ' + '))
-iv_to_dv <- dt[, glm(f1, data = .SD, family = binomial())]
-iv_to_m <- dt[, glm(f2, data = .SD, family = binomial())]
-iv_m_to_dv <- dt[, glm(f3, data = .SD, family = binomial())]
-
-mr_binary <- mediate(iv_to_m, 
-                     iv_m_to_dv, 
-                     mediator = 'late23', 
-                     treat = 'snp', 
-                     boot = TRUE, 
-                     sims = 100)
-
-# ordinal late mediation analysis
-dt[, late_ord := ordered(late)]
-f4 <- formula(paste0(c('late_ord ~ snp', covars), collapse = ' + '))
-f5 <- formula(paste0(c('hs ~ snp', 'late_ord', covars), collapse = ' + '))
-iv_to_m <- dt[, polr(f4, data = .SD, Hess = TRUE)]
-iv_m_to_dv <- dt[, glm(f5, data = .SD, family = binomial())]
-
-mr_ord <- mediate(iv_to_m, 
-                     iv_m_to_dv, 
[…]

### 2024-04-30 code/functions/make_or_95_ci.R

@@ -35,7 +35,7 @@ make_or_95_ci <- Vectorize(make_or_95_ci,
 make_95CI <- function(or, p, ci=0.95) {
   beta <- log(or)
   lt_zero <- beta < 0
-  z <- qnorm(p * 2, lower.tail = lt_zero)
+  z <- qnorm(p / 2, lower.tail = lt_zero)
   sd <- beta / z
   zrad <- qnorm((1 - ci) / 2, lower.tail = FALSE)
   upper <- round(exp(beta + zrad * sd), 2)
@@ -45,3 +45,12 @@ make_95CI <- function(or, p, ci=0.95) {
 }
 
 make_95CI <- Vectorize(make_95CI)
+
+estimate_se_from_beta_and_p <- function(beta, p) {
+  if (p <= 0 | p >= 1) {return(NA)}
+  z <- qnorm(p / 2, lower.tail = FALSE)
+  sd <- abs(beta) / z
+  return(sd)
+}
+
+estimate_se_from_beta_and_p = Vectorize(estimate_se_from_beta_and_p)

### 2024-04-30 code/gwas/format_metal_results.R

@@ -1,0 +1,90 @@
+# format METAL results for downstream work
+
+#!/usr/bin/env Rscript --vanilla
+
+# Load the required packages
+library(data.table)
+library(stringi)
+library(argparse)
+
+# function for flipping direction of effect in Direction column (e.g. "++-" -> "--+")
+flip_effect_directions_fast <- function(directions_vector) {
+  # Temporarily replace "+" with a placeholder ("p")
+  temp_vector <- stri_replace_all_regex(directions_vector, "\\+", "p")
+  
+  # Replace "-" with "+"
+  temp_vector <- stri_replace_all_regex(temp_vector, "-", "\\+")
+  
+  # Replace placeholder ("p") with "-"
+  flipped_vector <- stri_replace_all_regex(temp_vector, "p", "-")
+  
+  return(flipped_vector)
+}
+
+# function for flipping alleles, direction of effect, allele frequency, and direction for data table
+flip_alleles_effect_direction <- function(dt) {
+  setnames(merged_dt, c("Allele1", "Allele2"), c("Allele2", "Allele1"))
+  dt[, Freq1 := 1 - Freq1]
+  dt[, Effect := -Effect]
+  dt[, MinFreq1 := 1 - MaxFreq]
+  dt[, MaxFreq := 1 - MinFreq]
+  dt[, MinFreq := NULL]
+  setnames(dt, "MinFreq1", "MinFreq")
+  dt[, Direction := flip_effect_directions_fast(Direction)]
+  return(dt)
+}
+# Define parser for command-line arguments
+parser <- ArgumentParser(description = "Merge METAL meta-analysis GWAS file with PLINK .bim file")
+
+# Add arguments to the parser
+parser$add_argument("--metal_file", required = TRUE, help = "Path to METAL meta-analysis GWAS file")
+parser$add_argument("--bim_file", required = TRUE, help = "Path to PLINK .bim file")
+parser$add_argument("--output_file", required = TRUE, help = "Path to output merged file")
+parser$add_argument("--flip_alleles", default = "FALSE", help = "Flip alleles and effect sizes? Default = FALSE")
+
+# Parse the command-line arguments
+args <- parser$parse_args()
+args$flip_alleles = as.logical(args$flip_alleles)
+
+# Read METAL meta-analysis GWAS file into a data.table
[…]

### 2024-04-30 code/gwas/meta/metal_pipeline.sh

@@ -1,0 +1,26 @@
+
+meta_dir=./code/gwas/meta
+
+# bash "$meta_dir"/prep_polmm_results_for_metal.sh
+# bash "$meta_dir"/prep_saige_results_for_metal.sh
+
+# for pheno in hs grossinf microinf microinf grossinf cerad braak caa arteriol atheroscler lewy_body late braak_apoe cerad_apoe late_apoe caa_apoe
+#   do
+#   metal "$meta_dir"/"$pheno"_metal.txt
+#   done
+
+# for pheno in braak cerad caa hs microinf grossinf late lewy_body arteriol atheroscler diffuse_abeta
+#   do
+#   Rscript --vanilla code/gwas/format_metal_results.R \
+#     --metal_file output/gwas/metal/results/"$pheno"1.tbl \
+#     --bim_file data/adc/adc_np.bim \
+#     --output_file output/gwas/metal/results/"$pheno"1.csv 
+#   done
+
+for pheno in diffuse_abeta
+do
+python ./code/gwas/clump_snps.py \
+  -b data/adc/adc_np \
+  -r tmp/"$pheno"_magma_input.csv \
+  -o output/gwas/metal/results/"$pheno"1_clump
+done

### 2024-04-30 code/gwas/ordinal/act_diffuse_abeta_polmm.R

@@ -1,0 +1,45 @@
+library(GRAB)
+library(data.table)
+library(stringi)
+source("code/functions/strip_alleles.R")
+PhenoData = fread("data/act/act_np.pheno", na.strings = "-1")
+covar_data = fread("data/act/act_np.covar")
+PhenoData = merge(PhenoData, covar_data, c("FID", "IID"))
+PhenoData[, diffuse_abeta := factor(diffuse_abeta, levels = c(0, 1, 2, 3))]
+GenoFile = "data/act/act_np_pruned.bed"
+getTempFilesFullGRM("data/act/act_np_pruned", 
+                      nPartsGRM = 1, 
+                      partParallel = 1)
+
+getSparseGRM("data/act/act_np_pruned", 
+             nPartsGRM = 1,
+             SparseGRMFile = "tmp/sparse_grm.txt",
+             relatednessCutoff = 0.01)
+
+obj.POLMM = GRAB.NullModel(formula = diffuse_abeta ~ msex + age_death + ACT2 + ACT3 + pc1 + pc2 +pc3 + pc4 + pc5 + pc6 + pc7 + pc8 + pc9 + pc10,
+                           data = PhenoData, 
+                           subjData = PhenoData$IID, 
+                           method = "POLMM", 
+                           traitType = "ordinal",
+                           GenoFile = GenoFile,
+                           SparseGRMFile =  "tmp/sparse_grm.txt",
+                           control = list(showInfo = FALSE, 
+                                          LOCO = FALSE))
+
+gwas_plink = "data/act/act_np.bed"
+OutputFile = "output/gwas/act/polmm/diffuse_abeta_polmm_results.txt"
+GRAB.Marker(
+  obj.POLMM, 
+  GenoFile = gwas_plink,
+  OutputFile = OutputFile
+)
+results = fread("output/gwas/act/polmm/diffuse_abeta_polmm_results.txt")
+snp_info = stri_split_fixed(results$Info, ":", simplify = T)
+results[, chr := snp_info[, 1]]
+results[, A2 := snp_info[, 3]]
+results[, A1 := snp_info[, 4]]
+results[, N := obj.POLMM$N]
+setnames(results, c("Marker", "beta", "seBeta", "Pvalue", "AltFreq"), c("SNPID", "beta", "SE", "pval.spa", "MAF"))
+results = results[!is.na(pval.spa)]
[…]

### 2024-04-30 code/gwas/ordinal/polmm_rosmap.sh

@@ -1,29 +1,26 @@
 
 mkdir output/gwas/rosmap/polmm
 
+# create SNPMatrix
+mkdir data/rosmap/SNPMatrix
+./code/gwas/ordinal/polmm_create_SnpMatrix.R \
+  --bim data/rosmap/rosmap_np.bim \
+  --out data/rosmap/SNPMatrix/
+
+
 phenos=$(cat data/mega/rosmap_np_ord.pheno | \
          cut -d " " -f 3- | \
          awk 'NR == 1{print}')
 
-for pheno in $phenos
+
+for pheno in cerad diffuse_abeta
   do
-  # output top clumped SNPs from NACC analysis to SnpMatrix
-  snps=tmp/"$pheno"_nacc_snps.txt
-  cat output/gwas/adc/polmm/"$pheno"_polmm_results.clumped | \
-    awk 'NR>1{print $3}' \
-    > $snps
-
-  snpmat_prefix=data/mega/SnpMatrix/nacc_"$pheno"
-  ./code/gwas/ordinal/polmm_create_SnpMatrix.R \
-    --snp_list $snps \
-    -o $snpmat_prefix
-
   # create null model
   null_model=output/gwas/rosmap/polmm/"$pheno"_null_model.Rds
   ./code/gwas/ordinal/polmm_null_model.R \
-    --covar data/mega/rosmap_np.covar \
-    --pheno data/mega/rosmap_np_ord.pheno \
-    --plink tmp/mega_np_prune \
+    --covar data/rosmap/rosmap_np.covar \
+    --pheno data/rosmap/rosmap_np.pheno \
+    --plink data/rosmap/rosmap_np_pruned \
     --phenotype $pheno \
     --out $null_model \
     --ncores 100
@@ -32,10 +29,41 @@ for pheno in $phenos
   results=output/gwas/rosmap/polmm/"$pheno"_polmm_results.Rds
   ./code/gwas/ordinal/polmm_regression.R \
     -m $null_model \
-    --snp_matrix "$snpmat_prefix"SnpMatrix.Rds \
+    -l data/rosmap/SNPMatrix/SnpMatrix_list.Rds \
     -o $results
 
   Rscript --vanilla ./code/gwas/rds_to_text.R \
   -r $results \
   -o output/gwas/rosmap/polmm/"$pheno"_polmm_results.txt
+
+  python ./code/gwas/clump_snps.py \
[…]

### 2024-04-30 code/heritability/format_sumstats_plink2.R

@@ -1,94 +1,0 @@
-#!/usr/bin/Rscript --vanilla
-
-#=====================================
-# Format plink .assoc.logistic file
-# into LDAK summary statistics format
-# Predictor A1 A2 Z
-#=====================================
-
-library(data.table)
-library(magrittr)
-library(stringi)
-library(rtracklayer)
-library(argparse)
-
-parser <- ArgumentParser()
-parser$add_argument("--sumstats")
-parser$add_argument("--out")
-parser$add_argument("--build", default="hg38")
-parser$add_argument("--tags", default="raw_data/ldak.thin.hapmap.gbr.tagging",
-                    help="Tagging file for LDAK heritability model")
-args <- parser$parse_args()
-
-results_path <- args$sumstats
-out_path <- args$out
-tags_path <- args$tags
-
-if (args$build %in% c("hg19", "hg38")) {
-  build <- args$build
-  print(paste("Variant positions are assumed to map to build:", build))
-  if (build == "hg38") {
-    print("Variants will be remapped to build hg19.")
-  }
-} else {
-  stop("Build must be either: hg19, hg38")
-}
-
-results <- fread(results_path)
-results[A1 != ALT, T_STAT := -T_STAT]
-results <- results[, .(`#CHROM`, POS, ID, REF, ALT, OBS_CT, T_STAT)]
-setnames(results, colnames(results), c('CHR', 'BP', 'SNP', 'A2', 'A1', 'n', 'Z'))
-
-
-
-# LDAK only accepts SNPs for genetic correlation
-snp_alleles <- c("A", "C", "T", "G")
-sumstats <- results[A1 %in% snp_alleles & A2 %in% snp_alleles]
-sumstats[, ChrBP := paste0(CHR, ":", BP)]
-sumstats <- sumstats[!(duplicated(ChrBP))]
-sumstats[, seq_name := paste0("chr", CHR)]
-
-# convert to hg19 build if current build is hg38
-if (build == "hg38") {
-  setnames(sumstats, "BP", "BP_hg38", skip_absent = TRUE)
-  
-  # commented out because of issues installing GEOquery
-  # copy over chain file
-  if (!file.exists("raw_data/hg38ToHg19.over.chain")) {
-    file.copy("/data_global/liftOver/hg38ToHg19.over.chain.gz",
[…]

### 2024-04-30 code/magma/extract_markername_pvalue.R

@@ -1,0 +1,33 @@
+#!/usr/bin/env Rscript
+
+# Load necessary libraries
+library(argparse)
+library(data.table)
+
+# Setting up Argument Parser
+parser <- ArgumentParser()
+
+# Adding arguments
+parser$add_argument("-i", "--input", type = "character", help = "Input CSV file name")
+parser$add_argument("-s", "--snp_column", type = "character", default = "MarkerName", help = "Column name for SNP, default is 'MarkerName'")
+parser$add_argument("-p", "--pvalue_column", type = "character", default = "P.value", help = "Column name for P-value, default is 'P.value'")
+parser$add_argument("-o", "--output", type = "character", help = "Output CSV file name")
+
+# Parse the arguments
+args <- parser$parse_args()
+
+# Read the input file
+data <- fread(args$input)
+
+# Check if the SNP and P-value columns exist
+if (!(args$snp_column %in% colnames(data)) || !(args$pvalue_column %in% colnames(data))) {
+    stop("Specified columns not found in the data")
+}
+
+# Extract SNP and P-value columns
+output_data <- data[, .(SNP=get(..args$snp_column), P=get(..args$pvalue_column))]
+
+# Write the extracted data to a new CSV file
+fwrite(output_data, file = args$output, sep = " ")
+
+cat("File saved:", args$output, "\n")

### 2024-04-30 code/manuscript/figure_S1_dendrogram.R

@@ -6,18 +6,16 @@ library(data.table)
 library(magrittr)
 library(psych)
 library(pheatmap)
+library(polycor)
 
-bin_pheno <- fread("data/mega/mega_np.pheno", na.strings = "-1")
-ord_pheno <- fread("data/mega/mega_np_ord.pheno", na.strings = "-1")
-pheno <- merge(ord_pheno, bin_pheno[, .(FID, IID, hs, grossinf, microinf)],
-               by = c('FID', 'IID'))
+pheno <- fread("shared/nacc_rosmap_act_np.pheno", na.strings = "-1")
 covar <- fread("data/mega/mega_np.covar")
 
-# PCA on all variables except LATE-NC (has a lot of missingess)
-pheno_pca <- pheno[, -c("late", "FID", "IID")]
-labels = c("Braak NFT Stage", "CERAD Score", "Diffuse Amyloid Plaques", 
-           "Arteriolosclerosis",  "Atherosclerosis", "Lewy Body", 
-           "CAA", "HS", "Gross Infarction", "Microinfarct")
+# PCA on all variables
+pheno_pca <- pheno[, -c("FID", "IID")]
+labels = c("LATE-NC", "Braak NFT Stage", "CERAD Score", "Amyloid-Beta Plaques", 
+           "Arteriolosclerosis","Atherosclerosis", "Lewy Body", 
+           "CAA", "HS", "Microinfarct", "Gross Infarction")
 setnames(pheno_pca, colnames(pheno_pca), labels)
 
 pca <- principal(pheno_pca, 
@@ -29,12 +27,33 @@ pca <- principal(pheno_pca,
 # Two major clusters found corresponding roughly to:
 #   1. Alzheimer's NP : Braak, CERAD, diffuse amyloid, and CAA
 #   2. Vascular NP    : gross/micro infarcts, athero/arteriolosclerosis
-
-dendrogram <- pheatmap(polychoric(pheno_pca)$rho, 
+cors <- polychoric(pheno_pca)
+dendrogram <- pheatmap(cors$rho, 
                        # labels_row = labels,
                        # labels_col = rep("", 10),
                        filename = "./doc/figure_S1_dendrogram.png",
                        height = 7,
-                       width = 8)
+                       width = 8, 
+                       angle_col = "45")
 
 dendrogram
+
[…]

### 2024-04-30 code/manuscript/locuszoom_plots.sh

@@ -1,0 +1,43 @@
+# cat output/gwas/mega/polmm/atheroscler_polmm_results.txt | \
+#   ~/software/locuszoom/bin/locuszoom \
+#     --metal - \
+#     --markercol SNPID \
+#     --pvalcol pval.spa \
+#     --chr 13 \
+#     --refsnp rs2000660 \
+#     --flank 500000 \
+#     --source 1000G_March2002 \
+#     --build hg19 \
+#     --pop EUR \
+#     --delim space \
+#     --no-date \
+#     ymax=10 \
+#     legend="none" \
+#     geneFontSize=1.18 \
+#     axisTextSize=1.4 \
+#     refsnpTextSize=1.2 \
+#     showRecomb=FALSE \
+#     rfrows=2	\
+#     --prefix doc/ 
+
+cat "output/gwas/metal/results/caa_apoe1.csv" | \
+  ~/software/locuszoom/bin/locuszoom \
+    --metal - \
+    --markercol MarkerName \
+    --pvalcol P.value \
+    --chr 19 \
+    --refsnp rs4803778 \
+    --flank 500000 \
+    --source 1000G_March2012 \
+    --build hg19 \
+    --pop EUR \
+    --delim space \
+    --no-date \
+    ymax=13 \
+    legend="none" \
+    geneFontSize=1 \
+    axisTextSize=1 \
+    refsnpTextSize=1.2 \
+    showRecomb=FALSE \
+    signifLine="7.3" \
+    --prefix doc/

### 2024-04-30 code/manuscript/supplementary_caa_forest_plots.R

@@ -1,0 +1,121 @@
+library(forestplot)
+
+#list.files(path = "/home/kau232/locuszoom_test/lincoln/Forest_plot/")
+
+cil = function(beta, se) {
+  cil = exp(beta - 1.96 * se)
+}
+
+ciu = function(beta, se) {
+  ciu = exp(beta + 1.96 * se)
+}
+# NACC 
+# apoe33 -0.163  0.056
+# apoe34 -0.248 0.055
+# apoe44 -0.190112   0.103950
+# apoe24 -0.56 0.246
+# apoe23 -0.4 0.172
+# Create a data frame for forestplot
+forest_data <- data.frame(
+  labeltext = c("APOE 3/3", "APOE 3/4", "APOE 4/4", "APOE 2/4", "APOE 2/3"),
+  mean = c(exp(-0.163), exp(-0.248), exp(-0.19), exp(-0.56), exp(-0.4)),
+  lower = c(cil(-0.163, 0.056), cil(-0.248, 0.055), cil(-0.190112, 0.103950), cil(-0.56, 0.246), cil(-0.4, 0.172)),
+  upper = c(ciu(-0.163, 0.056), ciu(-0.248, 0.055), ciu(-0.190112, 0.103950), ciu(-0.56, 0.246), ciu(-0.4, 0.172))
+)
+
+# Create a forest plot with labels
+forestplot(
+  mean = forest_data$mean,
+  lower = forest_data$lower,
+  upper = forest_data$upper,
+  labeltext = forest_data$labeltext,
+  title = "NACC: Forest Plot of APOE-stratified rs7247551 assocation with CAA",
+  xlab = "Odds Ratio with 95% CI",
+  col = fpColors(box = "blue", line = "darkblue", summary = "royalblue"),
+  legend.pos = "bottom"  # Adjust legend position
+)
+
+# ROSMAP 
+# apoe33 -0.23656    0.10231
+# apoe34_44_24 -0.38474    0.15768
+# apoe23 -0.42770    0.28593
+# Create a data frame for forestplot
+forest_data <- data.frame(
+  labeltext = c("APOE 3/3", "APOE 4 carrier", "APOE 2/3"),
+  mean = c(exp(-0.23656), exp(-0.38474), exp(-0.42770)),
+  lower = c(cil(-0.23656, 0.10231), cil(-0.38474, 0.15768), cil(-0.42770, 0.28593)),
[…]

### 2024-04-30 code/manuscript/supplementary_table_AD_GWAS_results.R

@@ -1,0 +1,19 @@
+
+library(data.table)
+ad_gwas <- 
+  fread("/data_global/summary_statistics/GCST90027158_buildGRCh38.tsv.gz")
+
+top_snps <- c("rs6733839", "rs6460900", "rs4721058", "rs2000660", "rs72807981",
+              "rs769449", "rs12721051", "rs4420638", "rs4803778"
+)
+
+ad_gwas_top_snps <- ad_gwas[variant_id %in% top_snps]
+
+jansen_2019 <- fread("/data_global/summary_statistics/jansenIE2019_AD/AD_sumstats_Jansenetal_2019sept.txt.gz")
+
+jansen_2019_top_snps <- jansen_2019[SNP %in% top_snps]
+jansen_2019_top_snps[, chr_bp := paste0(CHR, "_", BP)]
+
+wightman <- fread("/data_global/summary_statistics/wightman2021_AD/PGCALZ2sumstatsExcluding23andMe.txt.gz")
+wightman[, CHR_BP := paste0(chr, "_", PosGRCh37)]
+wightman_top_snps <- wightman[CHR_BP %in% jansen_2019_top_snps$chr_bp]

### 2024-04-30 code/manuscript/supplementary_table_nacc_analysis_results.R

@@ -1,0 +1,281 @@
+#==========================================
+# Produce Table 2: GWAS results summary
+#==========================================
+
+library(data.table)
+library(magrittr)
+library(rtracklayer)
+library(flextable)
+source("code/functions/find_closest_gene.R")
+source("code/functions/make_or_95_ci.R")
+
+bim <- fread("data/mega/mega_np.bim")
+setnames(bim, 
+         c("V1", "V2", "V4", "V5", "V6"), 
+         c("CHR", "SNP", "BP", "A1", "A2"))
+pheno_ord <- fread("data/mega/nacc_np_ord.pheno")
+pheno_bin <- fread("data/mega/nacc_np.pheno")
+
+#-------------------
+# nacc results
+#--------------------
+results_files_ord <- paste0("output/gwas/adc/polmm/", 
+                            colnames(pheno_ord)[3:ncol(pheno_ord)],
+                            "_polmm_results.txt")
+results_files_bin <- paste0("output/gwas/adc/saige/",
+                            colnames(pheno_bin)[3:ncol(pheno_bin)],
+                            "_saige_results.txt")
+clump_files_ord <- paste0("output/gwas/adc/polmm/", 
+                          colnames(pheno_ord)[3:ncol(pheno_ord)],
+                          "_polmm_results.clumped")
+clump_files_bin <- paste0("output/gwas/adc/saige/",
+                          colnames(pheno_bin)[3:ncol(pheno_bin)],
+                          "_saige_results.clumped")
+
+get_results_list <- function(files) {
+  nfiles <- length(files)
+  file_list <- vector(mode = "list", length = nfiles)
+  for (i in 1:nfiles) {
+    file_list[[i]] <- fread(files[i])
+  }
+  return(file_list)
+}
+
+# collate binary and ordinal outcome results
+results_ord <- get_results_list(results_files_ord)
+for (i in 1:(ncol(pheno_ord) - 2)) {
+  results_ord[[i]][, Phenotype := colnames(pheno_ord)[i + 2]]
+}
+results_ord <- rbindlist(results_ord)
+
+results_bin <- get_results_list(results_files_bin)
+for (i in 1:(ncol(pheno_bin) - 2)) {
+  results_bin[[i]][, Phenotype := colnames(pheno_bin)[i + 2]]
+}
+results_bin <- rbindlist(results_bin)
+
+# collate clumped results
[…]

### 2024-04-30 code/multi_omics/03_make_tidy_caa_data_set.R

@@ -1,0 +1,50 @@
+#=======================================================
+# create data set with 4 CpG, SNP, CAA, and covariates
+#=======================================================
+
+library(data.table)
+library(readxl)
+library(ordinal)
+
+load("data/cpg.RData")
+covar <- setDT(read_xlsx("/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx"))
+rosmap_clin <- fread("data/synapse/ROSMAP_clinical.csv")
+rosmap_clin[, apoe := factor(apoe_genotype, 
+                             levels = c(33, 34, 44, 23, 22), 
+                             labels = c("apoe33", "apoe34", "apoe44", 
+                                        "apoe23", "apoe22"))]
+rosmap_clin[, projid := as.character(projid)]
+# rs7247551 genotype
+# cat 
+# plink --bfile ../np_phewas/data/rosmap/rosmap_np --snp rs7247551 --recode A --out ../multi_omics/data/rs7247551
+snp_geno <- fread("data/rs7247551.raw")
+snp_geno[,  rs7247551_G := 2 - rs7247551_A]
+snp_geno[, rs7247551_A := NULL]
+rep_zeros <- function(x) {
+    paste0(c(rep("0", 8 - nchar(x)), x), collapse = "")
+}
+rep_zeros <- Vectorize(rep_zeros)
+snp_geno[, projid := rep_zeros(IID)]
+
+cpg_apoe <- cpg[, .(IID, specimenID, conversion.efficiency, 
+                    cg04401876, cg09555818, cg10169327, cg13119609)]
+cpg_apoe[, IID := as.character(IID)]
+
+rna <- fread("../np_phewas/data/rnaseq/rnaseq.txt.gz")
+apoc2_col <- colnames(rna)[grep("ENSG00000234906", colnames(rna))]
+apoe_col <- colnames(rna)[grep("ENSG00000130203", colnames(rna))]
+rna_cols <- c("projid", "RIN", apoc2_col, apoe_col)
+rna_apo <- rna[, ..rna_cols]
+rna_apo[, projid := as.character(projid)]
+setnames(rna_apo, c("ENSG00000234906.3", "ENSG00000130203.5"), c("APOC2", "APOE"))
+
+# merge data into one data set
[…]

### 2024-04-30 code/multi_omics/05_caa_apoe_plots.R

@@ -1,0 +1,46 @@
+library(data.table)
+library(ggplot2)
+library(stringi)
+dt <- fread("data/rosmap_data.csv")
+dt[, caa_4gp := ordered(caa_4gp, 
+                        labels = c("None", "Mild", "Moderate", "Severe"))]
+dt[, apoe := factor(apoe, 
+                    levels = c("apoe33", "apoe34", "apoe44", 
+                               "apoe23", "apoe22"))]
+cpgs <- colnames(dt)[grep("cg", colnames(dt))]
+# regress cpg sites on covariates
+get_residuals <- function(cpg, 
+                          covars=c("conversion.efficiency", "msex", "apoe",
+                                   "age_death", "pmi", "study")) {
+    f <- as.formula(paste0(cpg, " ~ ", paste(covars, collapse = "+")))
+    m <- dt[, lm(f, .SD)]
+    res <- dt[, residuals(lm(f, .SD))]
+    res <- data.table(as.integer(names(res)), res)
+    setnames(res, colnames(res), c("row_num", "residual"))
+}
+
+for (i in 1:length(cpgs)) {
+    res <- get_residuals(cpgs[i])
+    dt[res[["row_num"]], (paste0(cpgs[i], "_res")) := scale(res[["residual"]])]
+    rm(res)
+}
+
+plot_data <- melt(dt[!is.na(caa_4gp), 
+                     .(caa_4gp, cg04401876_res, cg09555818_res,
+                       cg10169327_res, cg13119609_res)],
+                  id.vars = "caa_4gp",
+                  variable.factor = FALSE)
+
+plot_data[, variable := stri_replace_first_fixed(variable, "_res", "")]
+p1 <- plot_data[value > -4, ggplot(.SD, aes(caa_4gp, value, fill=variable, alpha=0.8))] +
+    facet_wrap("variable", nrow = 1) +
+    geom_boxplot(aes(alpha = 1), show.legend = FALSE, outlier.alpha = 0) +
+    geom_jitter(aes(alpha = 0.8), width = 0.15, show.legend = FALSE) +
+    labs(y = "Methylation at CpG site",
+         x = "CAA Pathology Severity",
+         fill = "CAA Severity",
+         title = "Methylation at CpG sites vs. cerebral amyloid angiopathy pathology") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, hjust=1, size = 11),
+          strip.text.x = element_text(size = 14))
+p1

### 2024-04-30 code/multi_omics/caa_additional_models.R

@@ -1,0 +1,71 @@
+library(data.table)
+library(ggplot2)
+library(ordinal)
+
+dt <- fread("data/rosmap_data.csv")
+dt[, caa_4gp := ordered(caa_4gp)]
+dt[, apoe := factor(apoe, levels = c("apoe33", "apoe22", "apoe23", "apoe34", "apoe44"))]
+
+apoc2_plot = dt[!is.na(rs7247551_G) & !is.na(APOC2), 
+                ggplot(.SD, aes(factor(rs7247551_G), sqrt(APOC2)))
+]
+
+apoc2_plot + 
+  geom_boxplot() +
+  labs(
+    x = "rs7247551 Genotype",
+    y = "sqrt(APOC2 expression)"
+  ) +
+  scale_x_discrete(labels = c("AA", "AG", "GG")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(hjust=1, size = 11),
+        strip.text.x = element_text(size = 14))
+
+apoe_plot = dt[!is.na(rs7247551_G) & !is.na(APOE), 
+                ggplot(.SD, aes(factor(rs7247551_G), log10(APOE)))
+]
+
+apoe_plot + 
+  geom_boxplot() +
+  labs(
+    x = "rs7247551 Genotype",
+    y = "log10(APOE expression)"
+  ) +
+  scale_x_discrete(labels = c("AA", "AG", "GG")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(hjust=1, size = 11),
+        strip.text.x = element_text(size = 14))
+
+apoc2_caa_plot = dt[!is.na(caa_4gp) & !is.na(APOC2) & sqrt(APOC2) < 5, 
+                    ggplot(.SD, aes(factor(caa_4gp), sqrt(APOC2)))
+]
+
+apoc2_caa_plot + 
+  geom_boxplot() +
+  labs(
+    x = "CAA pathology severity",
+    y = "sqrt(APOC2 expression)"
+  ) +
+  scale_x_discrete(labels = c("None", "Mild", "Moderate", "Severe")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(hjust=1, size = 11),
+        strip.text.x = element_text(size = 14))
+                    
+
+# regression models: SNP -> gene expression
[…]

### 2024-04-30 code/multi_omics/hs_grn_model.R

@@ -1,0 +1,30 @@
+library(data.table)
+library(ordinal)
+library(mediation)
+library(MASS)
+library(readxl)
+rosmap_clin <- setDT(read_xlsx("/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx"))
+rna <- fread("../np_phewas/data/rnaseq/rnaseq.txt.gz")
+grn_col <- colnames(rna)[grep("ENSG00000030582", colnames(rna))]
+rna_cols <- c("projid", "RIN", grn_col)
+rna <- rna[, ..rna_cols]
+rna[, projid := as.character(projid)]
+dt_rna <- merge(rosmap_clin, rna, "projid")
+
+# GRN and HS model
+m_hs_grn = dt_rna[, glm(hspath_typ ~ ENSG00000030582.12 + msex + age_death + RIN + pmi, family = "binomial")]
+# Coefficients:
+#   Estimate Std. Error z value Pr(>|z|)    
+# (Intercept)        -11.34677    3.03992  -3.733 0.000190 ***
+# ENSG00000030582.12   0.01027    0.02105   0.488 0.625673    
+# msex                -0.10972    0.36263  -0.303 0.762215    
+# age_death            0.09986    0.02770   3.605 0.000312 ***
+# RIN                 -0.08210    0.16905  -0.486 0.627215    
+# pmi                  0.01323    0.03167   0.418 0.676275    
+# ---
+#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+# GRN and LATE model
+dt_rna[, late := ordered(tdp_st4)]
+m_late_grn = dt_rna[, clm(late ~ ENSG00000030582.12 + msex + age_death + RIN + pmi, data = .SD)]
+summary(m_late_grn)

### 2024-04-30 code/multi_omics/prep/rna_seq/00_synapse_download.R

@@ -1,0 +1,9 @@
+library(synapser)
+library(synapserutils)
+
+synLogin()
+out_path <- "data/synapse/rna_seq/"
+syncFromSynapse('syn21088596', path=out_path, ifcollision="keep.local")
+syncFromSynapse('syn21323366', path=out_path, ifcollision="keep.local")
+syncFromSynapse('syn3505732', path=out_path, ifcollision="keep.local")
+syncFromSynapse('syn3505724', path=out_path, ifcollision="keep.local")

### 2024-04-30 code/np_coloc/hs_late_ad.R

@@ -3,7 +3,7 @@ library(ggplot2)
 library(coloc)
 
 bel2022 <- 
-  fread("/data_global/summary_statistics/GCST90027158_buildGRCh38.tsv.gz")
+  fread("/data_global/summary_statistics/Bellenguez/GCST90027158_buildGRCh38.tsv.gz")
 bel2022 <- bel2022[!which(duplicated(variant_id))]
 hs <- fread("output/gwas/mega/saige/hs_saige_results.txt")
 hs_pheno <- fread("data/mega/mega_np.pheno", na.strings = "-1")

### 2024-04-30 code/prep/act/00_act_np_ids.R

@@ -1,9 +1,9 @@
 
 library(data.table)
 library(readxl)
-act_np <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx", sheet = 2))
+act_np <- setDT(read_xlsx("raw_data/ACT_235_2023_08_21/np_2023_08_21.xlsx"))
 act_np[, FID := 0]
-fwrite(act_np[, .(FID, IDfromDave)], 
+fwrite(act_np[, .(FID, IDFromDave)], 
        file = "data/act_np_ids.txt", 
        quote = FALSE, 
        col.names = FALSE, 

### 2024-04-30 code/prep/act/09_make_covar.R

@@ -4,15 +4,22 @@
 #=================================
 
 # sex, age of death, ACT cohort, PCs 1-10
-pacman::p_load(data.table, GENESIS, SNPRelate, stringi, readxl)
+library(data.table)
+library(GENESIS)
+library(SNPRelate)
+library(stringi)
+library(readxl)
 
-act <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx",
-                               sheet = 1))
-act_np <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx",
-                                  sheet = 2))
-
-act_cohorts <- fread("data/act/act_ids_adgc.txt", 
-                     colClasses = c("character", "character", "factor"))
+act <- as.data.table(
+  read_xlsx("raw_data/ACT_235_2023_08_21/demo_2023_08_21.xlsx",sheet = 1)
+)
+act_np <- as.data.table(
+  read_xlsx("raw_data/ACT_235_2023_08_21/np_2023_08_21.xlsx", sheet = 1)
+)
+act_cohorts <- fread(
+  "data/act/act_ids_adgc.txt", 
+  colClasses = c("character", "character", "factor")
+)
 
 load('data/act/mypcair.Rdata')
 n_pcs <- 10
@@ -24,13 +31,10 @@ setnames(pcs, 'rn', 'IID')
 # from https://www.maelstrom-research.org/variable/act_recruitment:gender:Collected
 # gender coded as 2 female, 1 male in ACT
 
-act_np <- merge(act_np[, .(IDfromDave, autopsyage)],
-                act[, .(IDfromDave, gender)], 
-                by = "IDfromDave")
-
-act_np[gender == 2, gender := 0]
-setnames(act_np, colnames(act_np), c("IID", "age_death", "msex"))
-act_np[, age_death2 := age_death^2]
+act = act[, .(IDFromDave, gender, age_at_death)]
+act[gender == 2, gender := 0]
+setnames(act, colnames(act), c("IID", "msex", "age_death"))
+# act_np[, age_death2 := age_death^2]
 
 act_cohort_dummy <- model.matrix(~ V3, act_cohorts)
 act_cohorts <- cbind(act_cohorts, act_cohort_dummy)
@@ -38,21 +42,23 @@ act_cohorts <- act_cohorts[, .(V1, V2, V32, V33)]
 setnames(act_cohorts, colnames(act_cohorts), 
[…]

### 2024-04-30 code/prep/act/10_create_phenotypes.R

@@ -1,54 +1,0 @@
-#==================================
-# categorize the NACC NP variables
-# by outcome type, coding, etc.
-#==================================
-
-library(data.table)
-library(readxl)
-library(magrittr)
-
-act <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx",
-                               sheet = 2))
-
-act_fam <- fread("data/act/act_np.fam") %>% 
-  setnames(., c("V1", "V2"), c("FID", "IID"))
-
-adc <- fread("data/adc/adc_np.pheno")
-
-# brain arterioloscerosis binary None or Mild vs. Moderate or Severe
-act[micro_arteriolosclerosis_id < 3, 
-          asc_bin := 0]
-
-act[micro_arteriolosclerosis_id >= 3, 
-          asc_bin := 1]
-
-# Atherosclerosis binary None or Mild vs. Moderate or Severe
-act[ge_atherosclerosis_id < 3, 
-          ath_bin := 0]
-
-act[ge_atherosclerosis_id >= 3,
-          ath_bin := 1]
-
-# Cerebral amyloid angiopathy binary None vs. Mild, Moderate, or Severe
-act[micro_amyloidangiopathyoccipital == 1, caa_bin := 0]
-act[micro_amyloidangiopathyoccipital %in% 2:4, caa_bin := 1]
-
-# CERAD score binary None, Sparse, Moderate vs. Severe
-act[cerad %in% 0:2, cerad3 := 0]
-act[cerad %in% 3, cerad3 := 1]
-
-act <- act[, .(IDfromDave, any_hs, any_mvl, any_macro, any_lb4, asc_bin, 
-               ath_bin, cerad3, braak56, caa_bin)]
-setnames(act, "IDfromDave", "IID")
-
-act <- merge(act_fam[, .(FID, IID)], act, "IID")
-setcolorder(act, "FID")
-
-# IIDs missing cerad23 are also missing all other phenotypes, so remove those
-act <- act[!(is.na(cerad3))]
-
-fwrite(act,
-       file = "data/act/act_np.pheno",
-       quote = FALSE, 
-       sep = " ", 
-       na = -1)

### 2024-04-30 code/prep/act/10_make_pheno.R

@@ -1,0 +1,54 @@
+#==================================
+# categorize the act NP variables
+# by outcome type, coding, etc.
+#==================================
+
+library(data.table)
+library(readxl)
+library(magrittr)
+
+act <- setDT(read_xlsx("raw_data/ACT_235_2023_08_21/np_2023_08_21.xlsx"))
+
+act_fam <- fread("data/act/act_np.fam") %>% 
+  setnames(., c("V1", "V2"), c("FID", "IID"))
+
+setnames(act, "IDFromDave", "IID")
+
+act <- merge(act_fam[, .(FID, IID)], act, "IID")
+setcolorder(act, "FID")
+
+# IIDs missing cerad23 are also missing all other phenotypes, so remove those
+# update as of 2023-08-23 -- no longer true, both samples missing have LATE data
+# act <- act[!(is.na(cerad_score))]
+# 
+
+# rename phenotypes
+setnames(
+  act, 
+  c("late_stg", "cerad_score", "braak_stg", "amylangi_score", "athero_level",
+    "arteriolo_level", "thal_phase", "hipscl_any", "chronic_grossinfarcts_any", "chronic_microinfarcts_any"), 
+  c("late", "cerad", "braak", "caa", "atheroscler", "arteriol", "diffuse_abeta", "hs", "grossinf", "microinf")
+  )
+
+# create ordinal lewy body variable from categorical variable
+act[
+  lewybodydis_cat == 1 | lewybodydis_cat == 1.5 | lewybodydis_cat == 5,
+  lewy_body := 0
+]
+act[lewybodydis_cat == 3, lewy_body := 1]
+act[lewybodydis_cat == 2 | lewybodydis_cat == 4, lewy_body := 2]
+act[lewybodydis_cat == 6, lewy_body := 3]
+
+keep_vars = c(
+  "FID", "IID", "late", "cerad", "braak", "caa", "atheroscler", "arteriol", 
+  "diffuse_abeta", "lewy_body", "hs", "grossinf", "microinf"
+)
+
+act[diffuse_abeta == 5, diffuse_abeta := 4]
+act[diffuse_abeta > 0, diffuse_abeta := diffuse_abeta - 1]
+
+fwrite(act[, ..keep_vars],
+       file = "data/act/act_np.pheno",
+       quote = FALSE, 
+       sep = " ", 
+       na = -1)

### 2024-04-30 code/prep/act/13_make_covar_cerad.R

@@ -1,0 +1,8 @@
+library(data.table)
+
+covar = fread("data/act/act_np.covar")
+pheno = fread("data/act/act_np.pheno", na.strings = "-1")
+
+covar = merge(covar, pheno[, .(IID, cerad)], "IID")
+covar = covar[!is.na(cerad)]
+fwrite(covar, file = "data/act/act_np_cerad.covar")

### 2024-04-30 code/prep/adc/02_np_adgc.R

@@ -4,7 +4,7 @@ library(magrittr)
 
 na_strings <- c("-9", "-4", "9999", "999", "888")
 # Universal Data Set + NP
-uds <- fread("/data_global/nacc/investigator_nacc57.csv", 
+uds <- fread("/data_global/nacc/investigator_nacc61.csv", 
              na.strings = na_strings)
 # NP data not longitudinal, so just keep obs from last visit
 uds <- uds[NACCVNUM == NACCAVST]

### 2024-04-30 code/prep/adc/03_basic_qc.sh

@@ -10,8 +10,8 @@
 
 cat data/adc/np.csv | awk -F "," 'NR>1{print $1,$2}' > tmp/np_ids.tmp
 
-# apply initial MAF filter of 1%
-plink2 --pfile data/adc/adc --keep tmp/np_ids.tmp --maf 0.01 --make-pgen --out tmp/adc_np
+# apply initial MAF filter of 0.5%
+plink2 --pfile data/adc/adc --keep tmp/np_ids.tmp --maf 0.005 --make-pgen --out tmp/adc_np
 
 #-------------------------------------------------------------
 # Step 1: basic variant and individual quality filters
@@ -21,23 +21,9 @@ plink2 --pfile data/adc/adc --keep tmp/np_ids.tmp --maf 0.01 --make-pgen --out t
 # remove variants that have missing genotype rate > 0.05
 #--------------------------------------------------------------
 
-# 20% variant missingness threshold
-plink2 \
-  --pfile tmp/adc_np \
-  --geno 0.2 \
-  --make-pgen \
-  --out tmp/adc_qc1_geno.2.tmp
-
-# 20% IID missingness threshold
-plink2 \
-  --pfile tmp/adc_qc1_geno.2.tmp \
-  --mind 0.2 \
-  --make-pgen \
-  --out tmp/adc_qc1_mind.2.tmp
-
 # 5% variant missingness threshold
 plink2 \
-  --pfile tmp/adc_qc1_mind.2.tmp \
+  --pfile tmp/adc_np \
   --geno 0.05 \
   --make-pgen \
   --out tmp/adc_qc1_geno.05.tmp

### 2024-04-30 code/prep/adc/04_merge_with_1000g.sh

@@ -3,7 +3,7 @@
 #================================
 # Step 3: Remove ethnic outliers
 #================================
-prefix=tmp/adc_np
+prefix=tmp/adc_qc1.tmp
 
 # # convert to PLINK 1.9 fileset
 # # (merging in PLINK 2.0 is currently limited in features)
@@ -15,11 +15,11 @@ prefix=tmp/adc_np
 #   -b1 data/1000g/1000g \
 #   -o tmp/adc_1000g_merged
 
-# # prune
-# plink \
-#   --bfile tmp/adc_1000g_merged \
-#   --indep-pairwise 15000 1500 0.2 \
-#   --out tmp/adc_1000g_merged_prune
+# prune
+plink \
+  --bfile tmp/adc_1000g_merged \
+  --indep-pairwise 15000 1500 0.2 \
+  --out tmp/adc_1000g_merged_prune
 
 # pca
 plink \

### 2024-04-30 code/prep/adc/05_remove_ethnic_outliers.R

@@ -7,7 +7,9 @@ p_load(data.table, magrittr, ggplot2)
 # groundhog.library("ggplot2", day)
 
 # 1000g (otg) data population data
-otg <- fread("/data_global/1000g/integrated_call_samples_v3.20130502.ALL.panel") %>% 
+otg <- fread(
+  "/data_global/1000g/integrated_call_samples_v3.20130502.ALL.panel"
+  ) %>% 
   .[, sample, super_pop]
 
 # PCA data
@@ -82,15 +84,18 @@ InsideCircle <- function(x, y, r, center = c(0, 0)) {
 otg.merged[, include := 
              ifelse(
                is.na(super_pop) == T & 
-                 InsideCircle(PC1.norm, 
-                              PC2.norm, 
-                              radius, 
-                              c(otg.merged[super_pop == "EUR", 
-                                           mean(PC1.norm)], 
-                                otg.merged[super_pop == "EUR", 
-                                           mean(PC2.norm)]
-                              )
-                 ) == T, TRUE, FALSE)
+                 InsideCircle(
+                   PC1.norm, 
+                   PC2.norm, 
+                   radius, 
+                   c(
+                     otg.merged[super_pop == "EUR", mean(PC1.norm)], 
+                     otg.merged[super_pop == "EUR", mean(PC2.norm)]
+                   )
+                 ) == T, 
+               TRUE, 
+               FALSE
+             )
 ]
 
 adc <- otg.merged[include == T, ]

### 2024-04-30 code/prep/adc/11_create_phenotypes.R

@@ -1,255 +1,0 @@
-#==================================
-# categorize the NACC NP variables
-# by outcome type, coding, etc.
-#==================================
-
-library(data.table)
-library(magrittr)
-
-np <- readRDS("data/adc/np.Rds")
-covar <- fread("data/adc/adc_np.covar")
-np <- np[IID %in% covar$IID]
-
-apply_exclusion_criteria <- function(dt) {
-  dt %>% 
-    .[is.na(NACCDOWN) | NACCDOWN != 1] %>%
-    .[is.na(NPPDXA) | NPPDXA != 1] %>% .[is.na(NPPDXB) | NPPDXB != 1] %>%
-    .[is.na(NPPDXD) | NPPDXD != 1] %>% .[is.na(NPPDXE) | NPPDXE != 1] %>%
-    .[is.na(NPPDXF) | NPPDXF != 1] %>% .[is.na(NPPDXG) | NPPDXG != 1] %>%
-    .[is.na(NPPDXH) | NPPDXH != 1] %>% .[is.na(NPPDXI) | NPPDXI != 1] %>%
-    .[is.na(NPPDXJ) | NPPDXJ != 1] %>% .[is.na(NPPDXK) | NPPDXK != 1] %>% 
-    .[is.na(NPPDXL) | NPPDXL != 1] %>% .[is.na(NPPDXM) | NPPDXM != 1] %>% 
-    .[is.na(NPPDXN) | NPPDXN != 1] %>% .[is.na(NACCPRIO) | NACCPRIO != 1] %>%
-    .[is.na(NPPATH10) | NPPATH10 != 1] %>% 
-    .[is.na(NPALSMND) | NPALSMND != 1] %>% 
-    .[is.na(NPFTDTAU) | NPFTDTAU != 1] %>% 
-    .[is.na(NPOFTD) | NPOFTD != 1] %>% 
-    .[is.na(NPFTDTDP) | NPFTDTDP != 1] %>% 
-    as.data.table()
-}
-
-exclusion_vars <- c("NACCDOWN", "NPPDXA", "NPPDXB", "NPPDXD", "NPPDXE", 
-                    "NPPDXF", "NPPDXG", "NPPDXH", "NPPDXI", "NPPDXJ", 
-                    "NPPDXK", "NPPDXL", "NPPDXM", "NPPDXN", "NACCPRIO", 
-                    "NPPATH10", "NPALSMND", "NPFTDTAU", "NPOFTD", "NPFTDTDP")
-
[…]

### 2024-04-30 code/prep/adc/12_make_pheno.R

@@ -1,0 +1,61 @@
+# create NACC phenotyp file for POLMM analysis
+library(data.table)
+np = readRDS("data/adc/np_qced.Rds")
+
+
+
+# TDP-43 ordinal var
+# Stage 1: Amygdala only
+# Stage 2: + Hippocampus
+# Stage 3: + Middle frontal gyrus or cortex
+np[NPTDPB == 0 | NPTDPC == 0 | NPTDPD == 0 | NPTDPE == 0, late := 0]
+np[NPTDPB == 1, late := 1] 
+np[NPTDPC == 1 | NPTDPD == 1, late := 2]
+np[NPTDPE == 1, late := 3] 
+
+# neuritic plaque CERAD score ordinal
+np[!is.na(NACCNEUR), cerad := NACCNEUR]
+
+# braak score ordinal
+np[!is.na(NACCBRAA), braak := NACCBRAA]
+
+# Thal CERAD score / diffuse plaques ordinal and binary vars
+np[!is.na(NACCDIFF), diffuse_abeta := NACCDIFF]
+
+# arteriolosclerosis
+np[!is.na(NACCARTE), arteriol := NACCARTE]
+
+# atherosclerosis in circle of willis
+np[NACCAVAS %in% 0:3, atheroscler := NACCAVAS]
+
+# lewy body
+np[NACCLEWY %in% 0:3, lewy_body := NACCLEWY]
+np[NACCLEWY == 4 & NPLBOD == 5, lewy_body := 0] # olfactory
+
+# cerebral amyloid angiopathy
+np[NACCAMY %in% 0:3, caa := NACCAMY]
+
+# Hippocampal sclerosis
+np[NPHIPSCL == 0 | NPSCL == 0, hs := 0]
+np[NPHIPSCL %in% c(1, 2, 3) | NPSCL == 1, hs := 1]
+
+# microinfarcts
+np[NACCMICR == 0, microinf := 0]
+np[NACCMICR == 1, microinf := 1]
+
+# gross infarcts
+np[NACCINF == 0, grossinf := 0]
+np[NACCINF == 1, grossinf := 1]
+
+keep_vars_ord = c(
+  "FID", "IID", "late", "braak", "cerad", "diffuse_abeta", "arteriol", 
+  "atheroscler", "lewy_body", "caa"
+)
+
+keep_vars_bin = c("FID", "IID", "hs", "microinf", "grossinf")
[…]

### 2024-04-30 code/prep/mc_caa/01_download_synapse_data.py

@@ -1,0 +1,10 @@
+import synapseclient
+import synapseutils
+syn = synapseclient.Synapse()
+syn.login()
+
+
+outdir = "data/mc_caa/"
+# files = synapseutils.syncFromSynapse(syn, 'syn17010931', path = outdir + "Metadata/", ifcollision="keep.local") 
+# files = synapseutils.syncFromSynapse(syn, 'syn21547862', path = outdir + "Genetic_Variants/", ifcollision="keep.local") 
+# files = synapseutils.syncFromSynapse(syn, 'syn10850933', path = outdir + "Gene_Expression/", ifcollision="keep.local") 

### 2024-04-30 code/prep/mc_caa/analysis.R

@@ -1,0 +1,36 @@
+# attempt to replicate results using Mayo CAA study data
+
+library(data.table)
+library(ordinal)
+
+mccaa = fread("data/mc_caa/Metadata/MC-CAA_individual_human_metadata.csv")
+mccaa_geno = fread("data/mc_caa/Metadata/MC-CAA_assay_snparray_metadata.csv")
+dup_ids = c("18645", "18549", "815", "1287", "874", "3900", "925", "1014")
+batchA_chr17 = fread("data/mc_caa/plink/mc_caa_batchA_chr17_rs72844606.raw", check.names = TRUE)
+batchB_chr17 = fread("data/mc_caa/plink/mc_caa_batchB_chr17_rs72844606.raw", check.names = TRUE)
+batchA_chr19 = fread("data/mc_caa/plink/mc_caa_batchA_chr19_rs7247551.raw", check.names = TRUE)
+batchB_chr19 = fread("data/mc_caa/plink/mc_caa_batchB_chr19_rs7247551.raw", check.names = TRUE)
+
+mccaa = mccaa[!(individualID %in% dup_ids)]
+mccaa_geno[, individualID := as.character(specimenID)]
+mc = merge(mccaa, mccaa_geno, "individualID")
+mc[ageDeath == "90+", ageDeath := "90"] # ages >90 are truncated in synapse data
+mc[, ageDeath := as.numeric(ageDeath)]
+mc[, sex := factor(sex)]
+mc[, apoeGenotype := factor(apoeGenotype)]
+mc = mc[, .(individualID, AverageCAA, Braak, sex, ageDeath, PC1, PC2, PC3, PC4, PC5, apoeGenotype)]
+mc[, individualID := as.integer(individualID)]
+
+chr17 = rbind(batchA_chr17, batchB_chr17)
+mc17 = merge(mc, chr17, by.x = "individualID", by.y = "IID")
+
+model17 = mc17[Braak != 3.5, clm(as.ordered(Braak) ~ X17.8833591_T + ageDeath + sex + PC1 + PC2 + PC3 + PC4 + PC5)]
+
+# major allele A is minor in Batch A; switch alleles
+batchA_chr19[, X19.45454766_G := -X19.45454766_A + 2]
+batchA_chr19[, X19.45454766_A := NULL]
+chr19 = rbind(batchA_chr19, batchB_chr19)
[…]

### 2024-04-30 code/prep/mcbb/regression.R

@@ -1,13 +1,0 @@
-
-library(data.table)
-library(ordinal)
-mcbb <- fread("data/mcbb/MayoRNAseq_individual_metadata_031422.csv")
-snp <- fread("data/mcbb/tmem106b.raw", check.names = TRUE)
-
-dt <- merge(mcbb, snp, by.x = "individualID", by.y = "IID")
-
-dt[, Braak_rd := as.ordered(floor(Braak))]
-dt[ageDeath == "90_or_over", ageDeath := "90"]
-dt[, ageDeath := as.numeric(ageDeath)]
-m <- dt[race == "White", clm(as.ordered(Braak_rd) ~ X7.12269593_A + sex + ageDeath + factor(apoeGenotype))]
-summary(m)

### 2024-04-30 code/prep/mega/02_make_pheno.R

@@ -1,231 +1,0 @@
-library(data.table)
-library(stringi)
-library(readxl)
-
-adc <- fread("data/adc/adc_np.pheno", na.strings = '-1')
-ros <- fread("data/rosmap/rosmap_np.pheno", na.strings = '-1')
-act <- fread("data/act/act_np.pheno", na.strings = '-1')
-adni <- fread("data/adni/adni_np.pheno", na.strings = '-1')
-fam <- fread("data/mega/mega_np.fam")
-
-act_np <- as.data.table(read_xlsx("raw_data/act/DataForE235_20210421.xlsx",
-                                  sheet = 2))
-adc_np <- setDT(readRDS("data/adc/np_qced.Rds")) # %>%
-  # replace_with_na_all(condition = ~.x == -1)
-setDT(adc_np)
-ros_np <- setDT(read_xlsx(
-  "/data_global/ROSMAP/greg_20200109/dataset_843_basic_01-09-2020.xlsx"))
-adni_np <- fread("raw_data/NEUROPATH_07_06_21.csv")
-adni_fam <- fread("data/adni/adni_np.fam")
-adni_np <- merge(adni_fam, adni_np, by.x = 'V2', by.y = 'RID')
-rm(adni_fam)
-
-
-set_ids_as_char <- function(dt) {
-  dt[, IID := as.character(IID)]
-  dt[, FID := as.character(FID)]
-}
-
-set_ids_as_char(adc)
-set_ids_as_char(ros)
-set_ids_as_char(act)
-set_ids_as_char(adni)
-
-ros_np[, FID := '1']
-act_np[, FID := '2']
-setnames(ros_np, 'projid', 'IID')
-setnames(act_np, 'IDfromDave', 'IID')
-setnames(adni_np, c('V1', 'V2'), c('FID', 'IID'))
-set_ids_as_char(adc_np)
-set_ids_as_char(ros_np)
-set_ids_as_char(act_np)
-set_ids_as_char(adni_np)
-
-
-
-
-# rename ADC, ROSMAP, and ACT phenotypes to match those in ADNI
-# (since I renamed those myself already)
-phenotypes <- colnames(adni)[3:length(colnames(adni))]
-setnames(adc, c('LEWY', 'ATHCW', 'NACCINF', 'NACCMICR', 'HS', 'BRAAK', 
-                'NEUR', 'CAA', 'ASC'), 
-         phenotypes)
-setnames(ros, 
[…]

### 2024-04-30 code/prep/rosmap/02a_check_heterozygosity.R

@@ -4,7 +4,7 @@
 
 library(data.table)
 
-het <- fread("tmp/rosmap_1f.het")
+het <- fread("tmp/rosmap_1c.het")
 
 # (total number genotypes - number homozygous) / total number genotypes
 het[, het.rate := (`N(NM)` - `O(HOM)`) / `N(NM)`]
@@ -14,5 +14,5 @@ het.fail <- het[het.rate < mean(het.rate) - 3 * sd(het.rate) |
                   het.rate > mean(het.rate) + 3 * sd(het.rate)]
 het.fail[, het.dist := (het.rate - mean(het$het.rate)) / sd(het$het.rate)]
 
-write.table(het.fail, file = "tmp/rosmap_1f_fail_het.txt", 
+write.table(het.fail, file = "tmp/rosmap_1c_fail_het.txt", 
             quote = F, col.names = F, row.names = F)

### 2024-04-30 code/prep/rosmap/06_prep_for_pcair.sh

@@ -9,7 +9,7 @@ prefix_out="$prefix"_np
 # one final pass at variant QC
 plink \
   --bfile tmp/"$prefix"_3 \
-  --maf 0.05 \
+  --maf 0.005 \
   --geno 0.05 \
   --hwe 1e-6 midp include-nonctrl \
   --make-bed \
